---
title: "Data Processing Names"
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions
---

## Getting Started

```{r}
#| warning: false
#| output: false
# clear the global environment
rm(list = ls())
gc()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(readxl)
library(stringr)
library(lubridate)
library(stringi)
library(sqids)

# load custom functions
source("src/utils/custom_functions.r")
.load_quarto_dependencies()
```

## Functions

### Parse Names
**`parse_names()`** — split full names into given name(s), surname particle, and surname

This function takes a data frame with a column `naam` (full name as a single string) and returns a tibble with columns `naam, first_name, particle, and last_name`. It recognizes multi-word surname particles common in Dutch/German/Romance names (e.g., *van, van der, ’t, von der, de la, del, di, da, zu, op de*). For matching, it normalizes whitespace, unifies curly/straight apostrophes, lowercases, and transliterates diacritics to ASCII; the **longest** particle immediately before the final token (up to 3 words) is selected. The detected particle is returned in lowercase; first_name contains all tokens before the particle (or before the last token if none). If no particle is present, particle = NA. Single-token names are returned with that token as both first_name and last_name.

```{r}
parse_names = function(names){
  particles <- c(
    "de","den","der","het","te","ten","ter",
    "van","van de","van den","van der","van 't","van ’t",
    "'t","’t",
    "von","von der","von den",
    "la","le","du","del","della","di","da","dos",
    "das","de la","de los","de las",
    "zu","zum","zur", "op de"
  )

  # Normalization helper for matching
  .normalize <- function(x) {
    x |>
      str_squish() |>
      str_replace_all("’", "'") |>
      str_to_lower() |>
      stri_trans_general("Latin-ASCII")
    }

  p_norm <- .normalize(particles)

  names |>
    mutate(
      tokens_raw = str_split(naam, "\\s+"),
      tokens_norm = map(tokens_raw, ~ .normalize(.x)),
      n = map_int(tokens_raw, length),

      # find the longest particle match immediately before the final token
        particle_idx = map2_int(tokens_norm, n, function(tok, n_tok) {
          if (n_tok < 2) return(NA_integer_)
          # check 3, 2, 1-token particles that end at position n_tok-1
          for (k in 3:1) {
            start <- n_tok - k
            end   <- n_tok - 1
            if (start >= 1) {
              cand <- paste(tok[start:end], collapse = " ")
              if (cand %in% p_norm) return(start)
            }
          }
          NA_integer_
        }),

        has_particle = !is.na(particle_idx),

        first_name = pmap_chr(
          list(tokens_raw, particle_idx, n),
          function(tok, p_i, n_tok) {
            end_giv <- if (is.na(p_i)) n_tok - 1 else p_i - 1
            if (end_giv <= 0) tok[1] 
            else paste(tok[seq_len(end_giv)], collapse = " ")
          }
        ),

        particle = pmap_chr(
            list(tokens_raw, tokens_norm, particle_idx, n),
            function(tok_raw, tok_norm, p_i, n_tok) {
              if (is.na(p_i)) return(NA_character_)
              # output particle in lowercase, canonicalised apostrophes
              out <- paste(tok_norm[p_i:(n_tok-1)], collapse = " ")
              out
            }
          ) |> str_to_lower(),

        last_name = pmap_chr(
          list(tokens_raw, n),
          function(tok, n_tok) tok[n_tok]
        )
    ) |>
    select(naam, first_name, particle, last_name)
}
```

### Extract Initials

**`extract_initials()`** — detect leading initials and store them in initials

This function scans the start of the naam string for compact initials and writes them to a new column `initials`. It recognizes one- or two-letter uppercase blocks followed by a dot, repeated (e.g., J., J.F., A.B.), with optional hyphenated blocks (e.g., J.-P., J.F.-K.). A match is only made when these initials are immediately followed by whitespace. If initials are found, the existing first_name is set to NA (so we don’t misclassify initials as given names). Finally, initials is relocated to sit right after `naam`.

```{r}
extract_initials = function(names){
  names |>
    mutate(
      initials = str_extract(
          naam, 
          "^((?:\\p{Lu}{1,2}\\.)+(?:-\\p{Lu}{1,2}\\.)*)(?=\\s)"
        ),
      first_name = ifelse(is.na(initials), first_name, NA_character_)) |>
    relocate(initials, .after=naam)
}

```

### Patch Names

**`patch_names()`** — apply manual corrections to name fields by key (naam)

This function loads a table of curated fixes and updates matching rows in names using `dplyr::rows_update()`. For each value of naam that appears in the corrections file, overlapping columns (e.g., `first_name, particle, last_name, initials`, etc.) are overwritten with the corrected values. Rows without a match are left unchanged; no new rows are added and the original row order is preserved.

```{r}
patch_names = function(names){
  # read in dataset with corrections for name information
  corrections = readRDS(file.path('data', 'utils', 'name_corrections.Rds'))

  names |> rows_update(corrections, by='naam', unmatched='ignore')
}
```

```{r}
readRDS(file.path('data', 'utils', 'name_corrections.Rds')) |> head()
```


### Extract Maiden Name

**`extract_maiden_name()`** — split hyphenated surnames into last_name and maiden_name

This function looks for a hyphen (-) in the `last_name` column and, when present, assigns the part **before** the first hyphen back to last_name and the part **after** it to a new column maiden_name. If no second part exists (i.e., no hyphen), `maiden_name` is set to `NA`. All other columns are left unchanged.

```{r}
extract_maiden_name = function(names){
  # split last names
  last_name_splits = str_split(names$last_name, '-', simplify = TRUE)

  # add splits to names dataframe
  names['last_name'] = last_name_splits[,1]
  names['maiden_name'] = ifelse(
    last_name_splits[,2] == '',
    NA_character_,
    last_name_splits[,2]
  )

  return(names)
}
```


### Format names

**`format_names()`** — normalize characters, standardize case, and build composite name fields

This function harmonizes name components and constructs two cleaned display fields. It transliterates diacritics to ASCII for `initials`, `first_name`, `particle`, `last_name`, and `maiden_name` (using ICU via `stri_trans_general()`), lowercases particle, and then:
	1.	creates `clean_name` by uniting `initials:last_name` with spaces while skipping missing parts;
	2.	creates `clean_name_full` by appending `maiden_name` to `clean_name` with a hyphen (skipping if maiden_name is NA);
	3.	when both `initials` and `first_name` are present, removes the leading "initials " from both clean_name and `clean_name_full` to avoid duplication;
	4.	drops temporary helper columns.

```{r}
format_names = function(data){
  data |>
    mutate(
      initials = initials |> stri_trans_general("Latin-ASCII"),
      first_name = first_name |> stri_trans_general("Latin-ASCII"),
      particle = particle |> 
        str_to_lower() |>
        stri_trans_general("Latin-ASCII"),
      last_name = last_name |> stri_trans_general("Latin-ASCII"),
      maiden_name = maiden_name |> stri_trans_general("Latin-ASCII")
    ) |> 
    unite("clean_name", 
          initials:last_name, na.rm=TRUE, sep=" ", remove=FALSE) |>
    unite("clean_name_full",
          clean_name, maiden_name, na.rm=TRUE, sep="-", remove=FALSE) |>
    mutate(
      temp = str_remove(clean_name, paste0(initials, " ")),
      temp_full = str_remove(clean_name_full, paste0(initials, " ")),
      clean_name = case_when(
        !is.na(initials) & !is.na(first_name) ~ temp,
        .default=clean_name
      ),
      clean_name_full = case_when(
        !is.na(initials) & !is.na(first_name) ~ temp_full,
        .default=clean_name_full
      ),
    ) |>
    select(-starts_with("temp"))
    
}

```

### Add unique identifier

**`add_uid()`** — generate short unique IDs for each clean_name and attach them

This function creates a lookup table that maps every distinct clean_name to a short, URL-safe identifier uid using sqids. IDs are generated by encoding the sequence `i * 25` (for `i = 1..n`) with `sqids::encode()` and a minimum length of 10 characters, then joined back to the data and placed as the first column. The lookup is also saved to disk for external reuse.

```{r}
add_uid = function(names){
  # configure sqids
  settings <- sqids::sqids_options(min_length = 10)
  unique_names = names$clean_name |> unique()

  # generate a id for all names
  lookup = list()
  for (i in 1:length(unique_names)){
    name = unique_names[[i]]
    id = sqids::encode(i * 25, settings)
    lookup[[i]] = tibble(
      clean_name = name,
      uid = id
    )
  }

  lookup = bind_rows(lookup)
  saveRDS(lookup, file.path('data', 'utils', 'uid_key.Rds'))

  names |>
    left_join(lookup) |>
    relocate(uid, .before=1)
}

```

## Application

```{r}
#| warning: false
#| output: false
data = freadRDS2('scholarid') 

names = data |>
  select(naam) |>
  arrange(naam) |>
  distinct() |>
  drop_na() |>
  parse_names() |>
  extract_initials() |>
  patch_names() |>
  extract_maiden_name() |>
  format_names() |>
  add_uid()

fsaveRDS(names, 'names')
```

```{r}
names |> head()
```