---
title: "Data Processing Names"
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions
---

## Getting Started

```{r}
#| warning: false
#| output: false
# load custom functions
source("src/utils/custom_functions.r")

# clear the global environment and set dependencies
.clear_global_environment()
.load_quarto_dependencies()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(readxl)
library(stringr)
library(lubridate)
library(stringi)
library(sqids)
```

## Functions

### `parse_names()`
Split full names into given name(s), surname particle, and surname.

#### Arguments
**`names`**  

> Tibble or data frame containing a column `naam` with full names stored as a single string.

#### Method
The function defines a set of common surname particles (e.g., `van`, `van der`, `de la`, `von der`, `op de`) and a helper `.normalize()` that trims whitespace, unifies apostrophes, lowercases, and transliterates characters to ASCII. For each `naam`, it splits the name into tokens, normalizes them, and searches immediately before the final token for the **longest matching particle** of length 1–3 tokens. Using the detected particle position, it constructs `first_name` (all tokens before the particle, or before the last token if there is no particle), `particle` (the matched particle, returned in lowercase), and `last_name` (the final token). Single-token names are treated as both `first_name` and `last_name`. The function returns a tibble with `naam`, `first_name`, `particle`, and `last_name`.

```{r}
parse_names = function(names){
  particles <- c(
    "de","den","der","het","te","ten","ter",
    "van","van de","van den","van der","van 't","van ’t",
    "'t","’t",
    "von","von der","von den",
    "la","le","du","del","della","di","da","dos",
    "das","de la","de los","de las",
    "zu","zum","zur", "op de"
  )

  # Normalization helper for matching
  .normalize <- function(x) {
    x |>
      str_squish() |>
      str_replace_all("’", "'") |>
      str_to_lower() |>
      stri_trans_general("Latin-ASCII")
    }

  p_norm <- .normalize(particles)

  names |>
    mutate(
      tokens_raw = str_split(naam, "\\s+"),
      tokens_norm = map(tokens_raw, ~ .normalize(.x)),
      n = map_int(tokens_raw, length),

      # find the longest particle match immediately before the final token
        particle_idx = map2_int(tokens_norm, n, function(tok, n_tok) {
          if (n_tok < 2) return(NA_integer_)
          # check 3, 2, 1-token particles that end at position n_tok-1
          for (k in 3:1) {
            start <- n_tok - k
            end   <- n_tok - 1
            if (start >= 1) {
              cand <- paste(tok[start:end], collapse = " ")
              if (cand %in% p_norm) return(start)
            }
          }
          NA_integer_
        }),

        has_particle = !is.na(particle_idx),

        first_name = pmap_chr(
          list(tokens_raw, particle_idx, n),
          function(tok, p_i, n_tok) {
            end_giv <- if (is.na(p_i)) n_tok - 1 else p_i - 1
            if (end_giv <= 0) tok[1] 
            else paste(tok[seq_len(end_giv)], collapse = " ")
          }
        ),

        particle = pmap_chr(
            list(tokens_raw, tokens_norm, particle_idx, n),
            function(tok_raw, tok_norm, p_i, n_tok) {
              if (is.na(p_i)) return(NA_character_)
              # output particle in lowercase, canonicalised apostrophes
              out <- paste(tok_norm[p_i:(n_tok-1)], collapse = " ")
              out
            }
          ) |> str_to_lower(),

        last_name = pmap_chr(
          list(tokens_raw, n),
          function(tok, n_tok) tok[n_tok]
        )
    ) |>
    select(naam, first_name, particle, last_name)
}
```

---

### `extract_initials()`
Detect leading initials in `naam` and store them in a separate `initials` column.

#### Arguments
**`names`**  

> Tibble or data frame containing at least `naam` and `first_name` columns.

#### Method
The function uses a regular expression to detect compact leading initials at the start of `naam`, such as `J.`, `J.F.`, `A.B.`, or hyphenated variants like `J.-P.` or `J.F.-K.`. A match is only made when these initials are followed by whitespace, ensuring that only true initials are captured. The extracted string is stored in a new column `initials`. When initials are found, `first_name` is set to `NA` to avoid misclassifying initials as given names. Finally, the `initials` column is relocated to sit immediately after `naam`, leaving the rest of the structure unchanged.

```{r}
extract_initials = function(names){
  names |>
    mutate(
      initials = str_extract(
          naam, 
          "^((?:\\p{Lu}{1,2}\\.)+(?:-\\p{Lu}{1,2}\\.)*)(?=\\s)"
        ),
      first_name = ifelse(is.na(initials), first_name, NA_character_)) |>
    relocate(initials, .after=naam)
}

```

---

### `patch_names()`
Apply curated manual corrections to name fields by matching on `naam`.

#### Arguments
**`names`**  

> Tibble or data frame with a `naam` column and corresponding name components (e.g., `first_name`, `particle`, `last_name`, `initials`) that may require correction.

#### Method
The function reads a pre-compiled corrections table from `data/utils/name_corrections.Rds`. This table contains corrected values for specific `naam` entries and overlapping columns (such as `first_name`, `particle`, `last_name`, `initials`, etc.). Using `dplyr::rows_update()` with `by = "naam"` and `unmatched = "ignore"`, it overwrites the relevant columns in `names` wherever a matching `naam` is found, while leaving non-matching rows unchanged. No new rows are created and the original order of `names` is preserved. The result is a version of `names` where known problematic cases have been fixed using curated corrections.

```{r}
patch_names = function(names){
  # read in dataset with corrections for name information
  corrections = readRDS(file.path('data', 'utils', 'name_corrections.Rds'))

  names |> rows_update(corrections, by='naam', unmatched='ignore')
}
```

```{r}
readRDS(file.path('data', 'utils', 'name_corrections.Rds')) |> head()
```

---

### `extract_maiden_name()`
Split hyphenated surnames into `last_name` and `maiden_name`.

#### Arguments
**`names`**  

> Tibble or data frame containing a `last_name` column with surnames, some of which may be hyphenated (e.g., `"Jansen-Smits"`).

#### Method
The function splits each `last_name` on the first hyphen using `str_split(..., simplify = TRUE)`. The part before the hyphen is written back to `last_name`, and the part after the hyphen is stored in a new column `maiden_name`. When no second part exists (i.e., no hyphen in the surname), `maiden_name` is set to `NA_character_`. All other columns in `names` are left unchanged. The updated data frame, now carrying both `last_name` and `maiden_name`, is returned.

```{r}
extract_maiden_name = function(names){
  # split last names
  last_name_splits = str_split(names$last_name, '-', simplify = TRUE)

  # add splits to names dataframe
  names['last_name'] = last_name_splits[,1]
  names['maiden_name'] = ifelse(
    last_name_splits[,2] == '',
    NA_character_,
    last_name_splits[,2]
  )

  return(names)
}
```

---

### `format_names()`
Normalize characters, standardize case, and construct cleaned composite name fields.

#### Arguments
**`data`**  

> Tibble or data frame containing at least `initials`, `first_name`, `particle`, `last_name`, and optionally `maiden_name`.

#### Method
The function first transliterates diacritics to ASCII for `initials`, `first_name`, `particle`, `last_name`, and `maiden_name` using `stri_trans_general("Latin-ASCII")`. The `particle` is additionally forced to lowercase. It then creates two composite display fields:  
1. `clean_name` by uniting `initials:last_name` into a single string, skipping missing parts and using spaces as separators;  
2. `clean_name_full` by appending `maiden_name` to `clean_name` with a hyphen, again skipping missing values.

Because some names may contain both `initials` and `first_name`, the function removes the leading `"initials "` from both `clean_name` and `clean_name_full` to avoid duplication (e.g., `"J. Jan Jansen"` becomes `"Jan Jansen"`). This is done via temporary helper columns `temp` and `temp_full`, which are then used to conditionally overwrite the clean name fields and subsequently removed. The function returns the input data with two additional harmonized name display columns: `clean_name` and `clean_name_full`.

```{r}
format_names = function(data){
  data |>
    mutate(
      initials = initials |> stri_trans_general("Latin-ASCII"),
      first_name = first_name |> stri_trans_general("Latin-ASCII"),
      particle = particle |> 
        str_to_lower() |>
        stri_trans_general("Latin-ASCII"),
      last_name = last_name |> stri_trans_general("Latin-ASCII"),
      maiden_name = maiden_name |> stri_trans_general("Latin-ASCII")
    ) |> 
    unite("clean_name", 
          initials:last_name, na.rm=TRUE, sep=" ", remove=FALSE) |>
    unite("clean_name_full",
          clean_name, maiden_name, na.rm=TRUE, sep="-", remove=FALSE) |>
    mutate(
      temp = str_remove(clean_name, paste0(initials, " ")),
      temp_full = str_remove(clean_name_full, paste0(initials, " ")),
      clean_name = case_when(
        !is.na(initials) & !is.na(first_name) ~ temp,
        .default=clean_name
      ),
      clean_name_full = case_when(
        !is.na(initials) & !is.na(first_name) ~ temp_full,
        .default=clean_name_full
      ),
    ) |>
    select(-starts_with("temp"))
    
}

```

---

### `add_uid()`
Generate short unique identifiers for each `clean_name` and attach them as `uid`.

#### Arguments
**`names`**  

> Tibble or data frame containing a `clean_name` column with standardized name strings.

#### Method
The function first configures sqids with a minimum length of 10 characters and extracts the set of unique `clean_name` values. For each distinct `clean_name`, it generates a deterministic numeric sequence `i * 25` (for `i = 1..n`) and encodes this number into a short, URL-safe string `uid` using `sqids::encode()`. These are collected into a lookup tibble with columns `clean_name` and `uid`, which is saved to `data/utils/uid_key.Rds` for reuse outside this script. The original `names` data is then left-joined with the lookup on `clean_name`, and the resulting `uid` column is relocated to the first position. The function returns the input data augmented with a stable, compact identifier per unique name.

```{r}
add_uid = function(names){
  # configure sqids
  settings <- sqids::sqids_options(min_length = 10)
  unique_names = names$clean_name |> unique()

  # generate a id for all names
  lookup = list()
  for (i in 1:length(unique_names)){
    name = unique_names[[i]]
    id = sqids::encode(i * 25, settings)
    lookup[[i]] = tibble(
      clean_name = name,
      uid = id
    )
  }

  lookup = bind_rows(lookup)
  saveRDS(lookup, file.path('data', 'utils', 'uid_key.Rds'))

  names |>
    left_join(lookup) |>
    relocate(uid, .before=1)
}

```

## Application

```{r}
#| warning: false
#| output: false
data = freadRDS2('scholarid') 

names = data |>
  select(naam) |>
  arrange(naam) |>
  distinct() |>
  drop_na() |>
  parse_names() |>
  extract_initials() |>
  patch_names() |>
  extract_maiden_name() |>
  format_names() |>
  add_uid()

fsaveRDS(names, 'names')
```

```{r}
names |> head()
```