---
title: "Data Processing Gender"
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions

params:
  eval: false
---

## Getting Started

```{r}
#| warning: false
#| output: false
# load custom functions
source("src/utils/custom_functions.r")

# clear the global environment and set dependencies
.clear_global_environment()
.load_quarto_dependencies()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(readxl)
library(stringr)
library(lubridate)
library(httr2)
library(rvest)
library(xml2)
library(purrr)
#remotes::install_github("kalimu/genderizeR")
library(genderizeR)
library(dotenv)

```

## Functions
### `is_ok()`

Checks if the response status is succesfull

#### Arguments
**`resp`**

> `httr2` response object from `request()` and `perform_req()`

#### Method
This helper function checks if the response status is in the `2XX`s

```{r}
is_ok = function(resp) resp_status(resp) >= 200 && resp_status(resp) < 300
```

---

### `request_gender()`
Requests gender information for a given first name from the Meertens NVB website.

#### Arguments
**`first_name`**  

> Character string with the first name to be queried.

**`base`**  

> Base URL of the Meertens NVB endpoint. Defaults to `https://nvb.meertens.knaw.nl/naam/is/`.

**`pause`**  

> Numeric value indicating the number of seconds to wait (plus a small random jitter) before performing the request, to be polite to the server. Defaults to `0.5`.

#### Method
This function constructs a query URL by combining the base NVB endpoint with a URL-encoded, lowercased version of `first_name`. It then builds an `httr2` request with a realistic user agent, timeout, and HTTP headers. The request is configured with a retry strategy that retries on status codes `429` and `5xx`, using a jittered exponential backoff. After an optional polite pause, the function performs the request inside a `tryCatch` block to avoid throwing on transport errors. The result is returned as a uniform list containing: a logical `ok` flag (based on `is_ok()`), the HTTP status code, the URL, the raw response object (if available), and any captured error, so that the caller can handle successes and failures in a consistent way.

```{r}
extract_gender_information = function(
    resp,
    first_name
  ){
  # extract all the tables on the page
  html = read_html(resp_body_string(resp))
  tables = html_table(html, header=TRUE)

  # select the first table if a table if found
  if (length(tables) == 0) stop("No tables were found")
  tab = tables[[1]]

  # extract information from table
  male_count = tab[1, 3] |> pull()
  male_count = ifelse(male_count == '--', 0, as.numeric(male_count))
  female_count = tab[5, 3] |> pull() 
  female_count = ifelse(female_count == '--', 0, as.numeric(female_count))
  probability_male = male_count / (female_count + male_count) 

  # configure results table
  res = tibble::tribble(
    ~first_name, ~male_count, ~female_count, ~probability_male,
    first_name, male_count,  female_count,  probability_male
  )
  return(res)
}
```

---

### `get_gender_row()`
Retrieves a single gender-information row for a given first name, using a local cache when possible and the Meertens NVB website otherwise.

#### Arguments
**`name`**  

> Character string with the first name to be queried.

**`gender`**  

> Tibble containing cached gender information, including a `first_name` column. If `name` is found here, the cached row is returned instead of performing a new request.

#### Method
The function first checks whether `name` is already present in the cached `gender` tibble (via `first_name`). If so, it immediately returns the matching row. If not, it calls `request_gender(name)` to query the Meertens NVB endpoint. When the request is not successful (transport error or non-2xx status), it returns a tibble with `NA` values for all gender-related fields, so downstream code can handle failures consistently. When the request is successful, it passes the response and the name to `safe_extract()`, a `purrr::possibly()`-wrapped version of `extract_gender_information` that returns a default NA tibble on error. The resulting tibble is then augmented with `first_name = name` and returned as a single-row tibble with gender counts and probability.

```{r}
safe_extract = purrr::possibly(
  extract_gender_information,
  otherwise = tibble::tibble(
    first_name       = NA_character_,
    male_count       = NA_integer_,
    female_count     = NA_integer_,
    probability_male = NA_real_
  )
)

get_gender_row = function(name, gender) {
  # If cached, return from cache
  if (name %in% gender$first_name) {
    return(gender |> filter(first_name == name))
  }

  r <- request_gender(name)

  # If transport error or HTTP not OK, surface status & keep going
  if (!isTRUE(r$ok)) {
    return(tibble(
      first_name       = NA_character_,
      male_count       = NA_integer_,
      female_count     = NA_integer_,
      probability_male = NA_real_
    ))
  }

  out <- safe_extract(r$resp, name) |>
    mutate(first_name = name)

  out
}
```

---

### `patch_gender_on_splits()`
Patches missing gender information by splitting compound first names and aggregating gender estimates across the split parts.

#### Arguments
**`gender`**  

> Tibble containing gender information with at least `first_name`, `male_count`, `female_count`, and `probability_male` columns. Some rows may have missing gender statistics that need to be patched.

#### Method
The function first constructs a `gender_cache` by dropping rows with missing values from `gender`, so that only complete, reliable gender records are used as a cache. It then selects all rows in `gender` where `male_count` is `NA`, and splits their `first_name` values on spaces (e.g. `"Jan Pieter"` → `"Jan"`, `"Pieter"`), unnests these splits, and keeps the mapping between the original `first_name` and each `first_name_split`.

Next, it extracts the unique set of `first_name_split` values and calls `get_gender_row()` for each of these names, passing in the `gender_cache`. This yields `gender_patch`, a tibble with gender statistics for each split name, using the cache where possible and the NVB request pipeline otherwise.

The function then joins the split-name mapping back to `gender_patch` on `first_name_split`, drops rows that still lack gender information, and aggregates to the original `first_name` level. For first names with multiple split parts that have gender estimates, it computes the mean of `male_count`, `female_count`, and `probability_male` across the splits (coerced to integer for the counts). Finally, it uses `rows_update()` to update the original `gender` tibble with these patched values, returning a tibble where previously missing gender information has been filled in when possible.

```{r}
patch_gender_on_splits = function(gender){
  # set gender cache
  gender_cache = gender |> drop_na()
  # select authors without gender, and split their names
  selection = gender |>
    filter(is.na(male_count)) |>
    mutate(first_name_split = str_split(first_name, ' ')) |>
    unnest_longer(first_name_split) |>
    select(first_name, first_name_split)

  # get first names from selection
  first_names = selection |>
    select(first_name_split) |>
    pull() |>
    unique()

  # patch gender --------------------------------------
  gender_patch = purrr::map_dfr(
    first_names, get_gender_row, gender = gender_cache
  )

  # aggregate gender information
  gender_patch = selection |>
    left_join(
      gender_patch, 
      by=join_by(first_name_split == first_name)
    ) |>
    drop_na() |>
    # take the average gender count and probablity
    # for names where both splits yielded a gender
    # result
    group_by(first_name) |>
    summarise(
      male_count = as.integer(mean(male_count)),
      female_count = as.integer(mean(female_count)),
      probability_male = mean(probability_male)
    ) |>
    ungroup() 

  gender |> rows_update(gender_patch)
}

```

---

### `clean_gender()`
Creates a tidy gender summary for each first name based on gender probabilities and counts.

#### Arguments
**`data`**  

> Tibble or data frame containing at least `first_name`, `male_count`, `female_count`, and `probability_male` columns.

#### Method
This function takes a data frame with gender-count information and constructs a cleaner, more directly interpretable representation. It assigns a binary `gender` label based on whether `probability_male` is at least `0.5` (`"male"` if ≥ 0.5, otherwise `"female"`). The `count` variable is then set to `male_count` for rows classified as male and `female_count` for rows classified as female. The `prob` variable stores the probability that the assigned gender is correct: it is `probability_male` for male-labeled rows and `1 - probability_male` for female-labeled rows. Finally, the function keeps only `first_name`, `gender`, `prob`, and `count`, returning a compact tibble suitable for downstream analyses or merging.

```{r}
clean_gender = function(data){
  data |>
    mutate(
      gender = ifelse(
        probability_male >= 0.5,
        'male','female'
      ),
      count = ifelse(
        gender == 'male',
        male_count, female_count
      ),
      prob = ifelse(
        gender == 'male',
        probability_male, 1 - probability_male
      )
    ) |>
    select(first_name, gender, prob, count)
}
```

---

### `patch_missing_gender()`
Patches missing gender labels using curated lists of known female and male first names.

#### Arguments
**`data`**  

> Tibble or data frame containing at least `first_name` and `gender` columns, where some `gender` values may be missing.

#### Method
This function defines two curated character vectors, `female_names` and `male_names`, which contain first names that are known to be female or male, including several manually corrected or rare names. It then mutates the `gender` column using `case_when()`: existing non-missing `gender` values are preserved, names found in `female_names` are set to `"female"`, names found in `male_names` are set to `"male"`, and all other rows fall back to the original `gender` value via `.default = gender`. The result is a data frame where previously missing or ambiguous gender labels are patched based on these manually validated name lists, while leaving already coded cases unchanged.

```{r}
patch_missing_gender = function(data){
  female_names = c(
    "Alaxandra",   "Alinson",     "Avyanthi",
    "Brunilda",    "Busisiwe",    "Diliara",      
    "Dolive",      "Echo",        "Guangyu",
    # mistake in name, has been patched with _create_name_corrections
    "Guangye", 
    "Gul-i-Hina",  "Haebin",      "Haisu",
    # Phoebe Kisibi Mbasalaki was incorrectly coded, 
    # has been patched with _create_name_corrections 
    "Kisubi",      "Pheobe",
    "Liubov",      "Madalina",    "Majolijn",    
    "Mansoureh",   "Nankyung",    "Nilmawati",   
    "Nodira",      "Noyonika",    "Radostina",   
    "Rojika",      "Rozenmarijn", "Sayoni",
    "Seonoki",     "Shelliann",   "Shiming",     
    "Siggie",      "Siztine",     "Sungmi",      
    "Talinta",     "Teana",       "Xingna",
    "Yuliia",      "Zhiyi"
  )

  male_names = c(
    "Alborno",     "Chenchen",    "Chendi",
    "Chunglin",    "Chuyu",       "Diliara",
    "Gjovalin",    "Kirils",      "Kyohee",
    "Madhud",      "Quichen",     "Soeren",
    "Teana",       "Tanzhe",     "Vishwesh",
    "Weverthon"
  )

  data |>
    mutate(
      gender = case_when(
        !is.na(gender) ~ gender,
        first_name %in% female_names ~ 'female',
        first_name %in% male_names ~ 'male',
        .default = gender
      )
    )
}

```

---

### `scrape_gender()`
Scrapes, caches, and cleans gender information for a set of first names using the Meertens NVB data.

#### Arguments
**`idx`**  

> Tibble or data frame containing at least a `first_name` column with the names for which gender information should be obtained.

#### Method
The function first loads the existing NVB gender cache from `data/_cache/nvb_cache.Rds` and drops any rows with missing values to create a clean `gender_cache`. If the global flag `eval_ok` is `TRUE`, it extracts the unique, non-missing `first_name` values from `idx` and calls `get_gender_row()` for each of these names, passing along the `gender_cache` so that already-cached names can be reused and only new names are scraped. The resulting tibble is then passed through `patch_gender_on_splits()` to fill in missing gender information by splitting compound first names and aggregating estimates across the split parts. If `eval_ok` is `FALSE`, the function simply uses the existing `gender_cache` as `res` without making new requests.

The intermediate `res` (with any rows containing missing values removed) is written back to `data/_cache/nvb_cache.Rds`, updating the local cache. Finally, `res` is cleaned and standardized by applying `clean_gender()` (to derive a tidy `gender`, `prob`, and `count` representation) and `patch_missing_gender()` (to fix remaining missing or ambiguous labels using curated name lists). The function returns this fully processed tibble with gender information ready for downstream analysis or merging.

```{r}
scrape_gender = function(idx) {
  # load gender cache
  gender_cache = readRDS(file.path("data", "_cache", "nvb_cache.Rds")) |> 
    drop_na()

  if (eval_ok){
  first_names = idx |> pull(first_name) |> unique() |> na.omit()

  # scrape gender results
  res = purrr::map_dfr(
      first_names, 
      get_gender_row, 
      gender = gender_cache
    ) |>
    patch_gender_on_splits()
  } else {
    res =  gender_cache
  }

  # put gender scrape results
  saveRDS(res |> drop_na(), file.path('data', '_cache', "nvb_cache.Rds"))

  # clean gender results
  res = res |> 
    clean_gender() |>
    patch_missing_gender()

  return(res)
}

```

---

### `genderize_names()`
Fetches and caches gender information for a set of first names using the Genderize.io API.

#### Arguments
**`idx`**  

> Tibble or data frame containing at least a `term` column with first names (or name-like tokens) for which gender information should be retrieved.

#### Method
The function first loads an existing gender cache from `data/_cache/genderizer_cache.Rds`. It then reads the `GENDERIZE_API_KEY` from the environment (via `dotenv::load_dot_env()` and `Sys.getenv()`), which is required to authenticate requests to the Genderize.io API. From the input `idx`, it filters out any rows whose `term` already appears in the `name` column of the cached data, so that only uncached names are processed.

It extracts the remaining unique, non-missing `first_names` from `idx$term`. If the global flag `eval_ok` is `TRUE`, it loops over these names, calling `genderizeAPI(name, apikey = APIKEY)` for each, and stores each API response in a list `hold` keyed by the name. The new results in `hold` are combined with the existing `gender_cache` using `bind_rows()`, and duplicate rows are removed with `distinct(.keep_all = TRUE)` to ensure one record per name. The updated cache is written back to `data/_cache/genderizer_cache.Rds`, and the combined result (old cache plus any new observations) is returned for further use.

```{r}

genderize_names = function(idx) {
  # load cached gender information
  gender_cache = readRDS(file.path('data', '_cache', "genderizer_cache.Rds"))

  # load api key from secrets file
  dotenv::load_dot_env()
  APIKEY <- Sys.getenv("GENDERIZE_API_KEY")

  
  # select uncached names
  idx =  idx |> filter(!term %in% gender_cache$name)

  # select first_names
  first_names = idx$term |> na.omit() |> unique()
  first_names = first_names

  # fetch gender results
  hold = c()
  if (eval_ok) {
    for (name in first_names){
      resp = genderizeAPI(name, apikey = APIKEY)
      hold[[name]] = resp$response
    }
  }

  # combine cache with results and put new cache results
  res = bind_rows(gender_cache, bind_rows(hold)) |>
    distinct(.keep_all = TRUE)
  saveRDS(res, file.path('data', '_cache', "genderizer_cache.Rds"))

  return(res)
}

```

---

### `harmonize_gender()`

:::{.callout-note}
TODO: update description for function.
:::
Harmonizes gender information from multiple sources into a single, aggregated record per first name.

#### Arguments
**`gender`**  

> Tibble or data frame containing at least `first_name`, `term`, `gender.x`, `gender.y`, `count.x`, `count.y`, `prob`, and `probability` columns (typically resulting from a join between two gender sources, e.g. NVB and Genderize).

#### Method
The function starts by selecting the key identification and gender-related columns: `first_name`, `term`, all columns starting with `gender`, `count`, and `prob`. It then removes duplicate rows using `distinct(first_name, term, .keep_all = TRUE)` to ensure unique combinations of `first_name` and `term`. Two helper flags are created: `has_multiple`, which indicates if `first_name` contains a space or hyphen (potentially compound names), and `has_mismatch`, which marks cases where `gender.x` and `gender.y` disagree.

After dropping rows with missing `first_name`, it filters out problematic combinations where the name is compound (`has_multiple`), the two gender sources disagree (`has_mismatch`), and `term` contains a dot (typically indicating abbreviations or initials), as these are considered unreliable. Next, it resolves conflicts between sources: `gender` is set to `gender.y` when available and falls back to `gender.x` otherwise; `prob` is set to `probability` (e.g. from Genderize) when available and otherwise keeps the existing `prob`; `count` is set to `count.y` when present and falls back to `count.x` when missing.

Finally, the data is grouped by `first_name` and summarised: it takes the first non-missing `gender`, averages the `prob` across records, and sums the `count` to obtain a combined support count. The result is a tibble with one row per `first_name`, containing a harmonized gender label, an average probability, and a total count, integrating information from multiple gender sources in a principled way.

```{r}

harmonize_gender = function(gender) {
  gender |>
    mutate(
      gender = coalesce(gender.nvb, gender.g),
      gender_count = coalesce(count.nvb, count.g),
      gender_prob =coalesce(prob, probability)
    ) |>
    group_by(first_name) |>
    summarise(
      gender = first(gender, na_rm=TRUE),
      gender_prob = mean(gender_prob), 
      gender_count = sum(gender_count),
      gender.nvb = first(gender.nvb, na_rm=TRUE),
      gender.g = first(gender.g, na_rm=TRUE),
    ) |>
    patch_missing_gender()
}

```

---

## Application

```{r}
names = freadRDS2('names')

idx = names |>
  mutate(term = first_name |> 
    str_to_lower() |> 
    str_split('( |-)')
  ) |>
  unnest_longer(term)
res = scrape_gender(idx)
res2 = genderize_names(idx)
```


```{r}
gender = idx |>
  left_join(res) |>
  left_join(res2, by=join_by(term == name), suffix = c(".nvb", ".g")) |>
  harmonize_gender()

gender = names |> left_join(gender)
```

```{r}
fsaveRDS(gender, "gender")
```

### Example Data
```{r}
gender |> head()
```



