---
title: "Data Processing Ethnicity"
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions

params:
  eval: false
---

## Getting Started

```{r}
#| warning: false
#| output: false
# clear the global environment
rm(list = ls())
gc()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(readxl)
library(stringr)
library(lubridate)
library(httr2)
library(rvest)
library(xml2)
library(purrr)
library(RCurl)
library(fuzzyjoin)
library(stringi)
library(DemografixeR)

# load custom functions
source("src/utils/custom_functions.r")
.load_quarto_dependencies()
```

## Functions

### Implementation Nationalize.io

**`nationalize_name()`** — fetch nationality probabilities for names with persistent caching

Reads an existing cache of name→nationality results (`data/utils/origin_cache.Rds`), queries only uncached `search_name` values via the nationalize.io API, merges new results with the cache, de-duplicates, and writes the updated cache back to disk. Live API calls are performed only when a boolean flag `eval_ok` is `TRUE`; otherwise the function simply returns the current cache.

How it works:
- Loads `.env` with `dotenv::load_dot_env()` and reads the API key from `Sys.getenv("GENDERIZE_API_KEY")` (the Nationalize/Genderize/Agify services share the same key format).
- Filters out names already present in `origin_cache$name`.
- For remaining names (if `eval_ok` is `TRUE`), calls `nationalize(name, sliced = FALSE, apikey = APIKEY, simplify = FALSE)` and collects responses.
- Binds new rows with the cache and keeps distinct rows (`distinct(.keep_all = TRUE)`).
- Drops a top-level country column if present (some client returns a nested/list column).

```{r}
nationalize_name = function(idx) {
  origin_cache = readRDS(file.path('data', '_cache', "origin_cache.Rds"))

  # load api key from secrets file
  dotenv::load_dot_env()
  APIKEY <- Sys.getenv("GENDERIZE_API_KEY")

  # select uncached names
  idx =  idx |> filter(!search_name %in% origin_cache$name)

  # select last_names
  last_names = idx$search_name |> na.omit() |> unique() |> sort()
  # last_names = last_names[12:22]

  # fetch nationality results
  hold = c()
  if (-eval_ok) {
    for (name in last_names){
      resp = nationalize(name, sliced=FALSE, apikey = APIKEY, simplify = FALSE)
      hold[[name]] = resp 
    }
  }

  # combine cache with results and put new cache results
  origin = bind_rows(origin_cache, bind_rows(hold)) |> 
    distinct(.keep_all = TRUE)

  if ("country" %in% colnames(origin)) origin = origin |> select(-country)

  saveRDS(origin, file.path('data', '_cache', "origin_cache.Rds"))

  return(origin)
}
```

`nationalize_name` give multiple responses per individuals. The following function
selects the first response per name, which is the "best match"

```{r}

select_best_match = function(res){
  res |> 
    rename("last_name" = "name") |>
    # get the best match
    group_by(last_name) |>
    slice(1) 
}

```

### Scraper Configuration

This function builds a resilient GET request to a target page and returns a 
uniform result object instead of throwing on transport errors.

- **Success check.** `is_ok()` flags responses with HTTP status in the 200–299 range.
- **Browser-like request.** `request_last_name()` constructs an `httr2` request with a Mac Chrome user agent, Dutch/English Accept-Language, and broad Accept headers to mimic a real browser and reduce blocking.
- **Connection settings.** It sets a 30-second timeout and (deliberately) disables SSL peer verification to cope with misconfigured certificates on legacy servers. 
- **Retry policy.** On transient errors (HTTP 429 or 5xx), it retries up to 4 times with jittered exponential backoff `(runif(0.5–1.2) × 2^(try−1))`, which spreads load and avoids thundering herds.
- **Politeness delay.** If a global pause is set, it sleeps `pause + U(0, 0.4)` seconds before firing the request to throttle scraping.
- **Error handling.** The actual HTTP call is wrapped in `tryCatch()`. Instead of stopping, it returns a structured list:
  - **ok (logical)** – success per `is_ok()`
  - **status (integer)** – HTTP status or `NA` on transport error
  - **url (character)** – the requested URL
  - **resp (httr2 response)** – raw response on success, NULL on error
  - **error (condition)** – the caught error on failure

This design lets downstream code branch cleanly on `res$ok` and inspect `res$status` or `res$error` without breaking the pipeline.

```{r}
is_ok = function(resp) resp_status(resp) >= 200 && resp_status(resp) < 300

request_last_name = function(base_url, pause=0.5){
  # configure user agent
  ua = paste(
    "Mozilla/5.0 (Macintosh; Intel Mac OS X 15_5)",
    "AppleWebKit/537.36 (KHTML, like Gecko)",
    "Chrome/129.0.0.0 Safari/537.36"
  )

  req = request(base_url) |>
    req_user_agent(ua) |>
    # disable SSL verification
    req_options(ssl_verifypeer = 0) |>
    req_timeout(30) |>
    req_headers(
      "Accept" = paste0("text/html,application/xhtml+xml,",
                        "application/xml;q=0.9,*/*;q=0.8"),
      "Accept-Language" = "nl,en;q=0.8"
    ) |>
    req_retry(
      max_tries = 4,
      backoff = ~ runif(1, 0.5, 1.2) * (2 ^ (.x - 1)),  # jittered exponential
      is_transient = function(resp) {
        code <- resp_status(resp)
        isTRUE(code == 429L || (code >= 500L & code < 600L))
      }
    )

  # polite pause + jitter
  if (pause > 0) Sys.sleep(pause + runif(1, 0, 0.4))

  # CRITICAL: don't throw on transport errors
  resp <- tryCatch(
    req_perform(req),
    error = function(e) {
      attr(e, "nvb_url") <- url
      e
    }
  )

  # Return a uniform list the caller can inspect
  if (inherits(resp, "error")) {
    res = list(
      ok = FALSE,
      status = NA_integer_,
      url = url,
      resp = NULL,
      error = resp
    )
  } else {
    res = list(
      ok = is_ok(resp),
      status = resp_status(resp),
      url = url,
      resp = resp,
      error = NULL
    )
  }
  return(res)
}
```


### Configure URL and Extracters

#### `format_url()` – Build a query URL for CBG Familienamen
Takes a name and constructs a browser-ready URL for the CBG surnames site, encoding spaces as + and adding query parameters for multiple name fields. Optional `.what = "info"` appends the path to the analysis/etymology page. Returns the full URL string.

```{r}
format_url = function(
    name,
    base = "https://www.cbgfamilienamen.nl/nfb/detail_naam.php?",
    .what = "base"
  ){
  # format_name for URL
  formatted_name = name |> 
    URLencode(reserved = TRUE) |>
    str_replace_all(pattern = "%20", '+')

  # configure url
  url = paste0(
    base,
    "gba_naam=", formatted_name,
    "&gba_lcnaam=", tolower(formatted_name),
    "&nfd_naam=", formatted_name
  )

  if (.what == 'info'){
    url = paste0(url, "&info=analyse+en+verklaring")
  }
  
  return(url)
}

```

#### `extract_info()` – Parse analysis/etymology text from response
Given a response wrapper `r` (with `r$resp` from `httr2`), reads the HTML body and extracts the “kenmerken/verklaring” section. It splits the page text, trims boilerplate until the © footer, and returns a single "; "-separated string with the extracted info (or "" if not found).

```{r}
extract_info = function(r){
  html = read_html(resp_body_string(r$resp))
  text = html |>
    html_element("body") |>
    html_text()

  info = c("")
  if (str_detect(text, regex('kenmerken:|verklaring:'))){
    parts = str_split_fixed(text, pattern=regex('kenmerken:|verklaring:'), 2)[2] |>
      str_split(pattern = regex("[\\n|\\t|\\s]{3,20}")) |>
      unlist()

    idx = which(grepl(regex("©"), parts))

    if (length(idx) >= 1) {
      info = parts[seq_len(idx[1] - 1)]
      info = as.vector(info[nzchar(info)])
    }
  }

  # drop empty strings
  # info = info[nzchar(info)]
  return(paste(unname(info), collapse = "; "))
}

safe_info = function(r, default = NA_character_) {
  tryCatch(extract_info(r),
           error = function(e) default,
           warning = function(w) default)
}

```

#### `extract_count()` – Retrieve occurrence count from tables
Parses all HTML tables from `r$resp` and searches for the table layout containing the national count. If found, pulls the value at row 2, column 2 (per the site’s structure) and returns it as an integer; throws an error if no tables are present and returns `NA` when the expected layout is missing.

```{r}
extract_count = function(r){
  # extract all tables on the page
  html = read_html(resp_body_string(r$resp))
  tables = html_table(html, header=FALSE)

  # select the first table if a table if found
  if (length(tables) == 0) stop("No tables were found")
  
  # set count value
  count = NA_integer_
  if (length(tables) >= 4) {
    for (tab in tables){
      i = nrow(tab)
      if (i >= 5){
        count = tab[2,2] |> pull()
      }
    }
  }

  return(count)
}

safe_count = function(r, default = NA_integer_) {
  tryCatch(extract_count(r),
           error = function(e) default,
           warning = function(w) default)
}

```

### Tidy Scrape Wrappers

These utilities wrap the earlier extractors and assemble a tidy result for a single surname.
- `safe_count()` / `safe_info()` wrap `extract_count()` and `extract_info()` in `tryCatch()`, returning a default (`NA_integer_` / `NA_character_`) on any error or warning. This guarantees downstream code receives a value even when pages are missing or malformed.
- `get_name_row()` takes a name and performs up to two requests:
	1.	Builds the base URL with `format_url(name)` and fetches it via `request_last_name()`. If the HTTP result is ok, it parses the national occurrence count with `safe_count()`; otherwise it records `NA`.
	2.	Only if a non-missing count was obtained, it builds the “info” URL (`.what = "info"`) and requests it, then extracts the analysis/etymology info with `safe_info()` (again only when both requests succeeded).

Both `count` and `info` are coerced to character for consistency. The function returns a one-row tibble with `last_name`, `name_count`, and `info`, providing a compact, fault-tolerant record for each queried surname.


```{r}

get_name_row = function(name, count=NA_character_, info=NA_character_){
  # scrape the count information
  r1 = format_url(name) |>
    request_last_name() 

  count = if (isTRUE(r1$ok)) safe_count(r1) else NA_character_
  count = as.character(count)

  # scrape info if the first scrape yielded success
  if (!is.na(count)){
  r2 = format_url(name, .what="info") |>
    request_last_name()

  info = if (isTRUE(r2$ok) && isTRUE(r1$ok)) safe_info(r2) else NA_character_
  info = as.character(info)
  }


  as_tibble(list(
    last_name = name, 
    name_count = count,
    info = info
  ))
}
```


## patch ethnic origin 

**`add_origin5()`** — attach origin labels from a curated patch via fuzzy last-name matching

Loads a manual patch file (`data/utils/origin_patch.xlsx`), normalizes last names with `normalize_name()`, and performs a fuzzy left join on last_name_norm (maximum distance 0.5) to transfer an origin label. The matched label is renamed to `origin5`. Helper/diagnostic columns from the join (e.g., normalized names, counts) are dropped; unmatched rows are left unchanged.


### scrape nationality per row from achternamen bank

**`name_to_origin()`** — resolve origin info for names with on-disk caching

Loads a cached table of name to origin results (`data/utils/cbg_cache.Rds`), looks up only the uncached names from i`dx$search_name` using a resolver `get_name_row()`, appends new rows to the cache, saves the updated cache back to disk, and returns the combined result.å

How it works:
- Reads `cbg_cache.Rds` (expected to contain a last_name column used as the cache key).
- Filters `idx` to names not already in `ethnicity_cache$last_name`.
- Iterates over unique `search_name` values, calls `get_name_row(name)` for each miss, collects rows.
- åBinds new rows with the existing cache and saves the result.

```{r}
name_to_origin = function(idx) {
  ethnicity_cache = readRDS(file.path('data', '_cache', 'cbg_cache.Rds'))

  # select uncached names
  idx =  idx |> filter(!search_name %in% ethnicity_cache$last_name)

  # select last_names
  last_names = idx$search_name |> na.omit() |> unique() |> sort()

  # loop over remaining names
  hold = c()
  for (name in last_names){
    if (!name %in% ethnicity_cache$last_name){
      hold[[name]] = get_name_row(name)
    }
  }

  # combine cache with hold
  res = bind_rows(ethnicity_cache, bind_rows(hold))
  saveRDS(res,  file.path('data', '_cache', 'cbg_cache.Rds'))

  return(res)
}
```

```{r}
parse_origin = function(data){
  # 1) de herkomst staat vaak voor naam, achternaam, etc.
  origin_pattern = regex(paste0(
    "\\b[[:upper:]][[:lower:]]{2,}(?=\\s+(?:achter|familie|beroeps)?naam?\\b)"
  ))

  # 1c) detect particles
  particles_pattern = regex(paste0(
    "(de |den |der |het |te |ten |ter",
    "|van |van de |van den |van der |van 't |van ’t",
    "|'t |’t |op de)"
  ))

  # 2) patroniem/toponiem/etc. en taal/naam-typen in één keer definieren
  type_pattern <- regex(paste0(
    "(toponiem|patroniem|herkomstnaam|adresnaam|beroepsnaam|metonymische",
    "|beroepsnaam|samenstelling met voornaam|metafoor|betrekking)"
    ), ignore_case = TRUE
  )

  # 3) bijvoegelijk naamwoord of taal
  lang_pattern <- regex(paste0(
      "(Afrikaans(?:e)|Antilliaans(?:e)|Aziatisch(?:e)|Belgisch(?:e)|Catalaans(?:e)",
      "|Chines(?:e)|Deens(?:e)|Duits(?:e)|Egyptisch(?:e)|Engels(?:e)",
      "|Frans(?:e)|Fries(?:e)|Germaans(?:e)|Grieks(?:e)|Hongaars(?:e)",
      "|Iers(?:e)|Indias(?:e)|Indisch(?:e)|Italiaans(?:e)|Joegoslavisch(?:e)",
      "|Marokkaans(?:e)|Moluks(?:e)|Nederlands(?:e)|Pakistaans(?:e)|Koreaans(?:e)",
      "|Portuge(?:e)s(?:e)|Surinaams(?:e)|Turks(?:e)|Vietname(?:e)s(?:e))"
    ), ignore_case = TRUE
  )

  # 4) landen/gebieden die in jouw voorbeelden voorkomen
  country_pattern <- regex(paste0(
      "(Turkije|Pakistan|Afghanistan|India|Brits-Guyana|Suriname|Midden-Oosten",
      "|Zuidoost-Azi[eë]|Afrika|Antillen|Azi[eë]|Belgi[eë]|Cataloni[eë]",
      "|China|Denemarken|Duitsland|Egypte|Engeland|Verenigd Koninkrijk",
      "|Frankrijk|Friesland|Griekenland|Hongarije|Ierland|Itali[eë]",
      "|Joegoslavi[eë]|Marokko|Molukken|Nederland|Korea|Zuid-Korea|Noord-Korea",
      "|Portugal|Vietnam)"
    ), ignore_case = TRUE
  )
    
  # apply regex patterns
  data = data |>
    mutate(
      # 1) extract origin using general pattern
      o1 = str_extract(info, origin_pattern),
      o2 = str_extract_all(info, origin_pattern),
      o2 = map(o2, ~ if (length(.x) >= 1) .x[-1] else NULL),

      # 1b) indicates whether it is a regional name
      r1 = str_detect(info, paste0(
        "(dorp)|(plaats)|(gemeente)",
        "|(graafschap)|(stad)|(deel)|(Friesland)"
      )),
      # 1c) indicates whether the name contains a particle
      p1 = str_detect(last_name, particles_pattern),

      # 2) extract type of name
      t1 = str_extract(info, type_pattern),
      t2 = str_extract_all(info, type_pattern),
      t2 = map(t2, ~ if (length(.x) >= 1) .x[-1] else NULL),

      # 3) extract language or preverb
      l1 = str_extract(info, lang_pattern),
      l2 = str_extract_all(info, lang_pattern),
      l2 = map(l2, ~ if (length(.x) >= 1) .x[-1] else NULL),

      # 4) extract countries
      c1 = str_extract(info, country_pattern),
      c2 = str_extract_all(info, country_pattern),
      c2 = map(c2, ~ if (length(.x) >= 1) .x[-1] else NULL)
    ) |>
    select(-info)

  return(data)
}
```

```{r}
patch_origin = function(data){
  # load origin patch
  ethnicity_patch = readxl::read_excel(
      file.path('data', 'utils', 'origin_patch.xlsx')
    ) |>
    mutate(last_name_norm = normalize_name(last_name))

  # fuzzy match origin patch.
  data = data |>
    mutate(last_name_norm = normalize_name(last_name)) |>
    fuzzyjoin::stringdist_left_join(
      ethnicity_patch,
      by = "last_name_norm",
      max_dist = 0.5
    ) |>
    rename(
      "last_name" = "last_name.x",
      "o3" = "origin"
    ) |>
    mutate(o3 = str_remove(o3, regex('\\ naam', ignore_case = TRUE))) |>
    select(
      -last_name_norm.x, -count, 
      -last_name_norm.y, -last_name.y,
    )

  return(data)
}
```


```{r}

clean_origin = function(data){
  data = data |>
    mutate(
      origin = case_when(
        !is.na(o1) & !str_detect(o1, 'Christelijke') ~ o1,
        !is.na(l1) ~ l1,
        !is.na(c1) ~ c1,
        .default = NA_character_
      ),
      origin = ifelse(p1, 'Nederlandse', origin),
      origin = ifelse(!is.na(t1) & is.na(origin), 'Nederlandse', origin),
      origin = ifelse(isTRUE(r1) & is.na(origin), 'Nederlandse', origin),
      origin = ifelse(!is.na(o3), o3, origin),
      .after = last_name
    ) |>
    mutate(
      origin = str_replace(origin, "Als", "Nederlandse") |>
        str_replace("Azië", "Aziatisch") |>
        str_replace("Catalaanse", "Spaanse") |>
        str_replace("Engelstalig", "Engelse") |>
        str_replace("Catalaanse", "Spaanse") |>
        str_replace("germaanse", "Germaanse") |>
        str_replace("Germaanse", "Nederlandse") |>
        str_replace("Joegoslavië", "Joegoslavische") |>
        str_replace("Joodse", "Nederlandse") |>
        str_replace("nederlandse", "Nederlandse") |>
        str_replace("Pakistan", "Pakistaanse") |>
        str_replace("Turkije", "Turkse")|>
        str_replace("Zuid-Korea", "Zuid-Koreaanse")|>
        str_replace("Koreaanse", "Zuid-Koreaanse")
        
    ) |>
    select(!o1:o3)

  return(data)
}

  
```

## Harmonize ethnicity variables

**`harmonize_ethnicity()`** — map origin signals to ISO alpha-2 codes and attach country names

This function consolidates multiple origin indicators into a standardized two-letter country code and enriches it with a readable country name. It (1) normalizes name_count, (2) assigns `ethnicity` via rule-based matches on Dutch origin adjectives with a frequency threshold, (3) applies a set of surname-specific overrides, (4) blanks inconsistent statistics, and (5) joins ISO codes to add `country_name`.

How it works:
- Reads ISO codes from `data/utils/iso3611_codes.xlsx` (which include `alpha_2`, `country_name`).
- Cleans name_count: converts `"< 5"` to `5`, casts to integer, replaces NA with 0.
- Sets `ethnicity` using heuristics:
  - If `origin` contains "Nederlandse", "Chinese", "Duitse", "Marokkaanse", or "Turkse" and name_count > 200, assign "NL", "CN", "DE", "MA", "TR" respectively; otherwise fall back to country_id.
  - Applies bespoke overrides based on search_name (e.g., "Wachter" → "NL", "Metinsoy" → "TR", "Phillips" → "BR", etc.), replacing any prior value.
  - If the chosen `ethnicity` differs from `country_id`, sets count and probability to NA to avoid mixing mismatched stats.
- Keeps relevant columns, removes duplicates (distinct on uid, search_name, ethnicity), and left-joins ISO to add country_name.

```{r}
harmonize_ethnicity = function(origin) {
  iso = readxl::read_xlsx(file.path('data', 'utils', 'iso3611_codes.xlsx'))

  origin = origin |>
    mutate(
      name_count = name_count |> str_replace('< 5', '5') |> as.integer(),
      name_count = ifelse(is.na(name_count), 0, name_count),
      ethnicity = case_when(
        str_detect(origin, 'Nederlandse') & (name_count > 400) ~ "NL",
        str_detect(origin, 'Chinese') & (name_count > 200) ~ "CN",
        str_detect(origin, 'Duitse') & (name_count > 200) ~ "DE",
        str_detect(origin, 'Marokkaanse') & (name_count > 200) ~ "MA",
        str_detect(origin, 'Turkse') & (name_count > 200) ~ "TR",
        .default = country_id
      ),
      ethnicity = case_when(
        str_detect(search_name, 'Wachter') ~ "NL",
        str_detect(search_name, 'Das') ~ "NL",
        str_detect(search_name, "Metinsoy") ~ "TR",
        str_detect(search_name, "Wagner") ~ "DE",
        str_detect(search_name, "Mos") ~ "NL",
        str_detect(search_name, "Vlieg") ~ "NL",
        str_detect(search_name, "Knigge") ~ "NL",
        str_detect(search_name, "Bol") ~ "NL",
        str_detect(search_name, "Deen") ~ "NL",
        str_detect(search_name, "Vrooman") ~ "NL",
        str_detect(search_name, "Verdun") ~ "NL",
        str_detect(search_name, "Merlo") ~ "NL",
        str_detect(search_name, "Phillips") ~ "BR",
        str_detect(search_name, "Raekstad") ~ "NL",
        str_detect(search_name, "Aspide") ~ "IT",
        .default = ethnicity
      ),
      count = case_when(
        country_id != ethnicity ~ NA,
        .default = count
      ),
      probability = case_when(
        country_id != ethnicity ~ NA,
        .default = probability
      )
    ) |>
    select(uid, search_name, ethnicity, count, probability) |>
    distinct(uid, search_name, ethnicity, .keep_all=TRUE) |>
    left_join(
      iso |> select(alpha_2, country_name),
      by = join_by(ethnicity == alpha_2)
    )
}
```

## Application

```{r}
#| warning: false
#| output: false

names = freadRDS2('names')

# create an name index
idx = names |>
  distinct(particle, last_name, .keep_all = TRUE) |>
  unite(search_name, particle:last_name, sep=" ", na.rm=TRUE, remove = FALSE) |>
  unite(search_name_full, particle:maiden_name, sep=" ", na.rm=TRUE, remove = FALSE) 

# scrape using nationalizer
res = nationalize_name(idx) |>
  select_best_match()

# scrape using family name database
res2 = name_to_origin(idx) |>
  parse_origin() |>
  patch_origin() |>
  clean_origin()

# combine scraped information
origin = idx |>
  left_join(res, by=join_by(search_name==last_name)) |>
  left_join(res2, by=join_by(search_name==last_name)) |>
  select(uid, last_name, search_name, type:name_count) |>
  harmonize_ethnicity()

fsaveRDS(origin, 'ethnicity')

```
