---
title: Prune Scrape Data
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions

execute: 
  eval: false
---

## Getting Started

```{r}
#| warning: false
#| output: false
# load custom functions
source("src/utils/custom_functions.r")

# clear the global environment and set dependencies
.clear_global_environment()
.load_quarto_dependencies()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(stringr)
library(lubridate)
library(purrr)
```

## Functions

### `prune_invalid_publication_types()`
Filter out rare or irregular publication types and sort yearly counts.

#### Arguments
**`data`**  

> Named list that contains a `works` element, where `data[['works']]` is a named list of tibbles with OpenAlex works per scholar.

#### Method
The function first inspects all works by binding `data[['works']]` into one tibble and counting how often each `type` occurs. It defines a vector of valid types by keeping only types with at least 50 occurrences and explicitly excluding `dataset`. It then loops over the first ten elements of `data[['works']]`, skipping empty or invalid tibbles, and filters each scholar’s works to only those with a valid `type`. If a `counts_by_year` column is present, it carefully reorders each nested data frame by `year`, guarding against malformed or NULL entries. The updated per-scholar tibbles are written back into `data[['works']]`, and the pruned list is returned.

```{r}
prune_invalid_publication_types = function(data) {
  # excluding rare or irregular publication types from the works object. 
  # - data inspection showed a spike in datasets being published in 2020.
  valid_types = data[['works']] |> 
    bind_rows() |>
    group_by(type) |>
    summarise(n= n()) |>
    # excluding types with fewer than 50 hits (retraction: 1, libguides: 18, 
    # reference-entry: 39), and datasets (991). 
    filter(
      n >= 50, 
      type != 'dataset'
    ) |> 
    pull(type)

  # slice works object per author and only select valid publication types
  for (id in names(data[['works']][1:10])){
    tab = data[['works']][[id]]

    if(is.null(tab) | !is_tibble(tab) | nrow(tab) < 1) next
    tab = tab |> filter(type %in% valid_types) 
    
    # sort the counts_by_year table
    if(is_tibble(tab) && 'counts_by_year' %in% colnames(tab)) {
      tab = tab |>
        mutate(
          counts_by_year = map(
            counts_by_year,
            ~ if (!is.null(.x) && is.data.frame(.x) && "year" %in% names(.x)) {
                arrange(.x, year)
              } else {
                .x
              }
          )
        )
    }
    data[['works']][[id]] = tab
  }
  return(data)
}
```

### `get_author_id_lookup()`
Construct a lookup of intra-network co-authors across all works.

#### Arguments
**`data`**  

> Named list containing a `works` element, where `data[['works']]` is a named list of tibbles with OpenAlex works per scholar (including `uid`, `author_id`, and nested `authorships`).

#### Method
The function iterates over `data[['works']]` and, for each focal scholar `id`, first builds a global lookup of all `(uid, author_id)` pairs present in the combined works, excluding rows where `uid == id`. It then unnests each scholar’s `authorships` into a flat table and filters rows where `authors__author_id` is present in this global co-author lookup. Empty results are reset to `NULL`. After processing all scholars, the function binds all intermediate tables, de-duplicates `(uid, author_id, authors__author_id)` triples, and drops self-ties (`author_id != authors__author_id`). The resulting tibble captures pairs of focal authors and their intra-network co-authors, and is used downstream to prune works to scholars with co-publishing ties inside the observed network.

```{r}
get_author_id_lookup = function(data){

  # iterate over works, and select scholars with intranetwork ties
  hold = list()
  for (id in names(data[['works']])){
    # set lookup with intranetwork ties
    lookup = data[['works']] |> bind_rows() |>
      distinct(uid, author_id) |> drop_na() |>
      filter(uid != id) |> pull(author_id)

    # ids = lookup 
    tab = data[['works']][[id]] |>
      mutate(authors = authorships, .after = authorships) |>
      unnest(authors, names_sep = "__", keep_empty=FALSE) |>
      filter(authors__author_id %in% lookup)

    # set tab to missing if scholar has no intranetwork ties
    if (nrow(tab) == 0) tab = NULL

    hold[[id]] = tab
  }
  
  # combine valid scholars and create a lookup matrix with unique values
  # voor all author identifiers.
  res = hold |> bind_rows() |>
    distinct(uid, author_id, authors__author_id) |>
    filter(author_id != authors__author_id)

  return(res)
}
```

### `prune_works_by_author_id()`
Restrict works to scholars and author IDs with intra-network ties and fix types.

#### Arguments
**`data`**  

> Named list containing a `works` element (list of tibbles with `uid`, `author_id`, `doi`, `counts_by_year`, `oa_url`) and other components such as demographics and scholars.

#### Method
The function first calls `get_author_id_lookup()` to identify combinations of `uid` and `author_id` that participate in intra-network co-authorships. It then iterates over `data[['works']]`, filtering each scholar’s works to those rows where `uid` and `author_id` appear in this lookup. To avoid type inconsistencies across authors, it coerces logical `doi` values to `NA_character_`, logical `counts_by_year` to empty lists, and logical `oa_url` to `NA_character_`. If a filtered table ends up empty, it is replaced by an empty tibble. The pruned list of works is written back into `data[['works']]`, and the updated list object is returned.

```{r}
prune_works_by_author_id = function(data){
  lookup = get_author_id_lookup(data)

  works = list()
  for (id in names(data[['works']])){
    tab = data[['works']][[id]] |>
      filter(
        uid %in% unique(lookup$uid)
        & author_id %in% unique(lookup$author_id)
      ) |>
      mutate(
        doi = ifelse(is.logical(doi), NA_character_, doi),
        counts_by_year = ifelse(is.logical(counts_by_year), list(), counts_by_year),
        oa_url = ifelse(is.logical(oa_url), NA_character_, oa_url)
      )

    if (nrow(tab) == 0) tab = tibble()
    works[[id]] = tab
  }

  data[['works']] = works
  return(data)
}
```

### `prune_demographics()`
Drop demographic records for scholars without valid works.

#### Arguments
**`data`**  

> Named list that contains a `works` element (list of works tibbles with `uid`) and a `demographics` tibble keyed by `uid`.

#### Method
The function constructs a set of valid `uid`s by binding all works into a single tibble and extracting distinct `uid` values. It then filters `data[['demographics']]` to keep only rows with `uid` in this valid set, effectively removing demographic entries for scholars who do not have any retained works after pruning. The updated list is returned.

```{r}
prune_demographics = function(data){
  # create list of uids that were succesfully scraped
  valid_uids = data[['works']] |> bind_rows() |> distinct(uid) |> pull()

  # read in affiliations
  data[['demographics']] = data[['demographics']] |>
    filter(uid %in% valid_uids)

  return(data)
}
```

### `prune_scholars()`
Remove OpenAlex scholar records that lack valid author IDs after pruning.

#### Arguments
**`data`**  

> Named list containing a `works` element (with `author_id`) and a `scholars_oa` element, where `data[['scholars_oa']]` is a named list of OpenAlex author tibbles.

#### Method
The function first collects the set of valid `author_id`s by binding all works and extracting distinct `author_id`. It then loops over each element in `data[['scholars_oa']]`, skipping empty tibbles, and for each keeps only rows whose `author_id` (after stripping the `https://openalex.org/` prefix) is present in this valid set. The pruned per-scholar tibbles are stored in a new list, which replaces `data[['scholars_oa']]`. This ensures that the scholar metadata is consistent with the pruned works.

```{r}
prune_scholars = function(data){
  # create list of uids that were succesfully scraped
  valid_uids = data[['works']] |> bind_rows() |> distinct(author_id) |> pull()

  # read in affiliations
  hold = list()
  for (id in names(data[['scholars_oa']])) {
    tab = data[['scholars_oa']][[id]]
    if (nrow(tab) == 0) {
      hold[[id]] = tab
      next
    }

    hold[[id]] = tab |>
      mutate(author_id = str_remove(author_id, "^https://openalex.org/")) |> 
      filter(author_id %in% valid_uids)
  }

  data[['scholars_oa']] = hold
  return(data)
}
```

### `clean_demographics()`
Add region information and standardize demographic variable names.

#### Arguments
**`data`**  

> Named list with a `demographics` tibble (including `ethnicity`, `prob`, `count`, `universiteit_*`, `email_adres_*`, `functie_*`, `discipline_*`).

#### Method
The function first reads a country conversion table from `data/utils/country_conversion.csv`, keeps distinct `(alpha_2, region)` pairs, and standardizes the region labels, forcing `NL` to "Netherlands" and merging "Asia" and "Oceania" into "Asia and Oceania". It then left-joins this table onto `data[['demographics']]` by matching `ethnicity` to `alpha_2`, relocates the new `region` column ahead of `ethnicity`, and renames it to `origin`. It also renames several probability/count and longitudinal university/email/position columns to more concise, English variable names. Finally, it selects a compact set of columns from `uid` and `clean_name` through `discipline_25`, writes the cleaned tibble back into `data[['demographics']]`, and returns the updated list.

```{r}
clean_demographics = function(data) {
  # get country_conversion table
  country_conversion = read_csv(
      file.path('data', 'utils', 'country_conversion.csv'),
      show_col_types = FALSE
    ) |>
    distinct(alpha_2, region) |>
    mutate(
      region = ifelse(str_detect(alpha_2, 'NL'), 'Netherlands', region) |>
        str_replace('(Asia|Oceania)', 'Asia and Oceania') 
    )

  # add region to demographics and clean the dataframe
  data[['demographics']] = data[['demographics']] |> 
    left_join(country_conversion, by=join_by(ethnicity == alpha_2)) |>
    relocate(region, .before = ethnicity) |>
    rename(
      origin = region,
      gender_prob = prob,
      gender_count = count,
      university_22 = universiteit_22,
      university_24 = universiteit_24,
      university_25 = universiteit_25,
      email_22 = email_adres_22,
      email_24 = email_adres_24,
      email_25 = email_adres_25,
      position_22 = functie_22,
      position_24 = functie_24,
      position_25 = functie_25,
    ) |>
    select(uid, clean_name, gender, origin, ethnicity, email_22:discipline_25)

  return(data)
}
```

### `h_index_vec()`
Compute the h-index from a vector of citation counts.

#### Arguments
**`citations`**  

> Numeric or integer vector of citation counts (possibly containing NAs or non-finite values).

#### Method
The function first handles empty or NULL inputs by returning 0L. It then keeps only finite, non-missing citation counts, coerces them to integers, and sorts them in decreasing order. Using this ordered vector, it computes the h-index as the number of positions `i` where `c[i] >= i`, i.e., the largest value `h` such that at least `h` papers have at least `h` citations. The result is returned as an integer scalar.

```{r}
# 1) Core h-index calculator -----------------------------------------------
h_index_vec = function(citations) {
  if (is.null(citations) || length(citations) == 0L) return(0L)
  c = sort(as.integer(citations[is.finite(citations) & !is.na(citations)]), decreasing = TRUE)
  if (!length(c)) return(0L)
  # With c sorted desc, h is the count of entries with c_i >= i
  as.integer(sum(c >= seq_along(c)))
}
```

### `extract_citations()`
Robustly extract citation counts from various OpenAlex “works” structures.

#### Arguments
**`works_obj`**  

> Object containing works data; can be a data frame, a list with a `results` data frame, or a list of individual work objects with a `cited_by_count` element.

#### Method
The function defines an internal helper `df_to_cites()` that de-duplicates rows by `id` (if present) and then pulls `cited_by_count` as integers. It then branches on the type of `works_obj`. If it is a data frame, it calls `df_to_cites()` directly. If it is a list and has a `results` component, it applies `df_to_cites()` to `works_obj$results`. Otherwise, it assumes a list of work records and maps over them, extracting `cited_by_count` with `tryCatch()` to guard against malformed entries; NAs are dropped. For any unsupported object type, it returns an empty integer vector. The result is a clean integer vector of citation counts ready for index calculations.

```{r}
# 2) Robust extractor for cited_by_count from different "works" shapes ------
# Handles:
# - data.frame/tibble with column `cited_by_count`
# - list with element `results` that is a data.frame (e.g., API wrappers)
# - list of work-objects where each has $cited_by_count
# - optional de-duplication by `id` if present
extract_citations = function(works_obj) {
  # helper to deduplicate & pull citations from a data.frame
  df_to_cites = function(df) {
    if (!is.data.frame(df)) return(integer())
    if ("id" %in% names(df)) df = distinct(df, id, .keep_all = TRUE)
    if ("cited_by_count" %in% names(df)) return(as.integer(df$cited_by_count))
    integer()
  }

  if (is.data.frame(works_obj)) {
    return(df_to_cites(works_obj))
  }

  if (is.list(works_obj)) {
    # common wrapper: list(results = data.frame(...))
    if ("results" %in% names(works_obj)) return(df_to_cites(works_obj$results))

    # otherwise assume a list of work records
    cites = map_dbl(works_obj, ~ tryCatch(
      {
        v = .x[["cited_by_count"]]
        if (is.null(v)) NA_real_ else as.numeric(v)
      },
      error = function(e) NA_real_
    ))
    return(as.integer(cites[!is.na(cites)]))
  }

  integer()
}
```

### `h_index_by_author()`
Compute h-index and summary citation statistics per author.

#### Arguments
**`works_by_uid`**  

> Named list where `names(works_by_uid)` are scholar `uid`s and each value is that author’s works object (any structure accepted by `extract_citations()`).

#### Method
The function iterates over the named list with `imap_dfr()`. For each `(uid, works)` pair, it calls `extract_citations()` to obtain citation counts, then calculates three statistics: the number of works (`n_works`), the total citations (`total_citations`), and the h-index via `h_index_vec()`. These per-author rows are combined into a tibble and sorted in descending order of `h_index` and `total_citations`. The resulting table serves as a compact performance summary by scholar.

```{r}
# 3) Compute per-author h-index from a named list ---------------------------
# Input: works_by_uid — a named list where names(works_by_uid) are `uid`s,
#         each value is that author's OpenAlex works object
h_index_by_author = function(works_by_uid) {
  imap_dfr(works_by_uid, function(works, uid) {
    cites = extract_citations(works)
    tibble(
      uid             = uid,
      n_works         = length(cites),
      total_citations = sum(cites, na.rm = TRUE),
      h_index         = h_index_vec(cites)
    )
  }) |>
    arrange(desc(h_index), desc(total_citations))
}
```

### `i_index_vec()`
Compute the iN-index (default i10) from citation counts.

#### Arguments
**`citations`**  

> Numeric or integer vector of citation counts.

**`threshold`**  

> Integer threshold `N` for the iN-index (default `10L`), counting works with at least this many citations.

#### Method
The function keeps only finite, non-missing citation counts, coerces them to integers, and returns the number of works whose citation count is greater than or equal to the specified `threshold`. If the filtered vector is empty, it returns 0L. With `threshold = 10`, this corresponds to the classic i10-index.

```{r}
# 3) Compute per-author i10-index from a named list ---------------------------
# Input: works_by_uid — a named list where names(works_by_uid) are `uid`s,
#         each value is that author's OpenAlex works object
i_index_vec = function(citations, threshold = 10L) {
  c = as.integer(citations[is.finite(citations) & !is.na(citations)])
  if (!length(c)) return(0L)
  sum(c >= as.integer(threshold))
}
```

### `i_index_by_author()`
Compute the per-author iN-index (default i10) and citation summary.

#### Arguments
**`works_by_uid`**  

> Named list where `names(works_by_uid)` are scholar `uid`s and each value is that author’s works object.

**`threshold`**  

> Integer threshold `N` for the iN-index (default `10L`).

#### Method
The function constructs a dynamic column name such as `"i10_index"` from the chosen `threshold`. It then iterates over the named list with `imap_dfr()`, extracting citation counts for each author via `extract_citations()`. For each `uid`, it records the number of works, total citations, and the iN-index computed by `i_index_vec(citations, threshold)`. The resulting tibble is sorted by the iN-index (descending) and then by `total_citations`. This table complements the h-index summary by focusing on the volume of works above a fixed citation threshold.

```{r}
# Per-author iN (defaults to i10). Input: named list 'works_by_uid'
i_index_by_author = function(works_by_uid, threshold = 10L) {
  colname = paste0("i", threshold, "_index")
  imap_dfr(works_by_uid, function(works, uid) {
    cites = extract_citations(works)
    tibble(
      uid             = uid,
      n_works         = length(cites),
      total_citations = sum(cites, na.rm = TRUE),
      !!colname       := i_index_vec(cites, threshold = threshold)
    )
  }) |>
    arrange(desc(.data[[colname]]), desc(total_citations))
}
```

### `add_updated_stats_to_demographics()`
Attach h-index and i10-index statistics to the demographics table.

#### Arguments
**`data`**  

> Named list containing a `works` element (works per `uid`) and a `demographics` tibble keyed by `uid`.

#### Method
The function first computes per-author h-index statistics with `h_index_by_author(data[['works']])`. It also computes per-author i10-indices via `i_index_by_author(data[['works']])` and keeps only `uid` and `i10_index`. These two summary tables are then left-joined onto `data[['demographics']]` by `uid`, enriching each scholar’s demographic record with updated citation metrics. The augmented `demographics` tibble is written back into `data`, and the enhanced list object is returned.

```{r}
# Wrapper function addint these data to demographics
add_updated_stats_to_demographics = function(data){
  h_index = h_index_by_author(data[['works']]) 
  i10_index = i_index_by_author(data[['works']]) |> select(uid, i10_index)

  data[['demographics']] = data[['demographics']] |> 
    left_join(h_index) |>
    left_join(i10_index)

  return(data)
}
```

### `add_edge_table()`
Build an edge table of co-authorships from nested authorship data.

#### Arguments
**`data`**  

> Named list that contains a `works` element (list of works tibbles with `uid`, `work_id`, `authorships`, `publication_date`, `publication_year`, `type`).

#### Method
The function binds all per-scholar works together, selecting `uid`, `work_id`, `authorships`, `publication_date`, `publication_year`, and `type`. It renames `authorships` to `a` and fully unnests `a` into separate columns with a names separator. It then filters to rows where `a__uid` (the co-author’s `uid`) is non-missing. The resulting tibble, stored as `data[['edge_table']]`, represents a long edge list that links focal `uid` to their co-authors per work, along with basic publication metadata. The updated list is returned.

```{r}
add_edge_table = function(data){
  data[['edge_table']] = data[['works']] |>
    bind_rows() |>
    select(uid, work_id, authorships, publication_date, publication_year, type) |>
    # mutate(a = authorships, .after=authorships) |>
    rename(a = authorships) |>
    unnest(a, names_sep = "__", keep_empty=FALSE) |>
    filter(!is.na(a__uid))

  return(data)
}
```

## Application

```{r}
data = freadRDS2(file="scholars", location = "combined") |>
  prune_invalid_publication_types() |>
  prune_works_by_author_id() |>
  prune_demographics() |>
  prune_scholars() |>
  clean_demographics() |>
  add_updated_stats_to_demographics() |>
  add_edge_table()

fsaveRDS(data, 'scholars', location = "./data/clean/")
```
