---
title: Prune Scrape Data
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions

execute: 
  eval: false
---

## Getting Started

```{r}
#| warning: false
#| output: false
# load custom functions
source("src/utils/custom_functions.r")

# clear the global environment and set dependencies
.clear_global_environment()
.load_quarto_dependencies()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(slider)
```

## Functions

:::{.callout-note}
- h-index (jaarlijks)
- DONE vluchtige reputatie
  - citations
- DONE carriere leeftijd
  - years since first publication
- topic (node of edge)
- university
- gender
- herkomst
- ?? all metrics??
  - plaatsen (geographic distance)
  - afkomst,

- beurzen
- DONE productiviteit
  - yearly
  - window average < 3years
- topic similarity
:::

met descriptives
(varieren over tijd, gender, herkomst, positie)

## Application

```{r}
data = freadRDS2('scholars', location = "./data/clean/")
```


```{r}
select_works = function(data){
  # define list of valid data types
  valid_publication_types <- c(
    'article', 'book', 'book-chapter', 'dissertation', 
    'editorial', 'report', 'review',
    # optional, to consider:
    'erratum', 'letter', 'other', 'preprint'
  )

  selected_works <- data$works |>
    bind_rows() |> 
    filter(
      !is_retracted,
      type %in% valid_publication_types,
      !is.na(publication_year)
    ) |>
    select(
      uid, work_id, publication_year, publication_date, 
      type, counts_by_year) |>
    distinct(uid, work_id, .keep_all = TRUE)

  return(selected_works)
}
 
```


```{r}
publications_by_year = function(selected_works){
  pubs_by_year <- selected_works |>
    rename(year = publication_year) |>
    # complete date ranges per author
    group_by(uid, year) |>
    summarise(publications = n())|>
    arrange(uid, year) |>
    # count the number of publications per author year
    group_by(uid) |>
    complete(
      year = full_seq(range(year, na.rm = TRUE), 1),
      fill = list(publications = 0)
    ) |>
    mutate(
      year = as.integer(year),
      carreer_age = year - min(year),
      cum_publications = cumsum(publications),
      `2yr_publications` = slide_dbl(publications, sum, .before = 1, .complete=TRUE)
    ) |>
    ungroup() 

  return(pubs_by_year)
}

```


```{r}

citations_by_year = function(selected_works) {
  cited_by_year = selected_works |>
    unnest(counts_by_year) |>
    # complete date ranges per author
    arrange(uid, year) |>
    group_by(uid, year) |>
    summarise(citations = sum(cited_by_count)) |>
    drop_na(citations) |>
    # count the number of citations per author year
    group_by(uid) |>
    complete(
      year = full_seq(range(year, na.rm = TRUE), 1),
      fill = list(citations = 0)
    ) |>
    mutate(
      year = as.integer(year),
      cum_citations = cumsum(citations),
      `2yr_citations` = slide_dbl(citations, sum, .before = 1, .complete=TRUE)
    )

  return(cited_by_year)
}

```

```{r}

add_yearly_work_counts = function(data){
  pubs = data |> select_works() |> publications_by_year()
  cites = data |> select_works() |> citations_by_year()

  #combine publications and citations, and count relative statistics
  data$work_counts = pubs |>
    left_join(cites) |>
    # slicing data based on OpenAlex's citation data availability
    filter(year >= 2015) |>
    mutate(
      cited_by_cumpub = cum_citations / cum_publications,
      cited_by_pub = `citations` / `publications`
    )

  return(data)
}

```


```{r}
data = data |> add_yearly_work_counts()
```