---
title: Prune Scrape Data
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions

execute: 
  eval: false
---

## Getting Started

```{r}
#| warning: false
#| output: false
# load custom functions
source("src/utils/custom_functions.r")

# clear the global environment and set dependencies
.clear_global_environment()
.load_quarto_dependencies()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(slider)
```

## Functions

:::{.callout-note}

- DONE vluchtige reputatie
  - citations
- DONE carriere leeftijd
  - years since first publication
- topic (node of edge)
- DONE university
- DONE position
- DONE gender
- DONE herkomst 
- ?? all metrics??
  - (TODO) plaatsen (geographic distance)
  - DONE h-index (yearly)
  - DONE 2yr_mean_citedness (impact)
  - DONE i10-index (yearly)
- beurzen [external data required]
- DONE productiviteit
  - yearly
  - window average < 3years
- topic similarity
- interdisciplinariteit
:::

met descriptives
(varieren over tijd, gender, herkomst, positie)

## Application

```{r}
data = freadRDS2('scholars', location = "./data/clean/")
```

```{r}

categorize_ethnicity = function(data){
  minority_codes =  c('MA', 'TR', 'BQ', 'CW', 'SX', 'AW', 'SR')

  data[['demographics']] = data[['demographics']] |>
    mutate(migrant = case_when(
      str_detect(coalesce(ethnicity, ""), "NL") ~ "majority",
      ethnicity %in% minority_codes ~ "minority",
      is.na(ethnicity) ~ NA_character_,
      .default = "other"
    )) |>
    relocate(migrant, .after = 'ethnicity')

  return(data)
}

data = categorize_ethnicity(data)
```

```{r}
top_n <- 25

data[["demographics"]] |>
  count(ethnicity, migrant, name = "n") |>
  arrange(n) |>
  tail(25) |>
  mutate(ethnicity = factor(ethnicity, levels = unique(ethnicity))) |>
  ggplot(aes(x = ethnicity, y = n, fill = migrant)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Count", fill = "Migrant", title = "Counts by ethnicity, split by migrant") +
  theme_minimal()
```

```{r}
data[['demographics']] |>
  filter(!is.na(career_start)) |>
  ggplot(aes(x = career_start)) +
  geom_density() +
  labs(x = "Career start (year)", y = "Density", title = "Kernel density of career start") +
  theme_minimal()
```


```{r}
data[['demographics']] |>
  filter(!is.na(h_index_22)) |>
  ggplot(aes(x = h_index_22)) +
  geom_density() +
  labs(x = "h_index", y = "Density", title = "Kernel density of h-index in 2022") +
  theme_minimal()
```

```{r}
data[['demographics']] |>
  filter(!is.na(i10_index_22)) |>
  ggplot(aes(x = i10_index_22)) +
  geom_density() +
  labs(x = "i10_index", y = "Density", title = "Kernel density of i10-index in 2022") +
  theme_minimal()
```

```{r}
data[["demographics"]] |>
  group_by(position_22) |>
  summarise(has_valid_works = mean(has_valid_works), .groups = "drop") |>
  arrange(desc(has_valid_works)) |>
  mutate(position_22 = factor(position_22, levels = position_22)) |>
  ggplot(aes(x = position_22, y = has_valid_works)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = NULL,
    y = "Share with valid works",
    title = "Valid works by position (2025)"
  ) +
  theme_minimal()
```



```{r}
topic_data = data$works |>
  bind_rows() |>
  select(uid, work_id, publication_year, topics) |>
  unnest_longer(topics) |>
  unnest_wider(topics) |>
  mutate(
    id = str_remove(id, 'https://openalex.org/') |>
      str_replace('subfields/', 'S') |>
      str_replace('fields/', 'F') |>
      str_replace('domains/', 'D')
  )

topics = list()
```


```{r}
topics$counts$domains = topic_data |>
  filter(type == 'domain') |>
  select(uid, work_id, publication_year, id) |>
  arrange(id) |>
  mutate(value = 1L) |>
  pivot_wider(
    id_cols = c(uid, work_id, publication_year),
    names_from = id,
    values_from = value,
    values_fn = sum,
    values_fill = 0L
  )

topics$counts$domains
```


```{r}
topics$counts$fields = topic_data |>
  filter(type == 'field') |>
  select(uid, work_id, publication_year, id) |>
  arrange(id) |>
  mutate(value = 1L) |>
  pivot_wider(
    id_cols = c(uid, work_id, publication_year),
    names_from = id,
    values_from = value,
    values_fn = sum,
    values_fill = 0L
  )

topics$counts$fields
```


```{r}
topics$counts$subfields = topic_data |>
  filter(type == 'subfield') |>
  select(uid, work_id, publication_year, id) |>
  arrange(id) |>
  mutate(value = 1L) |>
  pivot_wider(
    id_cols = c(uid, work_id, publication_year),
    names_from = id,
    values_from = value,
    values_fn = sum,
    values_fill = 0L
  )

topics$counts$subfields
```

```{r}
topics$counts$topics = topic_data |>
  filter(type == 'topic') |>
  select(uid, work_id, publication_year, id) |>
  arrange(id) |>
  mutate(value = 1L) |>
  pivot_wider(
    id_cols = c(uid, work_id, publication_year),
    names_from = id,
    values_from = value,
    values_fn = sum,
    values_fill = 0L
  )

topics$counts$topics
```


```{r}
counts = topics$counts |>
  reduce(~ left_join(.x, .y))

valid_topics = counts |>
  select(matches("\\d$")) |>
  summarise(across(everything(), ~ sum(as.numeric(.x), na.rm = TRUE))) |>
  unlist(use.names = TRUE) |>
  (\(x) names(x)[x > 20])()
```


```{r}
topics_counts_cum <- counts |>
  select(-work_id) |>
  group_by(uid, publication_year) |>
  summarise(
    across(
      all_of(valid_topics),
      ~ sum(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  ) |>
  arrange(publication_year, .by_group = FALSE) |>
  group_by(uid) |>
  mutate(
    across(
      where(is.numeric) & !any_of(c("publication_year")),
      cumsum
    )
  ) |>
  ungroup()

```


```{r}
# alternative function for more 

# dissimilarity = counts |>
#   select(matches("\\d$")) |>
#   as.matrix() |>ba
#   coop::cosine()    # uses coop for optimized computation
```