---
title: Prune Scrape Data
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions

execute: 
  eval: false
---

## Getting Started

```{r}
#| warning: false
#| output: false
# load custom functions
source("src/utils/custom_functions.r")

# clear the global environment and set dependencies
.clear_global_environment()
.load_quarto_dependencies()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(slider)
```

## Functions

:::{.callout-note}

- DONE vluchtige reputatie
  - citations
- DONE carriere leeftijd
  - years since first publication
- topic (node of edge)
- DONE university
- DONE position
- DONE gender
- DONE herkomst 
- ?? all metrics??
  - plaatsen (geographic distance)
  - DONE h-index (yearly)
  - DONE 2yr_mean_citedness (impact)
  - DONE i10-index (yearly)
- beurzen [external data required]
- DONE productiviteit
  - yearly
  - window average < 3years
- topic similarity
:::

met descriptives
(varieren over tijd, gender, herkomst, positie)

## Application

```{r}
data = freadRDS2('scholars', location = "./data/clean/")
```


```{r}
select_works = function(data){
  # define list of valid data types
  valid_publication_types <- c(
    'article', 'book', 'book-chapter', 'dissertation', 
    'editorial', 'report', 'review',
    # optional, to consider:
    'erratum', 'letter', 'other', 'preprint'
  )

  selected_works <- data$works |>
    bind_rows() |> 
    filter(
      !is_retracted,
      type %in% valid_publication_types,
      !is.na(publication_year)
    ) |>
    select(
      uid, work_id, publication_year, publication_date, 
      type, counts_by_year) |>
    distinct(uid, work_id, .keep_all = TRUE)

  return(selected_works)
}
 
```


```{r}
publications_by_year = function(selected_works){
  pubs_by_year <- selected_works |>
    rename(year = publication_year) |>
    # complete date ranges per author
    group_by(uid, year) |>
    summarise(publications = n())|>
    arrange(uid, year) |>
    # count the number of publications per author year
    group_by(uid) |>
    complete(
      year = full_seq(range(year, na.rm = TRUE), 1),
      fill = list(publications = 0)
    ) |>
    mutate(
      year = as.integer(year),
      carreer_age = year - min(year),
      cum_publications = cumsum(publications),
      `2yr_publications` = slide_dbl(publications, sum, .before = 1, .complete=TRUE)
    ) |>
    ungroup() 

  return(pubs_by_year)
}

```


```{r}
citations_by_year = function(selected_works) {
  cited_by_year = selected_works |>
    unnest(counts_by_year) |>
    # complete date ranges per author
    arrange(uid, year) |>
    group_by(uid, year) |>
    summarise(citations = sum(cited_by_count)) |>
    drop_na(citations) |>
    # count the number of citations per author year
    group_by(uid) |>
    complete(
      year = full_seq(range(year, na.rm = TRUE), 1),
      fill = list(citations = 0)
    ) |>
    mutate(
      year = as.integer(year),
      cum_citations = cumsum(citations),
      `2yr_citations` = slide_dbl(citations, sum, .before = 1, .complete=TRUE)
    )

  return(cited_by_year)
}

```

```{r}

add_yearly_work_counts = function(data){
  pubs = data |> select_works() |> publications_by_year()
  cites = data |> select_works() |> citations_by_year()

  #combine publications and citations, and count relative statistics
  data$work_counts = pubs |>
    left_join(cites) |>
    # slicing data based on OpenAlex's citation data availability
    filter(year >= 2015) |>
    mutate(
      cited_by_cumpub = cum_citations / cum_publications,
      cited_by_pub = `citations` / `publications`
    )

  return(data)
}

```


```{r}
data = data |> add_yearly_work_counts()
```


```{r}

categorize_ethnicity = function(data){
  minority_codes =  c('MA', 'TR', 'BQ', 'CW', 'SX', 'AW', 'SR')

  data$demographics = data$demographics |>
    mutate(migrant = case_when(
      str_detect(coalesce(ethnicity, ""), "NL") ~ "majority",
      ethnicity %in% minority_codes ~ "minority",
      is.na(ethnicity) ~ NA_character_,
      .default = "other"
    )) |>
    relocate(migrant, .after = 'ethnicity')

  return(data)
}

data = categorize_ethnicity(data)

```

```{r}
h_index_vec = function(citations) {
  if (is.null(citations) || length(citations) == 0L) return(0L)
  c = sort(as.integer(citations[is.finite(citations) & !is.na(citations)]), decreasing = TRUE)
  if (!length(c)) return(0L)
  # With c sorted desc, h is the count of entries with c_i >= i
  as.integer(sum(c >= seq_along(c)))
}

extract_citations = function(works_obj) {
  # helper to deduplicate & pull citations from a data.frame
  df_to_cites = function(df) {
    if (!is.data.frame(df)) return(integer())
    if ("id" %in% names(df)) df = distinct(df, id, .keep_all = TRUE)
    if ("cited_by_count" %in% names(df)) return(as.integer(df$cited_by_count))
    integer()
  }

  if (is.data.frame(works_obj)) {
    return(df_to_cites(works_obj))
  }

  if (is.list(works_obj)) {
    # common wrapper: list(results = data.frame(...))
    if ("results" %in% names(works_obj)) return(df_to_cites(works_obj$results))

    # otherwise assume a list of work records
    cites = map_dbl(works_obj, ~ tryCatch(
      {
        v = .x[["cited_by_count"]]
        if (is.null(v)) NA_real_ else as.numeric(v)
      },
      error = function(e) NA_real_
    ))
    return(as.integer(cites[!is.na(cites)]))
  }

  integer()
}

```

```{r}
data$works |>
  bind_rows() |>
  select(work_id, counts_by_year) |>
  unnest_longer(counts_by_year)  |>
  unnest_wider(counts_by_year, names_sep = "_") |>
  rename(
    year = counts_by_year_year,
    citations = counts_by_year_cited_by_count
  )
```

----

```{r}
add_stats_by_year = function(data){
  cites = extract_citations_by_year(data)

  # calculate i10_indices
  i10_index_by_year = cites |>
    mutate(i10 = i > 10) |>
    group_by(uid, year) |>
    summarise(i10_index = sum(i10)) |>
    mutate(year = as.integer(year))

  # calculate h_indices
  h_index_by_year = cites |> 
    group_by(uid, year) |>
    summarise(h_index = sum(i > k)) |>
    mutate(year = as.integer(year))

  # aggregate citations and publications
  counts = cites |> 
    group_by(uid, year) |>
    summarize(
      citations = sum(i),
      publications = n_distinct(work_id)
    )

  # combine statistics in variable of data.frames
  statistics <- h_index_by_year |>
    full_join(i10_index_by_year, by = c("uid", "year")) |>
    full_join(counts, by = c("uid", "year")) |>
    filter(year > 2015, year < 2026) |>
    arrange(uid, year) |>
    group_by(uid) |>
    nest(stats_by_year = c(year, h_index, i10_index, citations, publications)) |>
    ungroup()

  data$demographics <- data$demographics |>
    left_join(statistics)

  return(data)
}

```

```{r}
data = add_stats_by_year(data)
```
