---
title: Prune Scrape Data
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions

execute: 
  eval: false
---

## Getting Started

```{r}
#| warning: false
#| output: false
# load custom functions
source("src/utils/custom_functions.r")

# clear the global environment and set dependencies
.clear_global_environment()
.load_quarto_dependencies()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(slider)
```

## Functions

:::{.callout-note}

- DONE vluchtige reputatie
  - citations
- DONE carriere leeftijd
  - years since first publication
- DONE topic (node of edge)
- DONE university
  - (optional) time varying (dummy tweede affiliatie (0/1))
  - hoofdaffiliatie
  - dyadisch, zit een van je affiliaties in de affiliaties van de ander
- DONE position
- DONE gender
- DONE herkomst 
- ?? all metrics??
  - (TODO) plaatsen (geographic distance)
  - DONE 2yr_mean_citedness (impact)
  - DONE i10-index (yearly)
  NOTE (consistente rolling window)
- DONE productiviteit
  - yearly
  - window average < 2years
- DONE topic similarity
- DONE aantal auteurs (binnen en buiten corpus) tijdsconstant 2021
:::

AP : schoon dataframe

NOTE:
- met hoeveel auteurs heb je samen gewerkt (binnen en buiten corpus)

met descriptives
(varieren over tijd, gender, herkomst, positie)

## Application

```{r}
data = freadRDS2('scholars', location = "./data/clean/")
```

```{r}

categorize_ethnicity = function(data){
  minority_codes =  c('MA', 'TR', 'BQ', 'CW', 'SX', 'AW', 'SR')
  country_codes = c('US', 'DE', 'BE')

  data[['demographics']] = data[['demographics']] |>
    mutate(migrant = case_when(
      country_id == 'NL' ~ 'NL',
      # country_id %in% country_codes ~ 'B3', # Big 3 origins
      country_id %in% minority_codes ~ 'M4', # 4 Biggest Migrant groups NL
      is.na(country_id) ~ NA_character_,
      .default = "other"
    )) |>
    relocate(migrant, .after = 'country_id')

  return(data)
}

data = categorize_ethnicity(data)
```

```{r}
top_n <- 25

data[["demographics"]] |>
  count(country_id, migrant, name = "n") |>
  arrange(n) |>
  tail(25) |>
  mutate(country_id = factor(country_id, levels = unique(country_id))) |>
  ggplot(aes(x = country_id, y = n, fill = migrant)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Count", fill = "Migrant", title = "Counts by country_id, split by migrant") +
  theme_minimal()
```

```{r}
data[['demographics']] |>
  filter(!is.na(career_start)) |>
  ggplot(aes(x = career_start)) +
  geom_density() +
  labs(x = "Career start (year)", y = "Density", title = "Kernel density of career start") +
  theme_minimal()
```


```{r}
data[['demographics']] |>
  filter(!is.na(h_index_22)) |>
  ggplot(aes(x = h_index_22)) +
  geom_density() +
  labs(x = "h_index", y = "Density", title = "Kernel density of h-index in 2022") +
  theme_minimal()
```

```{r}
data[['demographics']] |>
  filter(!is.na(i10_index_22)) |>
  ggplot(aes(x = i10_index_22)) +
  geom_density() +
  labs(x = "i10_index", y = "Density", title = "Kernel density of i10-index in 2022") +
  theme_minimal()
```

```{r}
data[["demographics"]] |>
  group_by(position_22) |>
  summarise(has_valid_works = mean(!is.na(career_start)), .groups = "drop") |>
  arrange(desc(has_valid_works)) |>
  mutate(position_22 = factor(position_22, levels = position_22)) |>
  ggplot(aes(x = position_22, y = has_valid_works)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = NULL,
    y = "Share with valid works",
    title = "Valid works by position (2025)"
  ) +
  theme_minimal()
```


```{r}
create_topic_table = function(data, base_year = 2022, window = 5){
  data[['works']] |>
    keep(~ nrow(.x) > 0) |>
    bind_rows() |>
    filter(
      publication_year >= (base_year - window),
      publication_year < base_year
    ) |>
    select(uid, work_id, publication_year, topics) |>
    unnest_longer(topics) |>
    unnest_wider(topics) |>
    mutate(
      id = str_remove(id, 'https://openalex.org/') |>
        str_replace('subfields/', 'S') |>
        str_replace('fields/', 'F') |>
        str_replace('domains/', 'D')
    ) 
}

topics = list()
```

```{r}
topic_table = data |> create_topic_table(base_year = 2025, window = 8)
```

```{r}
pivot_topic_table = function(topic_table, topic_type = NA){
  # assert topic_type is valid
  valid_types <- topic_table |> distinct(type) |> pull()
  if ((is.na(topic_type) && !topic_type %in% valid_types)) {
    warning(paste(topic_type, "is not valid"))
  }

  # select topic_type and pivot topics to wider
  topic_table |>
    filter(type == topic_type) |>
    select(uid, work_id, publication_year, id) |>
    arrange(id) |>
    mutate(value = 1L) |>
    pivot_wider(
      id_cols = c(uid, work_id, publication_year),
      names_from = id,
      values_from = value,
      values_fn = sum,
      values_fill = 0L
    )
}

aggregate_topic_table = function(table){
  table |>
    group_by(uid) |>
    summarise(across(matches("\\d$"), sum))
}

calculate_proximity_matrix = function(counts){
  d <- 1 - counts |>
    select(matches("\\d$")) |>
    coop::cosine()
  diag(d) = 0
  
  prox <- 1 - (d/max(d))^2
  
  return(prox)
}

calculate_interdisciplinarity = function(P, proximity){
  # align proximity once (same for every row)
  prox <- proximity[colnames(P), colnames(P)]
  P[is.na(P)] <- 0
  prox[is.na(prox)] <- 0

  res <- 1 - rowSums((P %*% prox) * P)  # numeric vector, length = nrow(P)
  names(res) <- rownames(P)

  return(res)
}
```

```{r}
topic_types = c('field', 'subfield')
for (topic_type in topic_types) {
  topics[['counts']][[topic_type]] = topic_table |>
    pivot_topic_table(topic_type = topic_type) |>
    aggregate_topic_table()

  # calculate proximity
  topics[['prox']][[topic_type]] = topics[['counts']][[topic_type]] |>
    calculate_proximity_matrix()
}
```

```{r}
years = c(2022, 2024, 2025)

for (topic_type in topic_types) {
  for (year in years){
    year_lab = as.character(year)
    
    # create a proportions table
    P = data |>
      create_topic_table(base_year = year, window = 2) |>
      pivot_topic_table(topic_type = topic_type) |>
      aggregate_topic_table() |>
      column_to_rownames(var="uid") |> 
      as.matrix() |>
      prop.table(margin = 1)

    # fetch proximity table
    prox = topics[['prox']][[topic_type]]

    # calculate interdisciplinarity
    topics[['intdis']][[topic_type]][[year_lab]] = P |>
      calculate_interdisciplinarity(prox) 
  }

  # combine results into matrix
  res = topics[['intdis']][[topic_type]] |> bind_rows() |> t()
  colnames(res) = as.character(paste('int', years, sep = '_'))
  topics[['intdis']][[topic_type]] = res
}

```

```{r}
create_empty_author_matrix = function(topics, topic_type){
  # get all author information
  authors = topics[['counts']][['field']] |> pull(uid)
  k_authors =  authors |> n_distinct()

  D = matrix(
    NA, nrow = k_authors, ncol = k_authors,
    dimnames = list(authors, authors)
  )
  return (D)
}

calculate_dissimilarity = function (topics, year, topic_type){
  # create empty dissimilarity table for all possible authors
  D = create_empty_author_matrix(topics)

  # get author data for specified year
  A = data |>
    create_topic_table(base_year = year, window = 2) |>
    pivot_topic_table(topic_type = topic_type) |>
    aggregate_topic_table() |>
    column_to_rownames(var="uid") |> 
    as.matrix() |>
    prop.table(margin = 1)

  # align A to prox columns
  prox = topics[['prox']][[topic_type]]
  M = 1 - prox[colnames(A), colnames(A)]
  M[is.na(M)] <- 0

  # calculate dissimilarity per author (i) coauthor (j) pair
  for (i in rownames(A)) {
    for (j in rownames(A)) {
      diff = abs(A[i,] - A[j,]) / 2
      D[i, j] = sum(diff %o% diff * M) * 2
    }
  }

  # normalize results
  D <- D / max(D, na.rm = TRUE)
  return(D)
}

# calculate dissimilarties per year per topic type
for (topic_type in topic_types){
  for (year in years){
    year_lab = as.character(year)
    topics[['dissim']][[topic_type]][[year_lab]] = topics |>
      calculate_dissimilarity(year, topic_type)
  }
}

```

```{r}
fsaveRDS(topics, 'topics', './data/clean/')
```

```{r}
topics = freadRDS2('topics', location = 'clean')
```


```{r}

clean_affilliations = function(data){
  valid_universities = c(
    'uu', 'uva', 'rug', 'ru', 'ucu', 'leiden', 'eur',
    'uvt', 'uvg', 'vu', 'wur'
  )

  data[['demographics']] = data[['demographics']] |>
    mutate(
      # get the first valid affiliation per author year
      across(
        starts_with('university'),
        # clean and split university strings
        ~ .x |> str_replace('\\.', '/') |>
          str_to_lower() |>
          str_split(pattern = '/') |>
          # select the first of valid universities
          sapply(\(chunks){
            chunks[chunks %in% valid_universities][1]
          }) |>
          str_to_upper() |>
          str_replace('LEIDEN', 'UL')  |>
          str_replace('RUG', 'UVG') |>
          str_replace('UCU', 'UU'),
        .names = "{.col}_first"
      ),
      # calculate the number of affiliations
      across(
        starts_with('university') & !ends_with('first'),
        ~ .x |> str_replace('\\.', '/') |>
          str_to_lower() |>
          str_split(pattern = '/') |>
          sapply(\(chunks){
            length(chunks[!is.na(chunks)])
          }),
        .names = "{.col}_k"
      )
    )    
  
  return(data)
}

data = clean_affilliations(data)

```

```{r}
set_time_matrix = function(){
  time_distance = c(
    UVG = c(0,    NA,   NA,    NA,    NA,    NA,    NA,    NA),
    VU  = c(2.52, 0,    NA,    NA,    NA,    NA,    NA,    NA),
    UVA = c(2.58, 0.52, 0,     NA,    NA,    NA,    NA,    NA),
    UL  = c(3.12, 0.91, 1.04,  0,     NA,    NA,    NA,    NA), 
    EUR = c(3.17, 1.32, 1.38,  1.18,  0,     NA,    NA,    NA),
    EVT = c(3.30, 1.70, 1.57,  1.67,  1.35,  0,     NA,    NA),
    UU  = c(2.47, 0.98, 0.88,  1.40,  1.33,  1.36,  0,     NA),
    RU  = c(2.83, 1.90, 1.68,  2.23,  2.33,  1.30,  1.53,  0)
  )
  labs = c('UVG', 'VU', 'UVA', 'UL', 'EUR', 'UVT', 'UU', 'RU')

  # convert to symmetrical matrix
  T = matrix(time_distance,
    ncol = 8, nrow = 8, dimnames = list(labs, labs)
  )
  T[lower.tri(T)] = T[upper.tri(T)]
  return(T)
}


```

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(purrr)

#' Create long affiliation table (uid-year-first affiliation)
create_affiliations_long <- function(demographics,
                                     affix = "first",
                                     year_regex = "\\d{2}") {
  demographics %>%
    select(uid, ends_with(affix)) %>%
    pivot_longer(
      cols = ends_with(affix),
      names_to = "name",
      values_to = "value"
    ) %>%
    mutate(year = paste0("20", str_extract(name, regex(year_regex)))) %>%
    select(uid, year, value)
}

#' Convert travel-time matrix T to long lookup table
create_time_lookup <- function(T) {
  as.data.frame(as.table(T)) %>%
    as_tibble() %>%
    rename(value_i = Var1, value_j = Var2, distance = Freq)
}

#' Create dyad table per year with affiliation values and travel-time distance
create_pairs_with_distance <- function(affiliations, T_long, drop_self = FALSE) {
  aff_key <- affiliations %>% distinct(year, uid, value)

  pairs <- aff_key %>%
    group_by(year) %>%
    summarise(pairs = list(crossing(uid_i = uid, uid_j = uid)), .groups = "drop") %>%
    unnest(pairs)

  if (drop_self) {
    pairs <- pairs %>% filter(uid_i != uid_j)
  }

  pairs %>%
    left_join(aff_key, by = c("year", "uid_i" = "uid")) %>%
    rename(value_i = value) %>%
    left_join(aff_key, by = c("year", "uid_j" = "uid")) %>%
    rename(value_j = value) %>%
    drop_na(value_i, value_j) %>%
    left_join(T_long, by = c("value_i", "value_j")) %>%
    drop_na(distance)
}

#' Build a full (all_uids x all_uids) matrix for a single year from dyad table
make_year_matrix <- function(df_year, all_uids) {
  M <- matrix(
    NA_real_,
    nrow = length(all_uids),
    ncol = length(all_uids),
    dimnames = list(all_uids, all_uids)
  )

  ii <- match(df_year$uid_i, all_uids)
  jj <- match(df_year$uid_j, all_uids)
  keep <- !is.na(ii) & !is.na(jj)

  M[cbind(ii[keep], jj[keep])] <- df_year$distance[keep]
  M
}

#' Create list of distance matrices per year (same dimnames for all years)
create_distance_matrices <- function(pairs_by_year, all_uids) {
  pairs_by_year %>%
    select(year, uid_i, uid_j, distance) %>%
    group_by(year) %>%
    group_split() %>%
    set_names(map_chr(., ~ unique(.x$year))) %>%
    map(~ make_year_matrix(.x, all_uids))
}

# affiliations
affiliations <- create_affiliations_long(data[["demographics"]])

# time lookup
T_long <- create_time_lookup(T)

# dyads + distance
pairs_by_year <- create_pairs_with_distance(affiliations, T_long)

# global uid set / ordering
all_uids <- colnames(topics[["dissim"]][["field"]][["2022"]])

# matrices per year
distances <- create_distance_matrices(pairs_by_year, all_uids)
```


```{r}
years <- c("2022", "2024", "2025")

dens_df <- map_dfr(years, function(y) {
  m <- topics[["dissim"]][["subfield"]][[y]]
  tibble(
    year = y,
    dissim = m[upper.tri(m)]
  )
}) |>
  filter(!is.na(dissim))

ggplot(dens_df, aes(x = dissim)) +
  geom_density() +
  facet_wrap(~ year, nrow = 1, scales = "free_y") +
  theme_minimal() +
  labs(x = "Dissimilarity (upper triangle)", y = "Density")
```

```{r}
years <- c("2022", "2024", "2025")

dens_df <- map_dfr(years, function(y) {
  m <- topics[["dissim"]][["field"]][[y]]
  tibble(
    year = y,
    dissim = m[upper.tri(m)]
  )
}) |>
  filter(!is.na(dissim))

ggplot(dens_df, aes(x = dissim)) +
  geom_density() +
  facet_wrap(~ year, nrow = 1, scales = "free_y") +
  theme_minimal() +
  labs(x = "Dissimilarity (upper triangle)", y = "Density")
```

```{r}
years <- c("2022", "2024", "2025")

dens_df <- map_dfr(years, function(y) {
  m <- topics[["dissim"]][["domain"]][[y]]
  tibble(
    year = y,
    dissim = m[upper.tri(m)]
  )
}) |>
  filter(!is.na(dissim))

ggplot(dens_df, aes(x = dissim)) +
  geom_density() +
  facet_wrap(~ year, nrow = 1, scales = "free_y") +
  theme_minimal() +
  labs(x = "Dissimilarity (upper triangle)", y = "Density")
```

```{r}
plot_df <- data[["demographics"]] |>
  select(uid, gender, university_22_first, university_24_first, university_25_first) |>
  pivot_longer(
    cols = starts_with("university_"),
    names_to = "year",
    values_to = "university"
  ) |>
  mutate(
    year = case_when(
      year == "university_22_first" ~ "2022",
      year == "university_24_first" ~ "2024",
      year == "university_25_first" ~ "2025",
      TRUE ~ year
    )
  ) |>
  filter(!is.na(university), university != "", !is.na(gender), gender != "") |>
  count(year, university, gender, name = "n") |>
  group_by(year, university) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

ggplot(plot_df, aes(x = university, y = prop, fill = gender)) +
  geom_col() +
  facet_wrap(~ year, scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Gender composition", fill = "Gender")
```

```{r}
citations = data[['demographics']] |>
  select(uid, stats_by_year) |>
  unnest(stats_by_year)
```


```{r}
plot_df <- data[["demographics"]] |>
  select(university_22_first, university_24_first, university_25_first) |>
  pivot_longer(
    cols = everything(),
    names_to = "panel",
    values_to = "university"
  ) |>
  filter(!is.na(university), university != "") |>
  count(panel, university, name = "n")

ggplot(plot_df, aes(x = university, y = n)) +
  geom_col() +
  facet_wrap(~ panel, scales = "free_x") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(1, "lines")
  ) +
  labs(x = NULL, y = "Count")
```

 check h-index en i10 index met oa data.
 - quick cum-cit / cum-pub 
 
