---
title: "Create patch for institutions"
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions
---

## Getting Started

```{r}
#| warning: false
#| output: false
# load custom functions
source("src/utils/custom_functions.r")

# clear the global environment and set dependencies
.clear_global_environment()
.load_quarto_dependencies()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(stringr)
library(lubridate)
```

## Functions

### `as_tibble_safe()`
Safely coerce nested list elements to tibbles.

#### Arguments
**`x`**  

> Object to be converted; may be `NULL`, a data frame, or another list-like structure.

#### Method
This helper inspects `x` and returns an empty tibble when it is `NULL`, `as_tibble(x)` when it is already a data frame, or `tibble::as_tibble(x)` for other list-like inputs. It is used to robustly normalize nested list columns (e.g., affiliations, last-known institutions) into rectangular structures before unnesting, avoiding errors when encountering `NULL` or heterogeneous types.

```{r}
as_tibble_safe <- function(x) {
  if (is.null(x)) {
    tibble()
  } else if (is.data.frame(x)) {
    as_tibble(x)
  } else {
    tibble::as_tibble(x)
  }
}
```

---

### `unnest_authors()`
Expand works data to one row per author, affiliation, and last-known institution.

#### Arguments
**`works`**  

> Tibble of works with nested `authorships`, `authorships__affiliations`, and `last_known_institutions` list-columns.

#### Method
The function sequentially normalizes and unnests nested columns. It first ensures that each element of `authorships` is a tibble using `as_tibble_safe()`, then unnests to a row-per-author layout. Next it applies the same pattern to `authorships__affiliations`, yielding one row per author–affiliation combination, and finally to `last_known_institutions`, yielding one row per last-known institution record. All unnests use `keep_empty = TRUE` to retain works without authors or affiliations as explicit placeholder rows. The result is a long-format table that exposes authors, current affiliations, and last-known institutions for each work.

```{r}
unnest_authors <- function(works) {
  works |>
    # 1) Make sure each element of `authorships` is a tibble (handles NULLs/lists)
    mutate(
      authorships = map(authorships, as_tibble_safe)
    ) |>
    # 2) One row per author record (keeps works with 0 authors via keep_empty)
    unnest(authorships, keep_empty = TRUE, names_sep = "__") |>
    # 3) Make sure each element of `authorships__affiliations` is a tibble
    mutate(
      authorships__affiliations = map(authorships__affiliations, as_tibble_safe)
    ) |>
    # 4) One row per affiliation record
    unnest(authorships__affiliations, keep_empty = TRUE, names_sep = "_") |>
    # 5) Make sure each element of `last_known_institutions` is a tibble
    mutate(
      last_known_institutions = map(last_known_institutions, as_tibble_safe)
    ) |>
    # 6) One row per last-known institution record
    unnest(last_known_institutions, keep_empty = TRUE, names_sep = "_")
}
```

---

### `prepare_works_authors()`
Construct a long works–authors–affiliations table for all scholars.

#### Arguments
**`data`**  

> List-like object containing at least `scholars_oa` (OpenAlex author records) and `works` (OpenAlex works) elements.

#### Method
The function first binds all OpenAlex author records in `data[["scholars_oa"]]` into a single tibble and merges these with the bound works table from `data[["works"]]` using a many-to-many join on `uid`. It then calls `unnest_authors()` to expand authorships, affiliations, and last-known institutions. Afterwards, it keeps only rows where the scholar’s `uid` matches `authorships__uid` (ensuring the row corresponds to the focal scholar) and filters to rows with either a non-missing `authorships__affiliations_id` or a match between `authorships__affiliations_id` and `last_known_institutions_id`. The result is a cleaned, long-format table that reliably links works, authors, and their institutional affiliations over time.

```{r}
prepare_works_authors <- function(data) {
  authors <- data[["scholars_oa"]] |>
    bind_rows()

  works <- data[["works"]] |>
    bind_rows() |>
    select(-display_name)

  works |>
    left_join(
      authors,
      by = join_by(uid),
      relationship = "many-to-many"
    ) |>
    unnest_authors() |>
    filter(uid == authorships__uid) |>
    filter(
      !is.na(authorships__affiliations_id) |
        authorships__affiliations_id == last_known_institutions_id
    )
}
```

---

### `derive_affiliation_date_ranges()`
Derive first and last publication dates per university affiliation.

#### Arguments
**`works_authors`**  

> Long-format tibble as returned by `prepare_works_authors()` with publication and affiliation information.

**`universities`**  

> Character vector of OpenAlex institution IDs to restrict the analysis to (e.g., Dutch universities).

#### Method
The function starts from the long `works_authors` table and collapses it to distinct combinations of display name, author ID, affiliation ID and name, and publication date. It then filters to rows where `authorships__affiliations_id` is in the supplied `universities`. After sorting by `publication_date`, it groups by `uid`, author, and affiliation, and summarizes the first and last publication dates where that affiliation appears. These dates are then expanded into broader affiliation intervals by subtracting 18 months from `first_date` and adding 6 months to `last_date`, capturing plausible entry and exit windows around the observed publication period. The resulting table contains one row per scholar–university pairing with inferred start and end dates of affiliation.

```{r}
derive_affiliation_date_ranges <- function(works_authors, universities) {
  works_authors |>
    distinct(
      authorships__display_name,
      authorships__author_id,
      authorships__affiliations_id,
      authorships__affiliations_display_name,
      publication_date,
      .keep_all = TRUE
    ) |>
    filter(authorships__affiliations_id %in% universities) |>
    arrange(publication_date) |>
    group_by(
      uid,
      authorships__author_id,
      authorships__affiliations_display_name,
      authorships__affiliations_id
    ) |>
    summarize(
      uid = first(uid),
      authorships__display_name = first(authorships__display_name),
      first_date = first(publication_date),
      last_date = last(publication_date),
      .groups = "drop"
    ) |>
    mutate(
      first_date = first_date %m-% months(18),
      last_date = last_date %m+% months(6)
    )
}
```

---

### `format_institutions()`
Convert affiliation histories into a wide panel of timepoint-specific affiliations.

#### Arguments
**`d`**  

> Tibble containing at least `uid`, `year`, and `authorships__affiliations_display_name`, typically produced by `build_affiliations_table()` before the final reshape.

#### Method
The function first reduces `d` to distinct combinations of `uid`, timepoint `year`, and affiliation name, keeping only years after 2015. Within each `(uid, year)` group, it concatenates unique affiliation names into a single `affiliations` string, separated by `"; "`. It then pivots the data wider with one row per `uid` and columns `affiliations_<year>`, restricting to a predefined set of evaluation years (2019, 2022, 2024, 2025). As a final convenience step, it back-fills `affiliations_25` with `affiliations_24` when the 2025 value is missing, ensuring a complete last column where possible. The result is a compact, wide-format table with one row per scholar and affiliation strings at key timepoints.

```{r}
format_institutions <- function(d) {
  d |>
    distinct(uid, year, authorships__affiliations_display_name) |>
    filter(year > 15) |>
    arrange(year) |>
    group_by(uid, year) |>
    reframe(
      affiliations = paste(
        unique(authorships__affiliations_display_name),
        collapse = "; "
      )
    ) |>
    pivot_wider(
      id_cols = uid,
      names_from = year,
      values_from = affiliations,
      names_prefix = "affiliations_"
    ) |>
    select(
      uid,
      num_range("affiliations_", c(19, 22, 24, 25))
    ) |>
    mutate(
      affiliations_25 = ifelse(
        is.na(affiliations_25),
        affiliations_24,
        affiliations_25
      )
    )
}
```

---

### `build_affiliations_table()`
Construct a timepoint-based affiliation panel from inferred date ranges.

#### Arguments
**`dates`**  

> Tibble with inferred `first_date` and `last_date` per `(uid, affiliation)` as returned by `derive_affiliation_date_ranges()`.

**`timepoints`**  

> Character vector of evaluation dates (e.g., `"2019-09-01"`, `"2022-12-19"`) to use as panel timepoints.

#### Method
The function first converts `timepoints` to dates and forms their cartesian product with `dates` using `expand_grid()`. For each scholar–affiliation–timepoint combination, it flags whether the timepoint falls within the inferred interval `[first_date, last_date]` and computes a short two-digit `year` label. It then filters to in-range observations, deduplicates by scholar, author, affiliation name, and year, and drops rows with missing affiliation names. The resulting long-format table is passed to `format_institutions()`, which reshapes it into a wide panel of affiliations at key timepoints. The final output is an affiliation patch table suitable for merging into the main scholars dataset.

```{r}
build_affiliations_table <- function(dates, timepoints) {
  tp <- ymd(timepoints)

  dates |>
    tidyr::expand_grid(timepoint = tp) |>
    arrange(timepoint) |>
    mutate(
      in_range = between(timepoint, first_date, last_date),
      year = as.integer(year(timepoint) - 2000)
    ) |>
    filter(in_range) |>
    distinct(
      uid,
      authorships__display_name,
      authorships__author_id,
      authorships__affiliations_display_name,
      year
    ) |>
    filter(!is.na(authorships__affiliations_display_name)) |>
    format_institutions()
}
```

## Application

```{r}
# Load scholars data
data <- freadRDS2("scholars", location = "clean")

# Build long works–authors–affiliations table
works_authors <- prepare_works_authors(data)

# Define OpenAlex institution IDs for the universities of interest
universities <- c(
  "https://openalex.org/I865915315", "https://openalex.org/I887064364",
  "https://openalex.org/I121797337", "https://openalex.org/I145872427",
  "https://openalex.org/I193662353", "https://openalex.org/I913958620",
  "https://openalex.org/I193700539", "https://openalex.org/I169381384",
  "https://openalex.org/I913481162"
)

# Derive first and last dates of university affiliations
dates <- derive_affiliation_date_ranges(
  works_authors = works_authors,
  universities = universities
)

# Define evaluation timepoints
timepoints <- c(
  "2019-09-01",
  "2022-12-19",
  "2024-04-19",
  "2025-10-02"
)

# Build final affiliations table
affiliations <- build_affiliations_table(
  dates = dates,
  timepoints = timepoints
)

# Save to disk
fsaveRDS(affiliations, "affiliations")
```

```{r}
affiliations |> head()
```
