---
title: Process Topic Information
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions

execute:
  eval: false
---

## Getting Started

```{r}
#| warning: false
#| output: false
# load custom functions
source("src/utils/custom_functions.r")

# clear the global environment and set dependencies
.clear_global_environment()
.load_quarto_dependencies()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(slider)
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(purrr)
```

## Functions

### `create_topic_table()`

Build a long-form topic table from works data within a specified look-back window.

#### Arguments

**`data`**

> Named list that contains a `works` element (list of works tibbles with at least `uid`, `work_id`, `publication_year`, and `topics`).

**`base_year`**

> Integer. The reference year; only works published in the `window` years preceding this year are included. Defaults to `2022`.

**`window`**

> Integer. Number of years to look back from `base_year`. Defaults to `5`.

#### Method

The function binds all non-empty works tibbles together and filters to works whose `publication_year` falls in `[base_year - window, base_year)`. It selects `uid`, `work_id`, `publication_year`, and `topics`, then unnests and widens the `topics` column. Finally it strips the OpenAlex base URL and maps `subfields/`, `fields/`, and `domains/` prefixes to compact short codes (`S`, `F`, `D`).

```{r}
create_topic_table = function(data, base_year = 2022, window = 5){
  data[['works']] |>
    keep(~ nrow(.x) > 0) |>
    bind_rows() |>
    filter(
      publication_year >= (base_year - window),
      publication_year < base_year
    ) |>
    select(uid, work_id, publication_year, topics) |>
    unnest_longer(topics) |>
    unnest_wider(topics) |>
    mutate(
      id = str_remove(id, 'https://openalex.org/') |>
        str_replace('subfields/', 'S') |>
        str_replace('fields/', 'F') |>
        str_replace('domains/', 'D')
    )
}
```

### `pivot_topic_table()`

Filter a long topic table to a given topic type and pivot it to a wide binary indicator matrix.

#### Arguments

**`topic_table`**

> A long-form topic tibble as returned by `create_topic_table()`, with at least `uid`, `work_id`, `publication_year`, `id`, and `type` columns.

**`topic_type`**

> Character string. The topic hierarchy level to select (e.g. `"field"`, `"subfield"`). Must be a value present in the `type` column. Defaults to `NA`.

#### Method

The function validates that `topic_type` is present in the `type` column, emitting a warning if not. It filters to the specified type, selects `uid`, `work_id`, `publication_year`, and `id`, adds a binary indicator column, and pivots wider so each unique topic `id` becomes a column. Duplicate topic–work pairs are summed; missing values are filled with `0L`.

```{r}
pivot_topic_table = function(topic_table, topic_type = NA){
  # assert topic_type is valid
  valid_types <- topic_table |> distinct(type) |> pull()
  if ((is.na(topic_type) && !topic_type %in% valid_types)) {
    warning(paste(topic_type, "is not valid"))
  }

  # select topic_type and pivot topics to wider
  topic_table |>
    filter(type == topic_type) |>
    select(uid, work_id, publication_year, id) |>
    arrange(id) |>
    mutate(value = 1L) |>
    pivot_wider(
      id_cols = c(uid, work_id, publication_year),
      names_from = id,
      values_from = value,
      values_fn = sum,
      values_fill = 0L
    )
}
```

### `aggregate_topic_table()`

Aggregate a wide topic indicator table to one row per scholar by summing across works.

#### Arguments

**`table`**

> A wide topic tibble as returned by `pivot_topic_table()`, with a `uid` column and numeric topic indicator columns whose names end in a digit.

#### Method

The function groups by `uid` and summarises all columns whose names end with a digit using `sum`, collapsing all per-work rows into a single per-scholar topic profile.

```{r}
aggregate_topic_table = function(table){
  table |>
    group_by(uid) |>
    summarise(across(matches("\\d$"), sum))
}
```

### `calculate_proximity_matrix()`

Compute a normalized cosine-proximity matrix between topics from a scholar-by-topic counts table.

#### Arguments

**`counts`**

> A wide tibble with topic indicator or count columns (column names ending in a digit), as returned by `aggregate_topic_table()`.

#### Method

The function extracts the numeric topic columns and computes pairwise cosine distances using `coop::cosine()`. The diagonal is set to `0`. The resulting distance matrix is normalized to a 0–1 proximity scale via `1 - (d / max(d))^2` and returned.

```{r}
calculate_proximity_matrix = function(counts){
  d <- 1 - counts |>
    select(matches("\\d$")) |>
    coop::cosine()
  diag(d) = 0

  prox <- 1 - (d/max(d))^2

  return(prox)
}
```

### `calculate_interdisciplinarity()`

Compute a scalar interdisciplinarity score for each scholar from their topic-proportion vector and the topic proximity matrix.

#### Arguments

**`P`**

> A numeric matrix of topic proportions (rows = scholars, columns = topics).

**`proximity`**

> A square numeric proximity matrix (topics × topics), as returned by `calculate_proximity_matrix()`.

#### Method

The function aligns `proximity` to the columns of `P` and replaces `NA` values with `0` in both matrices. Interdisciplinarity is computed as `1 - rowSums((P %*% prox) * P)`, yielding a named numeric vector of length `nrow(P)` where higher values indicate broader spread across dissimilar topics.

```{r}
calculate_interdisciplinarity = function(P, proximity){
  # align proximity once (same for every row)
  prox <- proximity[colnames(P), colnames(P)]
  P[is.na(P)] <- 0
  prox[is.na(prox)] <- 0

  res <- 1 - rowSums((P %*% prox) * P)  # numeric vector, length = nrow(P)
  names(res) <- rownames(P)

  return(res)
}
```

### `create_empty_author_matrix()`

Initialize an empty square `NA` matrix indexed by the scholars present in the topics counts table.

#### Arguments

**`topics`**

> Named list containing a `counts` element with a `field` tibble that has a `uid` column.

**`topic_type`**

> Placeholder argument for consistency with related functions; currently unused.

#### Method

The function extracts all `uid` values from `topics[['counts']][['field']]`, determines the number of distinct scholars, and returns an `n × n` matrix filled with `NA` with both row and column names set to the scholar `uid` values.

```{r}
create_empty_author_matrix = function(topics, topic_type){
  # get all author information
  authors = topics[['counts']][['field']] |> pull(uid)
  k_authors =  authors |> n_distinct()

  D = matrix(
    NA, nrow = k_authors, ncol = k_authors,
    dimnames = list(authors, authors)
  )
  return (D)
}
```

### `calculate_dissimilarity()`

Calculate a pairwise normalized dissimilarity matrix between scholars for a given year and topic type.

#### Arguments

**`topics`**

> Named list with `counts` and `prox` sub-lists as produced by the processing pipeline.

**`year`**

> Integer. The reference year passed to `create_topic_table()` with a look-back window of 2.

**`topic_type`**

> Character string. Topic hierarchy level to use (e.g. `"field"`, `"subfield"`).

#### Method

The function initializes an empty author matrix via `create_empty_author_matrix()`. It then builds a topic-proportion matrix `A` for the specified year and type by calling `create_topic_table()`, `pivot_topic_table()`, `aggregate_topic_table()`, and `prop.table()`. The stored proximity matrix for the topic type is converted to a distance matrix `M`. For every scholar pair `(i, j)`, dissimilarity is computed as the sum of the outer product of absolute half-differences weighted by `M`, multiplied by `2`. The resulting matrix is normalized by its maximum value and returned.

```{r}
calculate_dissimilarity = function (topics, year, topic_type){
  # create empty dissimilarity table for all possible authors
  D = create_empty_author_matrix(topics)

  # get author data for specified year
  A = data |>
    create_topic_table(base_year = year, window = 2) |>
    pivot_topic_table(topic_type = topic_type) |>
    aggregate_topic_table() |>
    column_to_rownames(var="uid") |>
    as.matrix() |>
    prop.table(margin = 1)

  # align A to prox columns
  prox = topics[['prox']][[topic_type]]
  M = 1 - prox[colnames(A), colnames(A)]
  M[is.na(M)] <- 0

  # calculate dissimilarity per author (i) coauthor (j) pair
  for (i in rownames(A)) {
    for (j in rownames(A)) {
      diff = abs(A[i,] - A[j,]) / 2
      D[i, j] = sum(diff %o% diff * M) * 2
    }
  }

  # normalize results
  D <- D / max(D, na.rm = TRUE)
  return(D)
}
```

## Application

```{r}
topics = list()
data = freadRDS2('scholars', location = "./data/clean/")
topic_table = data |> create_topic_table(base_year = 2025, window = 8)
```

```{r}
topic_types = c('field', 'subfield')
for (topic_type in topic_types) {
  topics[['counts']][[topic_type]] = topic_table |>
    pivot_topic_table(topic_type = topic_type) |>
    aggregate_topic_table()

  # calculate proximity
  topics[['prox']][[topic_type]] = topics[['counts']][[topic_type]] |>
    calculate_proximity_matrix()
}
```

```{r}
years = c(2022, 2024, 2025)

for (topic_type in topic_types) {
  for (year in years){
    year_lab = as.character(year)

    # create a proportions table
    P = data |>
      create_topic_table(base_year = year, window = 2) |>
      pivot_topic_table(topic_type = topic_type) |>
      aggregate_topic_table() |>
      column_to_rownames(var="uid") |>
      as.matrix() |>
      prop.table(margin = 1)

    # fetch proximity table
    prox = topics[['prox']][[topic_type]]

    # calculate interdisciplinarity
    topics[['intdis']][[topic_type]][[year_lab]] = P |>
      calculate_interdisciplinarity(prox)
  }

  # combine results into matrix
  res = topics[['intdis']][[topic_type]] |> bind_rows() |> t()
  colnames(res) = as.character(paste('int', years, sep = '_'))
  topics[['intdis']][[topic_type]] = res
}
```

```{r}
# calculate dissimilarties per year per topic type
for (topic_type in topic_types){
  for (year in years){
    year_lab = as.character(year)
    topics[['dissim']][[topic_type]][[year_lab]] = topics |>
      calculate_dissimilarity(year, topic_type)
  }
}
```

```{r}
fsaveRDS(topics, 'topics', './data/clean/')
```

```{r}
topics = freadRDS2('topics', location = 'clean')
```
