---
title: Descriptive Statistics
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions
        
# eval: false
---

## Getting Started

```{r}
#| warning: false
#| output: false
# clear the global environment
rm(list = ls())
gc()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(tidyplots)

# load custom functions 
source("src/utils/custom_functions.r")
.load_quarto_dependencies()

options(knitr.graphics.fig.width = 25)
tidyplots_options(
  unit = "cm"
)
```

## Application


```{r}
data = freadRDS2(file="scholars", location = "clean")
```


### Number of publications

To check the data quality of OpenAlex entries I inspect the number of
publications per year. This is a crude but common used strategy to assess the
completeness of repositories such as OpenAlex.

```{r}
#| warning: false
#| fig.width: 9.99

works = data[['works']] |>
  bind_rows() |>
  distinct(work_id, .keep_all = TRUE) 
  
works_count = works |>
  group_by(publication_year, type) |>
  summarise(n = n()) |>
  ungroup() |>
  filter(publication_year >= 1960)

works_count_plot = works_count |>
  tidyplot(x = publication_year, y = n, color = type) |>
  add_barstack_absolute() |>
  adjust_size(width = 15, height = 7) |>
  adjust_x_axis_title("Publication Year") |>
  adjust_x_axis(limits = c(1960, 2026)) |>
  adjust_y_axis_title("Number of publications") |>
  adjust_legend_title("Publication Types")

works_count_plot

```


In 2019 there is an unusual spike in the number of publication. This is
predominantly driven by a couple of scholars publishing numerous datasets. Also,
there are a number of publication types that might not constitute a valid type
of publication. Therefore, I select only valid types and check the distribution
of publications again.

```{r}
#| warning: false
#| fig.width: 10.5
valid_publication_types = c(
  'article', 'book', 'book-chapter', 'dissertation', 
  'editorial', 'report', 'review',
  # optional, to consider:
  'erratum', 'letter', 'other', 'preprint'
)

works_count_plot2 = works_count |>
  filter(type %in% valid_publication_types)|>
  tidyplot(x = publication_year, y = n, color = type) |>
  add_barstack_absolute() |>
  adjust_size(width = 15, height = 7) |>
  adjust_x_axis_title("Publication Year") |>
  adjust_x_axis(limits = c(1960, 2026)) |>
  adjust_y_axis_title("Number of publications") |>
  adjust_legend_title("Publication Types")

works_count_plot2
```

### Number of scholars

```{r}
dems = data[['demographics']] 

dems_long = dems |>
  pivot_longer(
    # only the repeated fields:
    cols = matches("^(email|university|position|discipline)_\\d{2}$"),
    # split into the base name (becomes columns) and the 2-digit year
    names_to = c(".value", "year2"),
    names_pattern = "^(.*)_(\\d{2})$",
    values_drop_na = TRUE  # drop rows where all new values are NA
  ) |>
  mutate(year = as.integer(paste0("20", year2))) |>
  select(-year2) |>
  arrange(uid, year) |>
  mutate(university = na_if(university, "")) |>
  separate_rows(university, sep = "\\s*/\\s*")
```


```{r}
dems_long = dems_long |>
  mutate(
    uni = str_to_lower(university),
    university = case_when(
        str_detect(uni, 'rug') ~ "University of Groningen",
        str_detect(uni, 'ru') ~ "Radboud University",
        str_detect(uni, '(uu|ucu)') ~ "Utrecht University",
        str_detect(uni, 'uva') ~ "University of Amsterdam",
        str_detect(uni, '(tilburg|uvt)') ~ "Tilburg University",
        str_detect(uni, 'eur') ~ "Erasmus University Rotterdam",
        str_detect(uni, 'leiden') ~ "Leiden University",
        str_detect(uni, 'vu') ~ "Vrije Universiteit",
        .default = NA_character_
    )
  )|>
  filter(!is.na(university)) |>
  select(-uni) |>
  distinct(uid, university, discipline, year, .keep_all = TRUE)
```


```{r}
#| warning: false
#| fig.width: 10.5
dems_long |>
  group_by(university, discipline, year) |>
  arrange(university, discipline, year) |>
  summarise(n = n()) |>
  filter(n > 5) |>
  tidyplot(y = discipline, x = n, color = university) |>
  add_barstack_absolute()  |>
  adjust_size(width = 8, height = 3)|>
  split_plot(by = year, ncol = 1)
```

## Network Statistics


```{r}
library(lubridate)
library(igraph)
library(Matrix)

priority <- c("UVA","EUR","RUG","VU","Leiden","RU","UU","Tilburg")
dems = dems %>%
  mutate(
    position = coalesce(!!!select(., starts_with("position_"))),
    university = coalesce(!!!select(., starts_with("university_")))
  ) |>
  mutate(
    s = str_to_lower(str_squish(university)),
    university = case_when(
      str_detect(s, "\\brug\\b|groningen") ~ "RUG",
      str_detect(s, "\\bvu\\b|vrij\\s*e?\\s*universiteit") ~ "VU",
      str_detect(s, "\\buva\\b|universiteit van amsterdam|univ(?:ersity)? of amsterdam") ~ "UVA",
      str_detect(s, "\\beur\\b|erasmus") ~ "EUR",
      str_detect(s, "\\bleiden\\b") ~ "LU",
      str_detect(s, "\\bru\\b|radboud") ~ "RU",
      str_detect(s, "\\buu\\b|utrecht") ~ "UU",
      str_detect(s, "\\buvt\\b|universiteit tilburg|\\btilburg\\b") ~ "TU",
      TRUE ~ NA_character_
    )
  )

edge = data[['edge_table']] |>
  filter(publication_date > ymd("2019-01-01"))
```


```{r}
# deps
library(dplyr)
library(igraph)
library(Matrix)

build_matrices <- function(D, E, directed = FALSE, weighted = TRUE,
                           sign_incidence = TRUE, exclude_isolates = FALSE) {
  stopifnot("uid" %in% colnames(D))
  stopifnot(all(c("uid", "a__uid") %in% colnames(E)))

  # --- 1) Clean edges: only between known nodes, drop NAs & self-loops
  nodes_vec <- unique(as.character(D$uid))
  Eclean <- E %>%
    transmute(uid = as.character(uid), a__uid = as.character(a__uid)) %>%
    filter(!is.na(uid), !is.na(a__uid),
           uid %in% nodes_vec, a__uid %in% nodes_vec,
           uid != a__uid)

  # --- 2) Aggregate parallel edges (counts -> weights if weighted)
  Eagg <- if (weighted) {
    Eclean %>% count(uid, a__uid, name = "weight")
  } else {
    Eclean %>% distinct(uid, a__uid) %>% mutate(weight = 1L)
  }

  # --- 3) Build graph (vertex order = nodes_vec)
  g <- graph_from_data_frame(Eagg, directed = directed,
                             vertices = tibble(name = nodes_vec))

  # --- 4) Optionally drop isolates (no edges to others)
  if (exclude_isolates) {
    keep_vids <- which(degree(g, mode = if (directed) "all" else "total") > 0)
    g <- induced_subgraph(g, vids = keep_vids)
  }

  # refresh node vector in case isolates were dropped
  nodes_vec_keep <- V(g)$name
  n <- length(nodes_vec_keep)

  # --- 5) Adjacency (sparse)
  A <- as_adjacency_matrix(
    g,
    attr   = if (weighted) "weight" else NULL,
    sparse = TRUE
  )
  if (!directed) A <- forceSymmetric(A, uplo = "U")

  # --- 6) Node × Edge incidence (sparse)
  el <- as_data_frame(g, what = "edges")
  m  <- nrow(el)

  if (m == 0L) {
    B <- sparseMatrix(i = integer(), j = integer(), x = numeric(),
                      dims = c(n, 0), dimnames = list(nodes_vec_keep, character()))
    edge_weight <- numeric(0)
  } else {
    tail_idx <- match(el$from, nodes_vec_keep)
    head_idx <- match(el$to,   nodes_vec_keep)

    row_idx <- c(tail_idx, head_idx)
    col_idx <- rep(seq_len(m), each = 2L)

    vals <- if (directed && sign_incidence) c(rep(-1, m), rep(+1, m)) else rep(1, 2L * m)

    B <- sparseMatrix(
      i = row_idx, j = col_idx, x = vals,
      dims = c(n, m),
      dimnames = list(nodes_vec_keep, paste0("e", seq_len(m)))
    )
    edge_weight <- if ("weight" %in% names(el)) el$weight else rep(1, m)
  }

  # --- 7) Ego × Alter rectangular counts (uid rows × a__uid cols), restricted to kept nodes
  Eclean_keep <- Eclean %>%
    filter(uid %in% nodes_vec_keep, a__uid %in% nodes_vec_keep)

  if (nrow(Eclean_keep) == 0L) {
    EgoAlter <- Matrix(0, nrow = n, ncol = n,
                       dimnames = list(nodes_vec_keep, nodes_vec_keep), sparse = TRUE)
  } else {
    tbl <- xtabs(~ uid + a__uid, Eclean_keep)
    # ensure full square ordering over nodes_vec_keep
    EgoAlter <- Matrix(0, nrow = n, ncol = n,
                       dimnames = list(nodes_vec_keep, nodes_vec_keep), sparse = TRUE)
    EgoAlter[rownames(tbl), colnames(tbl)] <- Matrix(tbl, sparse = TRUE)
  }

  # --- 8) Slice D to non-isolates and return as D
  D_out <- D %>% filter(as.character(uid) %in% nodes_vec_keep)

  list(
    graph        = g,
    A            = A,            # adjacency (n × n)
    B            = B,            # node–edge incidence (n × m)
    D            = D_out,        # nodes tibble, non-isolates only (if exclude_isolates=TRUE)
    edge_weight  = edge_weight,  # weights aligned to B's columns
    EgoAlter     = EgoAlter      # uid × a__uid counts among kept nodes
  )
}
```


```{r}
res <- build_matrices(dems, edge, directed = FALSE, weighted = TRUE, sign_incidence = TRUE, exclude_isolates = TRUE)

A <- res$A          # adjacency (sparse Matrix)
B <- res$B          # incidence (sparse Matrix)
w <- res$edge_weight
GA <- res$graph
EA <- res$EgoAlter  # uid × a__uid rectangular matrix

# Example: verify dimensions/names
dim(A); dim(B); head(rownames(A)); head(colnames(B))
```


```{r}
library(igraph)
library(ggraph)
library(ggplot2)
library(dplyr)
library(scales)

# ---------------------------------------------------------------
# 0) Helper: attach node attributes from D by uid
# ---------------------------------------------------------------
attach_node_attrs <- function(g, D, id_col = "uid") {
  # ensure character ids
  D <- D %>% mutate(!!id_col := as.character(.data[[id_col]]))
  V(g)$name <- as.character(V(g)$name)

  # align D to vertex order
  idx  <- match(V(g)$name, D[[id_col]])
  keep <- !is.na(idx)

  # add every column in D (except id) as a vertex attribute
  for (col in setdiff(names(D), id_col)) {
    vals <- D[[col]][idx[keep]]

    # coerce to simple atomic vectors (no tibbles/lists)
    if (is.factor(vals)) vals <- as.character(vals)
    if (is.list(vals)) {
      vals <- vapply(vals, function(z)
        if (length(z) == 0) NA_character_
        else if (length(z) == 1) as.character(z)
        else paste(z, collapse = "; "), character(1))
    }

    g <- set_vertex_attr(g, name = col, index = which(keep), value = vals)
  }
  g
}
# ---------------------------------------------------------------
# 1) One-mode plot from adjacency or existing GA
#    - sizes = strength (weighted degree)
#    - colors = gender; shapes = position; facets possible by university
# ---------------------------------------------------------------
plot_from_adjacency <- function(g, title = "Network (one-mode)", directed = FALSE) {
  # compute basic metrics for aesthetics
  V(g)$strength  <- graph.strength(g, weights = E(g)$weight %||% NULL) 
  V(g)$degree    <- degree(g, mode = if (directed) "all" else "total")
  E(g)$ewidth    <- rescale(E(g)$weight %||% rep(1, gsize(g)), to = c(0.2, 1.8))

  set.seed(42)
  ggraph(g, layout = "fr") +
    geom_edge_link(aes(width = ewidth),
                   edge_alpha = 0.25,
                   colour = "grey40",
                   lineend = "round",
                   arrow = if (directed) arrow(length = unit(2, "mm")) else NULL,
                   end_cap = if (directed) circle(2, "mm") else square(0.5, "mm")) +
    geom_node_point(aes(size = degree, color = university, shape = gender),
                    alpha = 0.9, stroke = 0.3) +
    scale_size_continuous(range = c(2, 9), breaks = pretty_breaks(4)) +
    scale_edge_width(range = c(0.2, 1.8), guide = "none") +
    guides(color = guide_legend(override.aes = list(size = 5)),
           shape = guide_legend(override.aes = list(size = 5))) +
    theme_graph(base_family = "sans") +
    labs(title = title, size = "Strength", shape = "Gender", color = "Position")
}


gA <- graph_from_adjacency_matrix(A, mode = "undirected", weighted = TRUE, diag = FALSE)
# gA <- attach_node_attrs(A, dems, id_col = "uid")
# plot_from_adjacency((gA))
```


```{r}

# plotly
# kateto.net
```

```{r}
library(dplyr)
library(scales)      # for rescale()
library(visNetwork)

data <- toVisNetworkData(gA)

N <- data$nodes |>
  inner_join(
    dems |>
      distinct(uid, .keep_all = TRUE) |>
      select(uid, clean_name, gender, origin, h_index, position, university),
    by = join_by(id == uid)
  ) |>
  # build per-node aesthetics
  mutate(
    label = ifelse(!is.na(clean_name) & nzchar(clean_name), clean_name, as.character(id)),
    university = ifelse(is.na(university) | !nzchar(university), "(Unknown)", university)
  )

# ---- color by university ----
uni_levels <- sort(unique(N$university))
pal <- scales::hue_pal()(length(uni_levels))
col_map <- setNames(pal, uni_levels)

# ---- size by h_index + shape by gender ----
N <- N |>
  mutate(
    color = unname(col_map[university]),
    size  = if (all(is.na(h_index))) 20 else
              suppressWarnings(rescale(h_index, to = c(12, 40),
                                       from = range(h_index, na.rm = TRUE))),
    size  = ifelse(is.na(size), 12, size),
    gender_clean = tolower(as.character(gender)),
    shape = dplyr::recode(
      gender_clean,
      "male" = "triangle",
      "female" = "dot",
      .default = "square"
    ),
    title = paste0(
      "<b>", label, "</b>",
      "<br/>University: ", university,
      "<br/>Gender: ", ifelse(is.na(gender) | !nzchar(gender), "Unknown", gender),
      "<br/>h-index: ", ifelse(is.na(h_index), "NA", h_index)
    )
  ) |>
  select(-gender_clean)  # cleanup helper col

E <- data$edges

p = visNetwork(nodes = N, edges = E, height = "800px", width = "100%") |>
  visNodes(borderWidth = 1) |>
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) |>
  visPhysics(solver = "barnesHut") |>
  visLayout(randomSeed = 42) |>
  visLegend()

p |> visSave(file = "full_network.html")
p
# library(webshot)
# webshot("full_network.html", zoom = 2, delay = 80, file = "full_network.png")
```


```{r}
# deps
library(igraph)
library(dplyr)
library(tidyr)

# ---------- helpers (unchanged) ----------
.harmonic_closeness <- function(g, weights = NULL) {
  D <- suppressWarnings(distances(g, weights = weights))
  diag(D) <- Inf
  (rowSums(1 / D, na.rm = TRUE)) / (vcount(g) - 1)
}
.same_attribute_share <- function(g, attr_vec) {
  if (is.null(attr_vec)) return(rep(NA_real_, vcount(g)))
  nbrs <- ego(g, order = 1, nodes = V(g), mode = "all")
  sapply(seq_along(nbrs), function(i) {
    neigh <- nbrs[[i]]
    if (length(neigh) == 0) return(NA_real_)
    mean(attr_vec[neigh] == attr_vec[i])
  })
}

# ========================================================================
# A) From res$A / res$graph / res$D  (one-mode adjacency)
# ========================================================================
compute_node_stats_from_res_adjacency <- function(
  res,
  node_id_col = "uid",
  use_weights = TRUE,
  invert_weights_for_paths = TRUE,
  attr_col = NULL
){
  stopifnot(all(c("A","graph","D") %in% names(res)))
  A     <- res$A
  nodes <- res$D
  mode  <- if (is_directed(res$graph)) "directed" else "undirected"

  g <- graph_from_adjacency_matrix(
    A,
    mode     = mode,
    weighted = use_weights,
    diag     = FALSE
  )
  V(g)$name <- rownames(A)

  w_strength <- if (use_weights && "weight" %in% edge_attr_names(g)) E(g)$weight else NULL
  w_dist <- if (!is.null(w_strength) && invert_weights_for_paths) {
    if (any(w_strength <= 0, na.rm = TRUE)) NULL else 1 / w_strength
  } else w_strength

  gu <- simplify(as_undirected(g, mode = "collapse", edge.attr.comb = "sum"),
                 remove.multiple = TRUE, remove.loops = TRUE)

  deg_all <- degree(g, mode = "all")
  str_all <- graph.strength(g, mode = "all", weights = w_strength)
  deg_in  <- if (mode == "directed") degree(g, mode = "in")  else NULL
  deg_out <- if (mode == "directed") degree(g, mode = "out") else NULL
  str_in  <- if (mode == "directed") graph.strength(g, mode = "in",  weights = w_strength) else NULL
  str_out <- if (mode == "directed") graph.strength(g, mode = "out", weights = w_strength) else NULL

  btw <- betweenness(g, directed = (mode == "directed"), weights = w_dist, normalized = TRUE)
  clo_harm <- .harmonic_closeness(g, weights = w_dist)

  eig  <- tryCatch(eigen_centrality(g, directed = (mode == "directed"), weights = w_strength)$vector,
                   error = function(e) rep(NA_real_, vcount(g)))
  pr   <- page_rank(g, directed = (mode == "directed"), weights = w_strength)$vector
  hubs <- if (mode == "directed") hub_score(g, weights = w_strength)$vector else rep(NA_real_, vcount(g))
  auth <- if (mode == "directed") authority_score(g, weights = w_strength)$vector else rep(NA_real_, vcount(g))

  clustering_local <- transitivity(gu, type = "local", isolates = "zero")
  triangles_cnt    <- count_triangles(gu, vids = V(gu))
  coreness_k       <- coreness(gu, mode = "all")
  constraint_burt  <- constraint(g, weights = w_strength)
  knn_vec          <- knn(gu)$knn
  ecc <- suppressWarnings(tryCatch(eccentricity(g, weights = w_dist),
                                   error = function(e) rep(NA_real_, vcount(g))))
  comp <- components(g)$membership
  comm_louvain <- membership(cluster_louvain(gu, weights = E(gu)$weight))

  same_attr_share <- {
    if (is.null(attr_col) || !(attr_col %in% names(nodes))) rep(NA_real_, vcount(g)) else {
      attr_vec <- nodes[[attr_col]][match(V(g)$name, nodes[[node_id_col]])]
      .same_attribute_share(gu, attr_vec)
    }
  }

  out <- tibble(
    !!node_id_col := V(g)$name,
    degree        = as.numeric(deg_all),
    strength      = as.numeric(str_all),
    degree_in     = if (!is.null(deg_in))  as.numeric(deg_in)  else NA_real_,
    degree_out    = if (!is.null(deg_out)) as.numeric(deg_out) else NA_real_,
    strength_in   = if (!is.null(str_in))  as.numeric(str_in)  else NA_real_,
    strength_out  = if (!is.null(str_out)) as.numeric(str_out) else NA_real_,
    betweenness   = as.numeric(btw),
    closeness_harmonic = as.numeric(clo_harm),
    eigenvector   = as.numeric(eig),
    pagerank      = as.numeric(pr),
    hubs          = as.numeric(hubs),
    authorities   = as.numeric(auth),
    clustering_local = as.numeric(clustering_local),
    triangles     = as.numeric(triangles_cnt),
    coreness_k    = as.numeric(coreness_k),
    constraint_burt = as.numeric(constraint_burt),
    avg_neighbor_degree = as.numeric(knn_vec),
    eccentricity  = as.numeric(ecc),
    component     = as.integer(comp),
    community_louvain = as.integer(comm_louvain),
    same_attr_share = as.numeric(same_attr_share)
  )

  nodes %>% left_join(out, by = setNames(node_id_col, node_id_col))
}

# ========================================================================
# B) From res$B / res$D  (two-mode incidence -> one-mode projection)
#     project = "rows" uses node rows (recommended here)
# ========================================================================
compute_node_stats_from_res_incidence <- function(
  res,
  node_id_col = "uid",
  project = c("rows","cols"),
  invert_weights_for_paths = TRUE,
  attr_col = NULL
){
  stopifnot(all(c("B","D") %in% names(res)))
  project <- match.arg(project)
  B     <- res$B
  nodes <- res$D

  g_bi <- graph_from_biadjacency_matrix(B, weighted = TRUE)
  V(g_bi)$name <- c(rownames(B), colnames(B))

  pr <- bipartite_projection(g_bi, multiplicity = TRUE)
  g1 <- if (project == "rows") pr$proj1 else pr$proj2

  gu <- simplify(g1, remove.multiple = TRUE, remove.loops = TRUE)
  w_strength <- if ("weight" %in% edge_attr_names(gu)) E(gu)$weight else NULL
  w_dist <- if (!is.null(w_strength) && invert_weights_for_paths) {
    if (any(w_strength <= 0, na.rm = TRUE)) NULL else 1 / w_strength
  } else w_strength

  deg   <- degree(gu)
  stren <- graph.strength(gu, weights = E(gu)$weight)
  btw   <- betweenness(gu, directed = FALSE, weights = w_dist, normalized = TRUE)
  clo_h <- .harmonic_closeness(gu, weights = w_dist)
  eig   <- eigen_centrality(gu, directed = FALSE, weights = E(gu)$weight)$vector
  pr    <- page_rank(gu, directed = FALSE, weights = E(gu)$weight)$vector
  clust <- transitivity(gu, type = "local", isolates = "zero")
  tri   <- count_triangles(gu, vids = V(gu))
  corek <- coreness(gu)
  cons  <- constraint(gu, weights = E(gu)$weight)
  knn_v <- knn(gu)$knn
  ecc   <- suppressWarnings(tryCatch(eccentricity(gu, weights = w_dist),
                                     error = function(e) rep(NA_real_, vcount(gu))))
  comp  <- components(gu)$membership
  comm  <- membership(cluster_louvain(gu, weights = E(gu)$weight))

  same_attr_share <- {
    if (is.null(attr_col) || !(attr_col %in% names(nodes))) rep(NA_real_, vcount(gu)) else {
      attr_vec <- nodes[[attr_col]][match(V(gu)$name, nodes[[node_id_col]])]
      .same_attribute_share(gu, attr_vec)
    }
  }

  out <- tibble(
    !!node_id_col := V(gu)$name,
    degree        = as.numeric(deg),
    strength      = as.numeric(stren),
    betweenness   = as.numeric(btw),
    closeness_harmonic = as.numeric(clo_h),
    eigenvector   = as.numeric(eig),
    pagerank      = as.numeric(pr),
    clustering_local = as.numeric(clust),
    triangles     = as.numeric(tri),
    coreness_k    = as.numeric(corek),
    constraint_burt = as.numeric(cons),
    avg_neighbor_degree = as.numeric(knn_v),
    eccentricity  = as.numeric(ecc),
    component     = as.integer(comp),
    community_louvain = as.integer(comm),
    same_attr_share = as.numeric(same_attr_share)
  )

  nodes %>% left_join(out, by = setNames(node_id_col, node_id_col))
}
```


```{r}
# # adjacency-based (uses res$A, res$graph, res$D)
# stats_adj <- compute_node_stats_from_res_adjacency(res, attr_col = "gender")

# # incidence-based row projection (uses res$B, res$D)
# stats_inc <- compute_node_stats_from_res_incidence(res, project = "rows", attr_col = "gender")
```


```{r}
# stats_adj |>
#   group_by(gender) |>
#   summarise(
#     degree = mean(degree),
#     betweenness = mean(betweenness),
#     closeness_harmonic = mean(closeness_harmonic),
#     eigenvector = mean(eigenvector),
#     clustering_local = mean(clustering_local),
#     coreness_k = mean(coreness_k),
#     constraint_burt = mean(constraint_burt)
#   )
```

```{r}
# stats_adj |>
#   group_by(origin) |>
#   summarise(
#     degree = mean(degree),
#     betweenness = mean(betweenness),
#     closeness_harmonic = mean(closeness_harmonic),
#     eigenvector = mean(eigenvector),
#     clustering_local = mean(clustering_local),
#     coreness_k = mean(coreness_k),
#     constraint_burt = mean(constraint_burt)
#   )
```

```{r}
# stats_adj |>
#   group_by(position) |>
#   summarise(
#     degree = mean(degree),
#     betweenness = mean(betweenness),
#     closeness_harmonic = mean(closeness_harmonic),
#     eigenvector = mean(eigenvector),
#     clustering_local = mean(clustering_local),
#     coreness_k = mean(coreness_k),
#     constraint_burt = mean(constraint_burt)
#   )
```

```{r}
# stats_adj |>
#   group_by(university) |>
#   summarise(
#     degree = mean(degree),
#     betweenness = mean(betweenness),
#     closeness_harmonic = mean(closeness_harmonic),
#     eigenvector = mean(eigenvector),
#     clustering_local = mean(clustering_local),
#     coreness_k = mean(coreness_k),
#     constraint_burt = mean(constraint_burt)
#   )
```