---
title: Descriptive Statistics
date: today

# eval: false
---

## Getting Started

```{r}
#| warning: false
#| output: false
# clear the global environment
rm(list = ls())
gc()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(tidyplots)

# load custom functions 
source("src/utils/custom_functions.r")
.load_quarto_dependencies()

options(knitr.graphics.fig.width = 20)
tidyplots_options(
  unit = "cm"
)
```

## Application


```{r}
data = freadRDS2(file="scholars", location = "clean")
```


### Number of publications

To check the data quality of OpenAlex entries I inspect the number of
publications per year. This is a crude but common used strategy to assess the
completeness of repositories such as OpenAlex.

```{r}
#| warning: false
#| fig.width: 9.99

works = data[['works']] |>
  bind_rows() |>
  distinct(work_id, .keep_all = TRUE) 
  
works_count = works |>
  group_by(publication_year, type) |>
  summarise(n = n()) |>
  ungroup() |>
  filter(publication_year >= 1960)

works_count_plot = works_count |>
  tidyplot(x = publication_year, y = n, color = type) |>
  add_areastack_absolute() |>
  adjust_size(width = 15, height = 7) |>
  adjust_x_axis_title("Publication Year") |>
  adjust_x_axis(limits = c(1960, 2026)) |>
  adjust_y_axis_title("Number of publications") |>
  adjust_legend_title("Publication Types")

works_count_plot

```


In 2019 there is an unusual spike in the number of publication. This is
predominantly driven by a couple of scholars publishing numerous datasets. Also,
there are a number of publication types that might not constitute a valid type
of publication. Therefore, I select only valid types and check the distribution
of publications again.

```{r}
#| warning: false
#| fig.width: 10.5
valid_publication_types = c(
  'article', 'book', 'book-chapter', 'dissertation', 
  'editorial', 'report', 'review',
  # optional, to consider:
  'erratum', 'letter', 'other', 'preprint'
)

works_count_plot2 = works_count |>
  filter(type %in% valid_publication_types)|>
  tidyplot(x = publication_year, y = n, color = type) |>
  add_barstack_absolute() |>
  adjust_size(width = 15, height = 7) |>
  adjust_x_axis_title("Publication Year") |>
  adjust_x_axis(limits = c(1960, 2026)) |>
  adjust_y_axis_title("Number of publications") |>
  adjust_legend_title("Publication Types")

works_count_plot2
```

### Number of scholars

```{r}
dems = data[['demographics']] 

dems_long = dems |>
  pivot_longer(
    # only the repeated fields:
    cols = matches("^(email|university|position|discipline)_\\d{2}$"),
    # split into the base name (becomes columns) and the 2-digit year
    names_to = c(".value", "year2"),
    names_pattern = "^(.*)_(\\d{2})$",
    values_drop_na = TRUE  # drop rows where all new values are NA
  ) |>
  mutate(year = as.integer(paste0("20", year2))) |>
  select(-year2) |>
  arrange(uid, year) |>
  mutate(university = na_if(university, "")) |>
  separate_rows(university, sep = "\\s*/\\s*")
```


```{r}
dems_long = dems_long |>
  mutate(
    uni = str_to_lower(university),
    university = case_when(
        str_detect(uni, 'rug') ~ "University of Groningen",
        str_detect(uni, 'ru') ~ "Radboud University",
        str_detect(uni, '(uu|ucu)') ~ "Utrecht University",
        str_detect(uni, 'uva') ~ "University of Amsterdam",
        str_detect(uni, '(tilburg|uvt)') ~ "Tilburg University",
        str_detect(uni, 'eur') ~ "Erasmus University Rotterdam",
        str_detect(uni, 'leiden') ~ "Leiden University",
        str_detect(uni, 'vu') ~ "Vrije Universiteit",
        .default = NA_character_
    )
  )|>
  filter(!is.na(university)) |>
  select(-uni) |>
  distinct(uid, university, discipline, year, .keep_all = TRUE)
```


```{r}
#| warning: false
#| fig.width: 10.5
dems_long |>
  group_by(university, discipline, year) |>
  arrange(university, discipline, year) |>
  summarise(n = n()) |>
  filter(n > 5) |>
  tidyplot(y = discipline, x = n, color = university) |>
  add_barstack_absolute()  |>
  adjust_size(width = 8, height = 3)|>
  split_plot(by = year, ncol = 1)
```
