---
title: Operationalize Variables
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions

---

## Getting Started

```{r}
#| warning: false
#| output: false
# load custom functions
source("src/utils/custom_functions.r")

# clear the global environment and set dependencies
.clear_global_environment()
.load_quarto_dependencies()
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(slider)
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(purrr)
```

## Functions

### `categorize_ethnicity()`

Classify scholars into migrant-origin groups based on their country of origin code.

#### Arguments

**`data`**

> Named list containing a `demographics` element with at least a `country_id` column.

#### Method

The function defines a set of minority country codes corresponding to the four largest migrant groups in the Netherlands (`MA`, `TR`, `BQ`, `CW`, `SX`, `AW`, `SR`) and adds a `migrant` column. Scholars from the Netherlands are coded `"NL"`, scholars from minority-coded countries are coded `"M4"`, scholars with a missing country code receive `NA`, and all others are coded `"other"`. The new column is relocated to follow `country_id`.

```{r}
categorize_ethnicity = function(data){
  minority_codes =  c('MA', 'TR', 'BQ', 'CW', 'SX', 'AW', 'SR')
  country_codes = c('US', 'DE', 'BE')

  data[['demographics']] = data[['demographics']] |>
    mutate(migrant = case_when(
      country_id == 'NL' ~ 'NL',
      # country_id %in% country_codes ~ 'B3', # Big 3 origins
      country_id %in% minority_codes ~ 'M4', # 4 Biggest Migrant groups NL
      is.na(country_id) ~ NA_character_,
      .default = "other"
    )) |>
    relocate(migrant, .after = 'country_id')

  return(data)
}
```

### `clean_affilliations()`

Standardize university affiliation strings, extract the first valid affiliation per year, and count the number of affiliations per year.

#### Arguments

**`data`**

> Named list containing a `demographics` element with `university_*` columns (one per measurement year).

#### Method

The function first normalizes raw affiliation strings across all `university_*` columns by replacing `.` with `/`, mapping `LEIDEN` to `UL`, `RUG` to `UVG`, and `UCU` to `UU`. It then splits each normalized string on `/` and extracts the first token that appears in a predefined list of valid university codes, storing the result in new `university_*_first` columns. Finally, it counts the number of non-missing affiliation tokens per year and stores this in `university_*_k` columns.

```{r}
clean_affilliations = function(data){
  valid_universities = c(
    'uu', 'uva', 'rug', 'ru', 'ucu', 'leiden', 'eur',
    'uvt', 'uvg', 'vu', 'wur', 'ul'
  )

  data[['demographics']] = data[['demographics']] |>
    mutate(
      across(
        starts_with('university'),
        # clean and split university strings
        ~ .x |> str_replace('\\.', '/') |>
          str_replace(regex('LEIDEN', ignore_case = TRUE), 'UL')  |>
          str_replace('RUG', 'UVG') |>
          str_replace('UCU', 'UU')
      ),
      # get the first valid affiliation per author year
      across(
        starts_with('university'),
        # clean and split university strings
        ~ .x |> str_replace('\\.', '/') |>
          str_to_lower() |>
          str_split(pattern = '/') |>
          # select the first of valid universities
          sapply(\(chunks){
            chunks[chunks %in% valid_universities][1]
          }),
        .names = "{.col}_first"
      ),
      # calculate the number of affiliations
      across(
        starts_with('university') & !ends_with('first'),
        ~ .x |> str_replace('\\.', '/') |>
          str_to_lower() |>
          str_split(pattern = '/') |>
          sapply(\(chunks){
            length(chunks[!is.na(chunks)])
          }),
        .names = "{.col}_k"
      )
    )

  return(data)
}
```

### `set_time_matrix()`

Build a symmetric 8×8 travel-time distance matrix between Dutch universities.

#### Arguments

None.

#### Method

The function defines the upper triangle of a travel-time matrix as a named numeric vector and constructs an 8×8 matrix with row and column labels `UVG`, `VU`, `UVA`, `UL`, `EUR`, `UVT`, `UU`, and `RU`. The lower triangle is filled by mirroring the upper triangle, producing a symmetric matrix which is returned.

```{r}
set_time_matrix = function(){
  time_distance = c(
    UVG = c(0,    NA,   NA,    NA,    NA,    NA,    NA,    NA),
    VU  = c(2.52, 0,    NA,    NA,    NA,    NA,    NA,    NA),
    UVA = c(2.58, 0.52, 0,     NA,    NA,    NA,    NA,    NA),
    UL  = c(3.12, 0.91, 1.04,  0,     NA,    NA,    NA,    NA),
    EUR = c(3.17, 1.32, 1.38,  1.18,  0,     NA,    NA,    NA),
    EVT = c(3.30, 1.70, 1.57,  1.67,  1.35,  0,     NA,    NA),
    UU  = c(2.47, 0.98, 0.88,  1.40,  1.33,  1.36,  0,     NA),
    RU  = c(2.83, 1.90, 1.68,  2.23,  2.33,  1.30,  1.53,  0)
  )
  labs = c('UVG', 'VU', 'UVA', 'UL', 'EUR', 'UVT', 'UU', 'RU')

  # convert to symmetrical matrix
  T = matrix(time_distance,
    ncol = 8, nrow = 8, dimnames = list(labs, labs)
  )
  T[lower.tri(T)] = T[upper.tri(T)]
  return(T)
}
```

### `create_affiliations_long()`

Reshape demographics to a long (uid × year × affiliation) table.

#### Arguments

**`demographics`**

> A tibble with a `uid` column and one or more `university_*_first` columns (one per measurement year).

**`affix`**

> Character string suffix used to select the relevant affiliation columns. Defaults to `"first"`.

**`year_regex`**

> Regular expression used to extract the two-digit year from column names. Defaults to `"\\d{2}"`.

#### Method

The function selects `uid` and all columns ending in `affix`, pivots them to long format, extracts the two-digit year from the column name using `year_regex` and prepends `"20"` to produce a four-digit year string, retains `uid`, `year`, and `value`, and uppercases the affiliation codes.

```{r}
create_affiliations_long <- function(demographics,
                                     affix = "first",
                                     year_regex = "\\d{2}") {
  demographics %>%
    select(uid, ends_with(affix)) %>%
    pivot_longer(
      cols = ends_with(affix),
      names_to = "name",
      values_to = "value"
    ) %>%
    mutate(year = paste0("20", str_extract(name, regex(year_regex)))) %>%
    select(uid, year, value) %>%
    mutate(value = toupper(value))
}
```

### `create_time_lookup()`

Convert the travel-time matrix to a long-format lookup table for use in joins.

#### Arguments

**`T`**

> A square numeric matrix of travel-time distances between universities, as returned by `set_time_matrix()`.

#### Method

The function coerces the matrix to a long-format data frame via `as.table()`, converts it to a tibble, and renames the columns to `value_i`, `value_j`, and `distance` to align with the naming conventions used in dyad join operations.

```{r}
create_time_lookup <- function(T) {
  as.data.frame(as.table(T)) %>%
    as_tibble() %>%
    rename(value_i = Var1, value_j = Var2, distance = Freq)
}
```

### `create_pairs_with_distance()`

Build a dyadic table of all scholar pairs per year, enriched with institutional travel-time distances.

#### Arguments

**`affiliations`**

> A long (uid × year × affiliation) tibble as returned by `create_affiliations_long()`.

**`T_long`**

> A long-format travel-time lookup table as returned by `create_time_lookup()`.

**`drop_self`**

> Logical. If `TRUE`, self-pairs (`uid_i == uid_j`) are removed. Defaults to `FALSE`.

#### Method

The function creates a distinct uid–year–affiliation key table and crosses all uid pairs within each year using `tidyr::crossing()`. If `drop_self` is `TRUE`, self-pairs are removed. It then left-joins affiliation codes for both the focal scholar and partner, drops rows where either affiliation is missing, and joins on `T_long` to attach `distance`, dropping pairs for which no travel-time is available.

```{r}
create_pairs_with_distance <- function(affiliations, T_long, drop_self = FALSE) {
  aff_key <- affiliations %>% distinct(year, uid, value)

  pairs <- aff_key %>%
    group_by(year) %>%
    summarise(pairs = list(tidyr::crossing(uid_i = uid, uid_j = uid)), .groups = "drop") %>%
    unnest(pairs)

  if (drop_self) {
    pairs <- pairs %>% filter(uid_i != uid_j)
  }

  pairs %>%
    left_join(aff_key, by = c("year", "uid_i" = "uid")) %>%
    rename(value_i = value) %>%
    left_join(aff_key, by = c("year", "uid_j" = "uid")) %>%
    rename(value_j = value) %>%
    drop_na(value_i, value_j) %>%
    left_join(T_long, by = c("value_i", "value_j")) %>%
    drop_na(distance)
}
```

### `make_year_matrix()`

Populate a full uid × uid distance matrix for a single year from a dyad table.

#### Arguments

**`df_year`**

> A tibble with `uid_i`, `uid_j`, and `distance` columns for a single year, as produced by splitting the output of `create_pairs_with_distance()`.

**`all_uids`**

> Character vector defining the complete set of row and column names for the output matrix.

#### Method

The function initializes an `NA` matrix of dimension `length(all_uids) × length(all_uids)` with row and column names set to `all_uids`. It uses `match()` to resolve each dyad to row and column indices, discards pairs whose UIDs are not found in `all_uids`, and fills the corresponding cells with their `distance` values. The populated matrix is returned.

```{r}
make_year_matrix <- function(df_year, all_uids) {
  M <- matrix(
    NA_real_,
    nrow = length(all_uids),
    ncol = length(all_uids),
    dimnames = list(all_uids, all_uids)
  )

  ii <- match(df_year$uid_i, all_uids)
  jj <- match(df_year$uid_j, all_uids)
  keep <- !is.na(ii) & !is.na(jj)

  M[cbind(ii[keep], jj[keep])] <- df_year$distance[keep]
  M
}
```

### `create_distance_matrices()`

Produce a named list of uid × uid travel-time distance matrices, one per year.

#### Arguments

**`pairs_by_year`**

> A dyadic tibble with `year`, `uid_i`, `uid_j`, and `distance` columns, as returned by `create_pairs_with_distance()`.

**`all_uids`**

> Character vector defining the shared set of row and column names for all yearly matrices.

#### Method

The function selects the four relevant columns, splits the tibble by `year` using `group_split()`, names each list element with its year string, and maps `make_year_matrix()` over each split to produce a consistently-dimensioned named list of distance matrices.

```{r}
create_distance_matrices <- function(pairs_by_year, all_uids) {
  pairs_by_year %>%
    select(year, uid_i, uid_j, distance) %>%
    group_by(year) %>%
    group_split() %>%
    set_names(map_chr(., ~ unique(.x$year))) %>%
    map(~ make_year_matrix(.x, all_uids))
}
```

### `add_distances()`

Orchestrate the institutional travel-time distance pipeline and attach the resulting per-year matrices to data.

#### Arguments

**`data`**

> Named list containing a `demographics` element with processed `university_*_first` affiliation columns.

**`topics`**

> Named list containing a `dissim` sub-list whose `field` element provides the reference UID ordering used to dimension the output matrices.

#### Method

The function calls `create_affiliations_long()` to reshape affiliations to long format, `set_time_matrix()` and `create_time_lookup()` to build the travel-time lookup, and `create_pairs_with_distance()` to compute all pairwise institutional distances per year. It extracts the reference UID set from `topics[["dissim"]][["field"]][["2022"]]` and passes it to `create_distance_matrices()` to produce consistently-dimensioned per-year matrices. The resulting list is stored in `data[['distances']]` and the updated list is returned.

```{r}
add_distances <- function(data, topics){
  # affiliations
  affiliations <- create_affiliations_long(data[["demographics"]])

  # time lookup
  T = set_time_matrix()
  T_long <- create_time_lookup(T)

  # dyads + distance
  pairs_by_year <- create_pairs_with_distance(affiliations, T_long)

  # global uid set / ordering
  all_uids <- colnames(topics[["dissim"]][["field"]][["2022"]])

  # matrices per year
  data[['distances']] <- create_distance_matrices(pairs_by_year, all_uids)

  return(data)
}
```

## Application

```{r}
data = freadRDS2('scholars', location = "./data/clean/")
topics = freadRDS2('topics', location = './data/clean/')
```

```{r}
data = data |>
  categorize_ethnicity() |>
  clean_affilliations() |>
  add_distances(topics)
```

```{r}
fsaveRDS(data, 'data', location = "./data/analysis/")
fsaveRDS(topics, 'topics', location = "./data/analysis/")
```

---
## Migrate these plots to another document

```{r}
top_n <- 25

data[["demographics"]] |>
  count(country_id, migrant, name = "n") |>
  arrange(n) |>
  tail(25) |>
  mutate(country_id = factor(country_id, levels = unique(country_id))) |>
  ggplot(aes(x = country_id, y = n, fill = migrant)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Count", fill = "Migrant", title = "Counts by country_id, split by migrant") +
  theme_minimal()
```

```{r}
data[['demographics']] |>
  filter(!is.na(career_start)) |>
  ggplot(aes(x = career_start)) +
  geom_density() +
  labs(x = "Career start (year)", y = "Density", title = "Kernel density of career start") +
  theme_minimal()
```


```{r}
data[['demographics']] |>
  filter(!is.na(h_index_22)) |>
  ggplot(aes(x = h_index_22)) +
  geom_density() +
  labs(x = "h_index", y = "Density", title = "Kernel density of h-index in 2022") +
  theme_minimal()
```

```{r}
data[['demographics']] |>
  filter(!is.na(i10_index_21)) |>
  ggplot(aes(x = i10_index_21)) +
  geom_density() +
  labs(x = "i10_index", y = "Density", title = "Kernel density of i10-index in 2022") +
  theme_minimal()
```

```{r}
data[["demographics"]] |>
  group_by(position_22) |>
  summarise(has_valid_works = mean(!is.na(career_start)), .groups = "drop") |>
  arrange(desc(has_valid_works)) |>
  mutate(position_22 = factor(position_22, levels = position_22)) |>
  ggplot(aes(x = position_22, y = has_valid_works)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = NULL,
    y = "Share with valid works",
    title = "Valid works by position (2025)"
  ) +
  theme_minimal()
```

```{r}
years <- c("2022", "2024", "2025")

dens_df <- map_dfr(years, function(y) {
  m <- topics[["dissim"]][["subfield"]][[y]]
  tibble(
    year = y,
    dissim = m[upper.tri(m)]
  )
}) |>
  filter(!is.na(dissim))

ggplot(dens_df, aes(x = dissim)) +
  geom_density() +
  facet_wrap(~ year, nrow = 1, scales = "free_y") +
  theme_minimal() +
  labs(x = "Dissimilarity (upper triangle)", y = "Density")
```

```{r}
years <- c("2022", "2024", "2025")

dens_df <- map_dfr(years, function(y) {
  m <- topics[["dissim"]][["field"]][[y]]
  tibble(
    year = y,
    dissim = m[upper.tri(m)]
  )
}) |>
  filter(!is.na(dissim))

ggplot(dens_df, aes(x = dissim)) +
  geom_density() +
  facet_wrap(~ year, nrow = 1, scales = "free_y") +
  theme_minimal() +
  labs(x = "Dissimilarity (upper triangle)", y = "Density")
```

```{r}
years <- c("2022", "2024", "2025")

dens_df <- map_dfr(years, function(y) {
  m <- topics[["dissim"]][["subfield"]][[y]]
  tibble(
    year = y,
    dissim = m[upper.tri(m)]
  )
}) |>
  filter(!is.na(dissim))

ggplot(dens_df, aes(x = dissim)) +
  geom_density() +
  facet_wrap(~ year, nrow = 1, scales = "free_y") +
  theme_minimal() +
  labs(x = "Dissimilarity (upper triangle)", y = "Density")
```

```{r}
plot_df <- data[["demographics"]] |>
  select(uid, gender, university_22_first, university_24_first, university_25_first) |>
  pivot_longer(
    cols = starts_with("university_"),
    names_to = "year",
    values_to = "university"
  ) |>
  mutate(
    year = case_when(
      year == "university_22_first" ~ "2022",
      year == "university_24_first" ~ "2024",
      year == "university_25_first" ~ "2025",
      TRUE ~ year
    )
  ) |>
  filter(!is.na(university), university != "", !is.na(gender), gender != "") |>
  count(year, university, gender, name = "n") |>
  group_by(year, university) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

ggplot(plot_df, aes(x = university, y = prop, fill = gender)) +
  geom_col() +
  facet_wrap(~ year, scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL, y = "Gender composition", fill = "Gender")
```


```{r}
plot_df <- data[["demographics"]] |>
  select(university_22_first, university_24_first, university_25_first) |>
  pivot_longer(
    cols = everything(),
    names_to = "panel",
    values_to = "university"
  ) |>
  filter(!is.na(university), university != "") |>
  count(panel, university, name = "n")

ggplot(plot_df, aes(x = university, y = n)) +
  geom_col() +
  facet_wrap(~ panel, scales = "free_x") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.spacing = unit(1, "lines")
  ) +
  labs(x = NULL, y = "Count")
```


```{r}
temp =  data[['demographics']] |>
  mutate(
    across(
      starts_with('university'),
      # clean and split university strings
      ~ .x |> str_replace('\\.', '/') |>
        str_replace(regex('LEIDEN', ignore_case = TRUE), 'UL')  |>
        str_replace('RUG', 'UVG') |>
        str_replace('UCU', 'UU')
    )) |>
  select(university_22, university_24, university_25)
```
