---
title: "Custom Functions"
---

## Functions

### Load core Quarto dependencies
This helper loads core packages needed for Quarto rendering (knitr and rmarkdown).  
Use it at the start of documents to ensure the required engines are available.
```{r}
#| id: load-quarto-dependencies
.load_quarto_dependencies = function(){
  library(knitr)
  library(rmarkdown)
}
```

### (Deprecated) Check and load packages

Legacy helper for checking and loading packages.
It now only issues a deprecation warning and advises using renv plus explicit `library()` calls instead.
```{r}
fpackage.check <- function(packages) {
  lifecycle::deprecate_warn(
    when = "2025-10-27",
    what = "fpackage.check()",
    details = paste0("Use renv for package installation and library()",
                     " for loading packages.")
  )
  message(
    "⚠️ fpackage.check() is deprecated. Use renv + library() calls instead."
  )
  
  lapply(packages, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  })

  invisible(TRUE)
}
```


### Save object to dated .RData file

Saves an R object to a .RData file in a processed data directory,
automatically prefixing the filename with the current date (YYYYMMDD).
```{r}
fsave <- function(x, file, location = "./data/processed/", ...) {
  lifecycle::deprecate_warn(
    when = "2025-10-27",
    what = "fsave()",
    details = paste0(
      "Use fsaveRDS or fsaveRDS2 to save data instead. function is",
      "deprecated in favor of renv support which conflicts with load"
    )
  )
  message(
    "⚠️ fsave() is deprecated. Use fsaveRDS or fsaveRDS2  calls instead."
  )
  # if directory does not exist, create it
  if (!dir.exists(location)) dir.create(location)
  
  # create filename with date
  datename <- substr(gsub("[:-]", "", Sys.time()), 1, 8)
  totalname <- paste(location, datename, file, sep = "")

  # save file
  print(paste("SAVING: ", totalname, sep = ""))
  save(x, file = totalname)
}
```

### Save object to dated RDS file

Saves an R object to a compressed `.Rds` file in a processed data directory,
automatically prefixing the filename with the current date (YYYYMMDD).
```{r}
fsaveRDS <- function(x, file, location = "./data/processed/", ...) {
  # if directory does not exist, create it
  if (!dir.exists(location)) dir.create(location)
  
  # create filename with date
  datename <- substr(gsub("[:-]", "", Sys.time()), 1, 8)
  totalname <- paste(location, datename, file, ".Rds", sep = "")

  # save file
  print(paste("SAVING: ", totalname, sep = ""))
  saveRDS(x, file = totalname, compress = "bzip2")
}
```

### Read an RDS file

Thin wrapper around `readRDS()` for symmetry with `fsaveRDS()`.
Given a file path, it loads and returns the stored object.
```{r}
freadRDS = function(fileName){
  # load file and return object
  readRDS(fileName)
}
```

### Read the most recent dated RDS file by name

Loads the latest (most recent by filename) .Rds file in a data subdirectory,
based on a base filename and a dated naming convention. Skips `old_` files by default.

```{r}
freadRDS2 = function(file, location = "processed", ...){
  # make sure that the location is nested within the data directory
  if (!str_detect(file, '.Rds')) file = paste0(file, ".Rds")

  # format location and create the location if it does not exist
  if (!str_detect(location, 'data/')) location = file.path('data', location)
  if (!file.exists(location)) warning("the location does not exist ", location)

  # list and filter by literal substring 'file'
  files = list.files(location, pattern = file)
  files = files[grepl(file, files, fixed = TRUE)]

  # exclude old_* unless the requested file itself starts with old_
  if (!startsWith(file, "old_")) files = files[!startsWith(files, "old_")] 
  if (startsWith(file, "old_")) files = files[startsWith(files, "old_")] 

  # create file_path for the most recent file
  file_name = sort(files, decreasing = TRUE, na.last = NA)[1]
  file_path = file.path(location, file_name)

  return(readRDS(file_path))
}
```

### Load .RData file and return its object

Loads a `.RData` file and returns the loaded object directly,
instead of leaving it in the global environment.

```{r}
fload <- function(filename) {
  load(filename)
  get(ls()[ls() != "filename"])
}

```

### Normalise personal names

Normalises name strings by transliterating Cyrillic to Latin,
stripping accents, lowercasing, and trimming whitespace for robust matching.
```{r}
normalize_name = function(x) {
  x = as.character(x)

  # mark Cyrillic
  has_cyr = stringi::stri_detect_charclass(x, "\\p{Script=Cyrillic}")

  # transliterate only Cyrillic -> Latin
  x[has_cyr] = stringi::stri_trans_general(x[has_cyr], "Cyrillic-Latin")

  # strip accents for all, lowercase, trim, squish
  x = stringi::stri_trans_general(x, "Latin-ASCII")
  x = tolower(x)
  x = str_squish(x)
  trimws(x)
}
# colorize <- function(x, color) {sprintf("<span style='color: %s;'>%s</span>", color, x) }
```

### Check whether code should be evaluated
Utility to check a `params$eval` (or `param$eval`) flag safely.
Returns `TRUE/FALSE` depending on the parameter, and `FALSE` if the parameter is missing.

```{r}
eval_ok <- tryCatch(isTRUE(params$eval), error = function(...) {
  tryCatch(isTRUE(param$eval), error = function(...) FALSE)
})
```