---
title: Network Descriptives
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions

execute:
  eval: false
---

## Getting Started

```{r}
#| warning: false
#| output: false
#| echo: false
BASEDIR = stringr::str_replace_all(getwd(), '/results', '')
setwd(BASEDIR)
```

```{r}
#| warning: false
#| output: false
# load custom functions
source("src/utils/custom_functions.r")

# clear the global environment and set dependencies
.clear_global_environment()
.load_quarto_dependencies()

BASEDIR = stringr::str_replace_all(getwd(), '/results', '')
setwd(BASEDIR)
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(sna)
library(RSiena)
library(igraph)
library(dplyr)
library(tibble)
library(purrr)
library(ape)
```

## Functions

### `make_adjacency_matrix()`

Build a co-authorship adjacency matrix for a specified set of scholars and time window.

#### Arguments

**`data`**

> Named list with a `works` element (named list of per-scholar works tibbles containing `uid`, `work_id`, `authorships`, `publication_date`, and `publication_year`).

**`uids`**

> Character vector of focal scholar UIDs defining the rows and columns of the output matrix.

**`type`**

> Character string controlling which author position is treated as the focal node. `"first"` and `"last"` use the author at that position; `"all"` creates a full crossing of all co-authors on each work. Defaults to `"first"`.

**`min_year`**

> Integer year or Date giving the lower bound of the publication window (inclusive).

**`max_year`**

> Integer year or Date giving the upper bound of the publication window (inclusive).

**`weighted`**

> Logical. If `TRUE`, matrix values reflect the count of shared publications; if `FALSE`, values are binarized to `1L`. Defaults to `FALSE`.

#### Method

The function binds the works for the requested `uids`, filters to the specified publication window (using `publication_date` for Date inputs and `publication_year` for integers), selects `uid`, `work_id`, and `authorships`, deduplicates by `work_id`, and unnests authorships. Depending on `type`, it either builds ego–alter pairs using the author at the named position or crosses all author pairs per work. Co-author pairs are counted and optionally binarized. The result is pivoted to a wide adjacency matrix, padded to the full `uids` dimension with zeros for absent pairs, and the diagonal is set to `0`.

```{r}
make_adjacency_matrix <- function(
    data, uids, type = 'first', min_year = 2020, max_year = 2026, weighted = FALSE
  ) {

  # make edgelist from works
  edges <- data[['works']][uids] |>
    purrr::keep(~ nrow(.x) > 0) |>
    bind_rows()

  if (inherits(min_year, "Date")) {
    edges <- edges |> filter(publication_date >= min_year, publication_date <= max_year)
  } else {
    edges <- edges |> filter(publication_year >= min_year, publication_year <= max_year)
  }
  edges <- edges |>
    select(uid, work_id, authorships) |>
    rename(e_uid = uid) |>
    distinct(work_id, .keep_all = TRUE) |>
    unnest(authorships)

  if (type != 'all') {
    edges <- edges |>
      group_by(work_id) |>
      mutate(
        e_uid = uid[author_position == type][1]
      ) |> ungroup() |>
      filter(!is.na(e_uid), e_uid != uid) |>
      select(work_id, e_uid, uid)
  } else {
    edges <- edges |>
      group_by(work_id) |>
      reframe(
        tidyr::crossing(
          e_uid = uid,
          uid   = uid
        )
      ) |>
      ungroup() |>
      filter(!is.na(e_uid), e_uid != uid) |>
      select(work_id, e_uid, uid)
  }

  # calculate the number of edges between ego and alters.
  edge_counts <- edges |> count(e_uid, uid, name="w")

  # convert counts to incidences
  if (!weighted) {
    edge_counts <- edge_counts %>% mutate(w = 1L)
  }

  # all nodes appearing as ego or alter
  nodes <- sort(unique(c(edge_counts$e_uid, edge_counts$uid)))

  # complete matrix grid and pivot to wide
  adj <- edge_counts %>%
    rename(from = e_uid, to = uid) %>%
    complete(from = nodes, to = nodes, fill = list(w = 0)) %>%
    pivot_wider(names_from = to, values_from = w) %>%
    arrange(from)

  # convert to matrix and set rownames
  mat <- adj %>% as.data.frame()

  row_ids <- mat$from
  mat$from <- NULL

  mat <- as.matrix(mat)
  rownames(mat) <- row_ids
  diag(mat) <- 0

  # make sure the matrices are consistent across waves
  all_uids <- as.character(uids)

  # indices of existing rows/cols in desired order
  ri <- match(all_uids, rownames(mat))
  ci <- match(all_uids, colnames(mat))

  # start with full NA matrix
  mat_full <- matrix(0L, nrow = length(all_uids), ncol = length(all_uids),
                    dimnames = list(all_uids, all_uids))

  # copy overlap block
  mat_full[!is.na(ri), !is.na(ci)] <- mat[ri[!is.na(ri)], ci[!is.na(ci)], drop = FALSE]
  mat <- mat_full

  mat
}
```

### `fcolnet2()`

Build a multi-wave co-authorship network filtered by university, position, and discipline.

#### Arguments

**`data`**

> Named list with a `demographics` element (containing `uid`, `university_*`, `discipline_*`, and `position_*` columns) and a `works` element used by `make_adjacency_matrix()`.

**`university`**

> Character vector of university abbreviations to include. Defaults to all eight valid Dutch universities.

**`position`**

> Character vector of academic positions to include. Defaults to all five valid positions.

**`discipline`**

> Character vector of disciplines to include. Defaults to `"Sociology"` and `"Political Sciences"`.

**`type`**

> Character string or `NULL`. If `NULL`, adjacency matrices are built for `"first"`, `"last"`, and `"all"` authorship types; otherwise the single specified type is used.

**`waves`**

> List of two-element vectors specifying the start and end of each observation wave. Defaults to three waves roughly corresponding to 2019–2022, 2023–2024, and 2025–2026.

#### Method

The function validates each filter argument against its allowed values using an internal helper. It filters `demographics` to scholars who appear with at least one of the specified universities, disciplines, and positions across the three measurement years. For each wave (and optionally each authorship type), it calls `make_adjacency_matrix()` with the corresponding year bounds. The function returns a list with `data` (the filtered demographics tibble) and `nets` (the nested list of adjacency matrices).

```{r}
fcolnet2 <- function(
    data,
    university = NULL,
    position   = NULL,
    discipline = NULL,
    type = NULL,
    waves = list(c(2019, 2022), c(2023, 2024), c(2025, 2026))
  ){

  valid_disciplines  <- c("Sociology", "Political Sciences")
  valid_universities <- c("UVA","EUR","UL","VU","UVG","UU","RU","UVT")
  valid_positions    <- c("Associate Professor","PhD Candidate","Full Professor",
                          "Researcher or Lecturer","Assistant Professor")
  valid_dates        <- list(c(ymd('20201219'), ymd('20221219')),
                             c(ymd('20221220'), ymd('20240419')),
                             c(ymd('20240420'), ymd('20260201')))

  # helper: set default + validate
  set_and_check <- function(x, valid, name = deparse(substitute(x))) {
    x <- if (is.null(x)) valid else x
    bad <- setdiff(unique(na.omit(x)), valid)
    if (length(bad)) stop("Invalid ", name, ": ", paste(bad, collapse = ", "), call. = FALSE)
    x
  }

  # set and check values
  discipline <- set_and_check(discipline, valid_disciplines,  "discipline")
  university <- set_and_check(university, valid_universities, "university")
  position   <- set_and_check(position,   valid_positions,    "position")
  waves      <- if(is.null(waves)) valid_dates else waves
  min_year   <- min(unlist(waves))
  max_year   <- max(unlist(waves))

  # step 1: make selection of nodes
  authors <- data[["demographics"]] |>
    filter(
      if_any(c(university_22, university_24, university_25),
            ~ .x |>
              toupper() |>
              str_detect(paste(university, collapse = "|"))),
      if_any(c(discipline_22, discipline_24, discipline_25),
            ~ .x %in% discipline),
      if_any(c(position_22, position_24, position_25),
            ~ .x %in% position)
    ) |>
    arrange(uid)
  uids <- authors |> pull(uid) |> sort()

  # step 3: create empty matrixes (wave, i, j)
  nwaves <- length(waves)
  nets <- list()
  if (is.null(type)){
    types = c('first', 'last', 'all')
    for (t in types){
      for (w in 1:length(waves)){
        nets[[t]][[w]] <- make_adjacency_matrix(
          data, uids, type = t, min_year = waves[[w]][1], max_year = waves[[w]][2])
      }
    }
  } else {
    for (w in 1:length(waves)){
      nets[[w]] <- make_adjacency_matrix(
        data, uids, type, min_year = waves[[w]][1], max_year = waves[[w]][2])
    }
  }

  # step 4: fill nets
  output <- list(
    data = authors,
    nets = nets
  )

  return(output)
}
```

### `harmonize_covariates()`

Standardize and rename covariate columns in a colnet data sub-list for use in analysis.

#### Arguments

**`data`**

> A colnet list whose `data` (or other named) sub-element contains demographics columns for `gender`, `discipline_*`, `position_*`, `position2_*`, and `university_*_first`.

**`what`**

> Character string naming the sub-element of `data` to transform. Defaults to `'data'`.

#### Method

The function adds a `female` integer indicator and a `soc` integer indicator for Sociology (using the first non-missing discipline across the three measurement years). It also reclassifies scholars whose `position2_*` columns indicate a postdoctoral role as `"Postdoctoral Researcher"` in the corresponding `position_*` columns. It renames `university_*_first` columns to `uni*p` and converts all `uni*p` columns to ordered factors using a predefined set of eight university abbreviations as levels.

```{r}
harmonize_covariates <- function(data, what = 'data'){
  uni_levels = c("EUR", "RU" , "UL", "UU", "UVA", "UVG", "UVT", "VU")

  data[[what]] = data[[what]] |>
    mutate(
      female = as.integer(gender == 'female'),
      soc = pmap_chr(
        list(discipline_22, discipline_24, discipline_25),
        ~ unique(na.omit(c(...)))[1]
      ),
      soc = as.integer(soc == 'Sociology'),
      position_22 = ifelse(str_detect(position2_22, 'Postdoctoral'), 'Postdoctoral Researcher', position_22),
      position_24 = ifelse(str_detect(position2_24, 'Postdoctoral'), 'Postdoctoral Researcher', position_24),
      position_25 = ifelse(str_detect(position2_25, 'Postdoctoral'), 'Postdoctoral Researcher', position_25)
    ) |>
    rename_with(
      ~ .x |>
        str_replace('university_', 'uni') |>
        str_replace('_first$', 'p')
    ) |>
    mutate(
      across(
        starts_with('uni') & ends_with('p'),
        ~ factor(tolower(.x),
        levels = tolower(uni_levels),
        ordered = TRUE)
      )
    )

  return(data)
}
```

### `make_graph_from_wave()`

Construct an igraph object from an adjacency matrix wave and attach node-level attributes.

#### Arguments

**`wave`**

> A square numeric adjacency matrix with row and column names set to scholar UIDs.

**`node_data`**

> A tibble of node attributes whose row order matches the adjacency matrix. All columns are attached as vertex attributes.

**`directed`**

> Logical. If `TRUE`, the resulting graph is directed. Defaults to `TRUE`.

#### Method

The function calls `igraph::graph_from_adjacency_matrix()` with the appropriate mode. It then asserts that the vertex count equals the number of rows in `node_data` and iterates over all columns of `node_data`, attaching each as a named vertex attribute via `set_vertex_attr()`.

```{r}
make_graph_from_wave <- function(wave, node_data, directed = TRUE) {
  graph <- igraph::graph_from_adjacency_matrix(
    wave,
    mode = if (directed) "directed" else "undirected",
    weighted = NULL,
    diag = FALSE,
    add.colnames = NULL
  )

  stopifnot(vcount(graph) == nrow(node_data))

  for (nm in names(node_data)) {
    graph <- set_vertex_attr(graph, nm, value = node_data[[nm]])
  }

  graph
}
```

### `detect_university_var()`

Identify the university affiliation variable present in a node data frame.

#### Arguments

**`node_data`**

> A data frame or tibble of node attributes, typically `colnet$data` or the vertex data frame of a graph.

#### Method

The function checks a set of candidate column names (`"university"`, `"university_first"`, `"first_affiliation"`, `"uni"`) against the columns present in `node_data`. It returns the first match, or stops with an error if none are found.

```{r}
detect_university_var <- function(node_data) {
  candidates <- c(
    "university", "university_first", "first_affiliation",
    "uni"
  )

  hit <- intersect(candidates, names(node_data))
  if (length(hit) == 0) {
    stop("No university variable found in colnet$data.")
  }
  hit[[1]]
}
```

### `detect_position_var()`

Identify the position variable present in a node data frame.

#### Arguments

**`node_data`**

> A data frame or tibble of node attributes, typically `colnet$data` or the vertex data frame of a graph.

#### Method

The function checks a set of candidate column names (`"position_22"`, `"position_24"`, `"position_25"`) against the columns present in `node_data`. It returns the first match, or stops with an error if none are found.

```{r}
detect_position_var <- function(node_data) {
  candidates <- c(
    "position_22", "position_24", "position_25"
  )

  hit <- intersect(candidates, names(node_data))
  if (length(hit) == 0) {
    stop("No position variable found in colnet$data.")
  }
  hit[[1]]
}
```

### `make_distance_weights()`

Build a binary weight matrix identifying all node pairs at an exact geodesic distance.

#### Arguments

**`graph`**

> An igraph object.

**`nb_distance`**

> Integer. The geodesic distance at which to set weights to `1`. Defaults to `1`.

**`mode`**

> Character string passed to `igraph::distances()`. One of `"out"`, `"in"`, or `"all"`. Defaults to `"all"`.

#### Method

The function computes the full pairwise distance matrix via `igraph::distances()`, sets cells equal to `nb_distance` to `1`, sets all other cells (including unreachable pairs) to `0`, and zeroes the diagonal.

```{r}
make_distance_weights <- function(graph, nb_distance = 1, mode = "all") {
  D <- distances(graph, mode = mode)
  W <- (D == nb_distance) * 1
  W[is.infinite(D)] <- 0
  diag(W) <- 0
  W
}
```

### `graph_morans_i()` (igraph distances)

Compute Moran's I for a numeric vertex attribute at a given graph-theoretic neighbor distance using igraph-based weights.

#### Arguments

**`graph`**

> An igraph object with the target variable set as a vertex attribute.

**`var`**

> Character string. Name of the numeric vertex attribute to test.

**`nb_distance`**

> Integer. Neighbor distance at which to define the spatial weights. Defaults to `1`.

**`mode`**

> Character string passed to `make_distance_weights()`. Defaults to `"all"`.

#### Method

The function extracts the vertex attribute, subsets to non-missing observations, and checks for minimum sample size and non-zero variance. It constructs binary spatial weights via `make_distance_weights()` and calls `ape::Moran.I()` (unscaled). Returns a one-row tibble with `variable`, `nb_distance`, `observed`, `expected`, `sd`, and `p_value`; all numeric fields are `NA` when the statistic cannot be computed.

```{r}
graph_morans_i <- function(graph, var, nb_distance = 1, mode = "all") {
  x <- vertex_attr(graph, var)

  if (!is.numeric(x)) {
    stop(sprintf("Variable '%s' must be numeric for Moran's I.", var))
  }

  keep <- !is.na(x)

  if (sum(keep) < 3) {
    return(tibble::tibble(
      variable = var,
      nb_distance = nb_distance,
      observed = NA_real_,
      expected = NA_real_,
      sd = NA_real_,
      p_value = NA_real_
    ))
  }

  g_sub <- induced_subgraph(graph, vids = V(graph)[keep])
  x_sub <- x[keep]

  # no variance -> Moran's I undefined
  if (length(unique(x_sub)) < 2 || isTRUE(all(stats::var(x_sub) == 0))) {
    return(tibble::tibble(
      variable = var,
      nb_distance = nb_distance,
      observed = NA_real_,
      expected = NA_real_,
      sd = NA_real_,
      p_value = NA_real_
    ))
  }

  W <- make_distance_weights(g_sub, nb_distance = nb_distance, mode = mode)

  # no usable ties at this distance
  if (all(W == 0) || sum(W) == 0) {
    return(tibble::tibble(
      variable = var,
      nb_distance = nb_distance,
      observed = NA_real_,
      expected = NA_real_,
      sd = NA_real_,
      p_value = NA_real_
    ))
  }

  out <- tryCatch(
    ape::Moran.I(x_sub, W, scaled = FALSE),
    error = function(e) NULL
  )

  if (is.null(out) || anyNA(c(out$observed, out$expected, out$sd, out$p.value))) {
    return(tibble::tibble(
      variable = var,
      nb_distance = nb_distance,
      observed = NA_real_,
      expected = NA_real_,
      sd = NA_real_,
      p_value = NA_real_
    ))
  }

  tibble::tibble(
    variable = var,
    nb_distance = nb_distance,
    observed = unname(out$observed),
    expected = unname(out$expected),
    sd = unname(out$sd),
    p_value = unname(out$p.value)
  )
}
```

### `graph_morans_i()` (sna geodistances)

Compute Moran's I for a numeric vertex attribute using `sna::geodist`-based weights (scaled). Overrides the igraph-based definition above.

#### Arguments

**`graph`**

> An igraph object with the target variable set as a vertex attribute.

**`var`**

> Character string. Name of the numeric vertex attribute to test.

**`nb_distance`**

> Integer. Neighbor distance at which to define the spatial weights. Defaults to `1`.

**`mode`**

> Unused in this version; retained for interface compatibility.

#### Method

Identical guards to the igraph-distances version (minimum sample size, non-zero variance, non-empty weight matrix). Constructs the binary weight matrix using `sna::geodist()` on the adjacency matrix extracted from the induced subgraph, selecting rows where geodesic distance equals `1`. Calls `ape::Moran.I()` with `scaled = TRUE`. Returns the same one-row tibble schema as the previous definition.

```{r}
graph_morans_i <- function(graph, var, nb_distance = 1, mode = "all") {
  x <- vertex_attr(graph, var)

  if (!is.numeric(x)) {
    stop(sprintf("Variable '%s' must be numeric for Moran's I.", var))
  }

  keep <- !is.na(x)

  if (sum(keep) < 3) {
    return(tibble::tibble(
      variable = var,
      nb_distance = nb_distance,
      observed = NA_real_,
      expected = NA_real_,
      sd = NA_real_,
      p_value = NA_real_
    ))
  }

  g_sub <- induced_subgraph(graph, vids = V(graph)[keep])
  x_sub <- x[keep]

  # no variance -> Moran's I undefined
  if (length(unique(x_sub)) < 2 || isTRUE(all(stats::var(x_sub) == 0))) {
    return(tibble::tibble(
      variable = var,
      nb_distance = nb_distance,
      observed = NA_real_,
      expected = NA_real_,
      sd = NA_real_,
      p_value = NA_real_
    ))
  }

  geodistances <- sna::geodist(as.matrix(as_adjacency_matrix(g_sub)), count.path = TRUE)
  geodistances <- geodistances$gdist
  W <- geodistances == 1

  # no usable ties at this distance
  if (all(W == 0) || sum(W) == 0) {
    return(tibble::tibble(
      variable = var,
      nb_distance = nb_distance,
      observed = NA_real_,
      expected = NA_real_,
      sd = NA_real_,
      p_value = NA_real_
    ))
  }

  out <- tryCatch(
    ape::Moran.I(x_sub, W, scaled = TRUE),
    error = function(e) NULL
  )

  if (is.null(out) || anyNA(c(out$observed, out$expected, out$sd, out$p.value))) {
    return(tibble::tibble(
      variable = var,
      nb_distance = nb_distance,
      observed = NA_real_,
      expected = NA_real_,
      sd = NA_real_,
      p_value = NA_real_
    ))
  }

  tibble::tibble(
    variable = var,
    nb_distance = nb_distance,
    observed = unname(out$observed),
    expected = unname(out$expected),
    sd = unname(out$sd),
    p_value = unname(out$p.value)
  )
}
```

### `prepare_graph_attributes()`

Add binary vertex attribute indicators for gender, discipline, university, and position to a graph.

#### Arguments

**`graph`**

> An igraph object, optionally without vertex attributes already attached.

**`node_data`**

> Optional tibble of node attributes. If provided, the graph is rebuilt via `make_graph_from_wave()` before attributes are added. Defaults to `NULL`.

#### Method

If `node_data` is supplied, the function first calls `make_graph_from_wave()`. It then extracts the vertex data frame and adds: a `gender_bin` indicator (`1` = male, `0` = female); a `discipline_bin` indicator from the `soc` column; one binary `university__<name>` indicator per unique university level; and a `full_prof_bin` indicator (`1` = full professor). Helper functions `detect_university_var()` and `detect_position_var()` are used to locate the relevant columns automatically.

```{r}
prepare_graph_attributes <- function(graph, node_data = NULL) {
  if (!is.null(node_data)) {
    graph <- make_graph_from_wave(
      wave = as_adjacency_matrix(graph, sparse = FALSE),
      node_data = node_data,
      directed = is_directed(graph)
    )
  }

  vdf <- igraph::as_data_frame(graph, what = "vertices")

  # gender: binary
  if ("gender" %in% names(vdf)) {
    gender_low <- tolower(as.character(vdf$gender))
    graph <- set_vertex_attr(
      graph, "gender_bin",
      value = case_when(
        gender_low %in% c("male", "man", "m") ~ 1,
        gender_low %in% c("female", "woman", "f") ~ 0,
        TRUE ~ NA_real_
      )
    )
  }

  # discipline: binary sociology indicator
  if ("soc" %in% names(vdf)) {
    graph <- set_vertex_attr(
      graph, "discipline_bin",
      value = vdf$soc
    )
  }

  # university: one binary indicator for each category
  uni_var <- tryCatch(detect_university_var(vdf), error = function(e) NULL)
  if (!is.null(uni_var)) {
    uni_vals <- as.character(vdf[[uni_var]])
    uni_levels <- sort(unique(stats::na.omit(uni_vals)))

    for (u in uni_levels) {
      nm <- paste0("university__", make.names(tolower(u)))
      graph <- set_vertex_attr(
        graph,
        nm,
        value = ifelse(uni_vals == u, 1, ifelse(is.na(uni_vals), NA, 0))
      )
    }
  }

  # full professor indicator
  pos_var <- tryCatch(detect_position_var(vdf), error = function(e) NULL)
  if (!is.null(pos_var)) {
    pos_vals <- tolower(as.character(vdf[[pos_var]]))
    graph <- set_vertex_attr(
      graph, "full_prof_bin",
      value = case_when(
        tolower(pos_vals) %in% c("full professor", "professor", "hoogleraar") ~ 1,
        !is.na(pos_vals) ~ 0,
        TRUE ~ NA_real_
      )
    )
  }

  graph
}
```

### `get_basic_network_stats()`

Compute a standard set of graph-level descriptive statistics for a single wave.

#### Arguments

**`graph`**

> An igraph object.

**`node_data`**

> A tibble of node attributes (currently unused in the computation; retained for interface consistency).

#### Method

The function computes: total node count, count of non-isolate nodes, edge density (undirected, no loops), global clustering coefficient (on the collapsed undirected graph), mean normalized betweenness centrality, and mean path distance. Results are returned as a two-column tibble with `statistic` and `value` columns.

```{r}
get_basic_network_stats <- function(graph, node_data) {
  n_total <- vcount(graph)
  n_non_isolates <- sum(igraph::degree(graph, mode = 'all') > 0)

  tibble(
    statistic = c(
      "N nodes total",
      "N nodes excluding isolates",
      "density",
      "clustering coefficient",
      "average betweenness centrality",
      "mean path distance"
    ),
    value = c(
      n_total,
      n_non_isolates,
      edge_density(graph, loops = FALSE),
      transitivity(as.undirected(graph, mode = "collapse"), type = "globalundirected"),
      mean(igraph::betweenness(graph), na.rm = TRUE),
      mean_distance(graph, directed = is_directed(graph), unconnected = TRUE)
    )
  )
}
```

### `graph_jaccard()`

Compute the edge-level Jaccard similarity index between two network waves.

#### Arguments

**`wave_a`**

> A square numeric adjacency matrix for the first wave.

**`wave_b`**

> A square numeric adjacency matrix for the second wave.

**`directed`**

> Logical. If `TRUE`, all cells are compared; if `FALSE`, only the upper triangle is used. Defaults to `TRUE`.

#### Method

Both matrices are binarized (`> 0`). For directed networks the full vectorized matrices are compared; for undirected networks only the upper triangle is used. The Jaccard index is computed as `intersection / union` over present edges, returning `NA` if the union is empty.

```{r}
graph_jaccard <- function(wave_a, wave_b, directed = TRUE) {
  A1 <- wave_a > 0
  A2 <- wave_b > 0

  if (!directed) {
    A1 <- A1[upper.tri(A1)]
    A2 <- A2[upper.tri(A2)]
  } else {
    A1 <- as.vector(A1)
    A2 <- as.vector(A2)
  }

  inter <- sum(A1 & A2, na.rm = TRUE)
  union <- sum(A1 | A2, na.rm = TRUE)

  if (union == 0) return(NA_real_)
  inter / union
}
```

### `summarize_colnet_wave()`

Produce a full descriptive summary for a single network wave, including basic statistics, Moran's I tests, and optional inter-wave stability.

#### Arguments

**`wave`**

> A square numeric adjacency matrix for the wave to summarize.

**`node_data`**

> A tibble of node attributes aligned to the adjacency matrix.

**`wave_next`**

> Optional adjacency matrix for the subsequent wave. If provided, the Jaccard index between `wave` and `wave_next` is appended to the summary. Defaults to `NULL`.

**`directed`**

> Logical. Whether to treat the network as directed. Defaults to `TRUE`.

**`mode`**

> Character string passed to `graph_morans_i()`. Defaults to `"all"`.

**`i_index_var`**

> Character string. Name of the vertex attribute holding the i-index measure. Defaults to `"i_index"`.

#### Method

The function builds the igraph object via `make_graph_from_wave()`, enriches it with binary indicators via `prepare_graph_attributes()`, and computes basic statistics via `get_basic_network_stats()`. It then runs `graph_morans_i()` at neighbor distances 1 and 2 for gender, discipline, university dummy indicators (from `uni*p` columns), full professor status, and the i-index variable. Moran's I results are relabeled to human-readable statistic names. If `wave_next` is supplied, a Jaccard stability row is appended. The function returns a named list with `graph`, `summary` (basic stats + stability), and `morans_i` (the Moran's I table).

```{r}
summarize_colnet_wave <- function(
    wave,
    node_data,
    wave_next = NULL,
    directed = TRUE,
    mode = "all",
    i_index_var = "i_index"
) {
  graph <- make_graph_from_wave(wave, node_data, directed = directed)
  graph <- prepare_graph_attributes(graph)

  basic <- get_basic_network_stats(graph, node_data)

  moran_results <- list()

  # gender
  if ("gender_bin" %in% vertex_attr_names(graph)) {
    moran_results <- append(moran_results, list(
      graph_morans_i(graph, "gender_bin", 1, mode),
      graph_morans_i(graph, "gender_bin", 2, mode)
    ))
  }

  # discipline
  if ("discipline_bin" %in% vertex_attr_names(graph)) {
    moran_results <- append(moran_results, list(
      graph_morans_i(graph, "discipline_bin", 1, mode),
      graph_morans_i(graph, "discipline_bin", 2, mode)
    ))
  }

  # university dummies from uni22p:uni25p
  uni_vars <- vertex_attr_names(graph) |>
    (\(x) x[grepl("^uni\\d{2}p$", x)])()

  if (length(uni_vars) > 0) {
    moran_results <- append(
      moran_results,
      purrr::map(uni_vars, \(u) {
        vals <- vertex_attr(graph, u)
        levs <- levels(vals)

        # keep only actually observed levels in this wave
        levs <- levs[levs %in% unique(as.character(stats::na.omit(vals)))]

        dummy_results <- purrr::map(levs, \(lev) {
          tmp_name <- paste0(u, "__", lev)

          graph <<- set_vertex_attr(
            graph,
            tmp_name,
            value = dplyr::case_when(
              is.na(vals) ~ NA_real_,
              as.character(vals) == lev ~ 1,
              TRUE ~ 0
            )
          )

          dplyr::bind_rows(
            graph_morans_i(graph, tmp_name, 1, mode),
            graph_morans_i(graph, tmp_name, 2, mode)
          )
        })

        dplyr::bind_rows(dummy_results)
      })
    )
  }

  # full professor
  if ("full_prof_bin" %in% vertex_attr_names(graph)) {
    moran_results <- append(moran_results, list(
      graph_morans_i(graph, "full_prof_bin", 1, mode),
      graph_morans_i(graph, "full_prof_bin", 2, mode)
    ))
  }

  # i-index
  if (i_index_var %in% vertex_attr_names(graph)) {
    moran_results <- append(moran_results, list(
      graph_morans_i(graph, i_index_var, 1, mode),
      graph_morans_i(graph, i_index_var, 2, mode)
    ))
  }

  moran_tbl <- bind_rows(moran_results)

  moran_tbl <- moran_tbl |>
    mutate(
      statistic = case_when(
        variable == "gender_bin" & nb_distance == 1 ~ "gender Moran's I (nb distance 1)",
        variable == "gender_bin" & nb_distance == 2 ~ "gender Moran's I (nb distance 2)",
        variable == "discipline_bin" & nb_distance == 1 ~ "discipline Moran's I (nb distance 1)",
        variable == "discipline_bin" & nb_distance == 2 ~ "discipline Moran's I (nb distance 2)",
        variable == "full_prof_bin" & nb_distance == 1 ~ "full prof Moran's I (nb distance 1)",
        variable == "full_prof_bin" & nb_distance == 2 ~ "full prof Moran's I (nb distance 2)",
        variable == i_index_var & nb_distance == 1 ~ "I index Moran's I (nb distance 1)",
        variable == i_index_var & nb_distance == 2 ~ "I index Moran's I (nb distance 2)",
        grepl("^university__", variable) & nb_distance == 1 ~
          paste0(gsub("^university__", "", variable), " Moran's I (nb distance 1)"),
        grepl("^university__", variable) & nb_distance == 2 ~
          paste0(gsub("^university__", "", variable), " Moran's I (nb distance 2)"),
        TRUE ~ paste(variable, "Moran's I (nb distance", nb_distance, ")")
      )
    ) |>
    select(statistic, observed, expected, sd, p_value)

  stability_tbl <- tibble(
    statistic = character(),
    value = numeric()
  )

  if (!is.null(wave_next)) {
    stability_tbl <- tibble(
      statistic = "Jaccard's Index (period 1 - 2)",
      value = graph_jaccard(wave, wave_next, directed = directed)
    )
  }

  list(
    graph = graph,
    summary = bind_rows(basic, stability_tbl),
    morans_i = moran_tbl
  )
}
```

### `get_nodes_at_distance()`

Return the indices of all vertices at an exact geodesic distance from a focal node.

#### Arguments

**`graph`**

> An igraph object.

**`v`**

> Integer or vertex selector. The focal node from which distances are measured.

**`k`**

> Integer. The exact geodesic distance to select.

**`mode`**

> Character string passed to `igraph::distances()`. Defaults to `"all"`.

#### Method

Computes the full distance vector from `v` to all other vertices and returns the integer indices of vertices whose distance equals exactly `k`.

```{r}
get_nodes_at_distance <- function(graph, v, k, mode = "all") {
  d <- distances(graph, v = v, to = V(graph), mode = mode)
  which(as.numeric(d[1, ]) == k)
}
```

### `safe_mean()`

Compute the mean of a numeric vector after removing `NA` values, returning `NA` if no non-missing values remain.

#### Arguments

**`x`**

> A numeric vector.

#### Method

Removes `NA` values from `x`; returns `NA_real_` if the result is empty, otherwise returns `mean(x)`.

```{r}
safe_mean <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  mean(x)
}
```

### `safe_sum()`

Compute the sum of a numeric vector after removing `NA` values, returning `NA` if no non-missing values remain.

#### Arguments

**`x`**

> A numeric vector.

#### Method

Removes `NA` values from `x`; returns `NA_real_` if the result is empty, otherwise returns `sum(x)`.

```{r}
safe_sum <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  sum(x)
}
```

### `safe_n_distinct()`

Count the number of distinct non-missing values in a vector, returning `NA` if no non-missing values remain.

#### Arguments

**`x`**

> A vector of any type.

#### Method

Removes `NA` values from `x`; returns `NA_real_` if the result is empty, otherwise returns `dplyr::n_distinct(x)`.

```{r}
safe_n_distinct <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  dplyr::n_distinct(x)
}
```

### `ei_index()`

Compute the EI (External–Internal) index of a focal node's neighborhood composition.

#### Arguments

**`ego_value`**

> The category value of the focal node (scalar).

**`alter_values`**

> A vector of category values for the focal node's neighbors.

#### Method

After removing `NA`s from `alter_values`, the function counts internal ties (alters sharing the ego's category) and external ties (alters in a different category). The EI index is computed as `(external - internal) / (external + internal)`, ranging from `–1` (fully homophilous) to `+1` (fully heterophilous). Returns `NA` if `ego_value` is missing or `alter_values` is empty after cleaning.

```{r}
ei_index <- function(ego_value, alter_values) {
  alter_values <- alter_values[!is.na(alter_values)]

  if (is.na(ego_value) || length(alter_values) == 0) {
    return(NA_real_)
  }

  internal <- sum(alter_values == ego_value, na.rm = TRUE)
  external <- sum(alter_values != ego_value, na.rm = TRUE)
  total <- internal + external

  if (total == 0) return(NA_real_)
  (external - internal) / total
}
```

### `neighborhood_density()`

Compute the edge density among a set of vertices (e.g. an ego's neighborhood).

#### Arguments

**`graph`**

> The full igraph object from which the subgraph is induced.

**`vids`**

> Integer vector of vertex indices defining the neighborhood.

#### Method

If fewer than two vertices are supplied, the function returns `NA`. Otherwise it inducts a subgraph on `vids`, collapses it to an undirected graph, and returns `igraph::edge_density()` (without loops).

```{r}
neighborhood_density <- function(graph, vids) {
  if (length(vids) < 2) return(NA_real_)

  subg <- induced_subgraph(graph, vids = vids)

  # for co-author neighborhoods, density is usually easier to interpret undirected
  subg <- as.undirected(subg, mode = "collapse")

  edge_density(subg, loops = FALSE)
}
```

### `compute_ego_objects()`

Compute ego-level network statistics for every scholar in a single wave.

#### Arguments

**`wave`**

> A square numeric adjacency matrix for the wave.

**`node_data`**

> A tibble of node attributes aligned to the adjacency matrix.

**`directed`**

> Logical. Whether to treat the network as directed. Defaults to `TRUE`.

**`mode`**

> Character string controlling directionality when computing neighbor sets. Defaults to `"all"`.

**`gender_var`**, **`discipline_var`**, **`university_var`**, **`position_var`**, **`i_index_var`**

> Character strings naming the vertex attribute columns to use for composition and capital measures. `university_var` is required and validated.

#### Method

The function builds an igraph object from `wave` and attaches `node_data` as vertex attributes. For each scholar it finds neighbors at distances 1 and 2 using `get_nodes_at_distance()`, then computes: neighborhood size, neighborhood density, local clustering coefficient, betweenness centrality, mean path length, gender/discipline/university EI indices, reachable full-professor counts, and mean/total i-index at each distance. Results are returned as a tibble with one row per scholar.

```{r}
compute_ego_objects <- function(
    wave,
    node_data,
    directed = TRUE,
    mode = "all",
    gender_var = "gender",
    discipline_var = "discipline_22",
    university_var,
    position_var = "position_22",
    i_index_var = "i10_index_21"
) {
  stopifnot(university_var %in% names(node_data))

  graph <- make_graph_from_wave(wave, node_data, directed = directed)

  n <- vcount(graph)

  # graph-level / ego-level structure measures
  bet <- igraph::betweenness(graph)
  mnd <- mean_distance(graph, directed = is_directed(graph), unconnected = TRUE)
  loc_clust <- transitivity(
    as.undirected(graph, mode = "collapse"),
    type = "localundirected",
    isolates = "zero"
  )

  # extract node attributes once
  gender <- vertex_attr(graph, gender_var)
  discipline <- vertex_attr(graph, discipline_var)
  university <- vertex_attr(graph, university_var)
  position <- vertex_attr(graph, position_var)
  i_index <- vertex_attr(graph, i_index_var)

  full_prof <- case_when(
    tolower(as.character(position)) %in% c("full professor", "professor", "hoogleraar") ~ 1,
    is.na(position) ~ NA_real_,
    TRUE ~ 0
  )

  out <- purrr::map_dfr(seq_len(n), function(i) {
    d1 <- get_nodes_at_distance(graph, i, 1, mode = mode)
    d2 <- get_nodes_at_distance(graph, i, 2, mode = mode)

    tibble(
      node_id = i,
      uid = if ("uid" %in% names(node_data)) node_data$uid[i] else i,

      #------------------------------------------
      # size
      #------------------------------------------
      `N co-authors distance 1` = length(d1),
      `N co-authors distance 2` = length(d2),

      #------------------------------------------
      # structure
      #------------------------------------------
      `density distance 1` = neighborhood_density(graph, d1),
      `density distance 2` = neighborhood_density(graph, d2),
      `clustering coefficient` = loc_clust[i],
      `betweenness centrality` = bet[i],
      `avg. path length` = mnd,

      #------------------------------------------
      # composition
      #------------------------------------------
      `gender EI index distance 1` = ei_index(gender[i], gender[d1]),
      `gender EI index distance 2` = ei_index(gender[i], gender[d2]),

      `discipline EI index distance 1` = ei_index(discipline[i], discipline[d1]),
      `discipline EI index distance 2` = ei_index(discipline[i], discipline[d2]),

      `university EI index distance 1` = ei_index(university[i], university[d1]),
      `university EI index distance 2` = ei_index(university[i], university[d2]),

      #------------------------------------------
      # scientific capital
      #------------------------------------------
      `reachable full prof distance 1` = safe_sum(full_prof[d1]),
      `reachable full prof distance 2` = safe_sum(full_prof[d2]),

      `average i-index distance 1` = safe_mean(i_index[d1]),
      `average i-index distance 2` = safe_mean(i_index[d2]),

      `total i-index distance 1` = safe_sum(i_index[d1]),
      `total i-index distance 2` = safe_sum(i_index[d2]),
    )
  })

  out
}
```

### `calculate_ego_results()`

Compute ego-level network statistics for a wave and summarize them by gender.

#### Arguments

**`wave1`**

> A square numeric adjacency matrix for the wave to analyze.

**`node_data`**

> A tibble of node attributes aligned to the adjacency matrix, including `uid` and `gender`.

**`mode`**

> Character string passed to `compute_ego_objects()`. Defaults to `"all"`.

#### Method

The function calls `compute_ego_objects()` with fixed variable names (`gender`, `uni22p`, `i10_index_21`), then joins the result to `node_data`, groups by `gender`, takes column means across all numeric columns, transposes the result, and returns a tibble with one row per statistic and one column per gender category.

```{r}
calculate_ego_results <- function(wave1, node_data, mode = 'all'){
  ego_w1 <- compute_ego_objects(
    wave = wave1,
    node_data = node_data,
    directed = TRUE,
    mode = mode,
    gender_var = "gender",
    university_var = "uni22p",
    i_index_var = "i10_index_21"
  )
  out <- node_data |>
    select(uid, gender) |>
    filter(!is.na(gender)) |>
    left_join(ego_w1) |>
    select(-node_id, -uid) |>
    group_by(gender) |>
    summarise_if(is.numeric, mean, na.rm=T) |>
    column_to_rownames(var = 'gender') |> t()

  names = rownames(out)
  out = out |> as_tibble() |> mutate(name = names, .before = 1)

  return(out)
}
```

### `exclude_junior_staff()`

Subset a colnet object to senior scholars only, removing PhD candidates, postdoctoral researchers, and staff.

#### Arguments

**`colnet`**

> A colnet list with `data` (demographics tibble with a `position_22` column) and `nets` (list of adjacency matrices).

#### Method

The function filters `colnet[['data']]` to exclude scholars whose `position_22` is `"PhD Candidate"`, `"Staff"`, or `"Postdoctoral Researcher"`. It then subsets each adjacency matrix in `colnet[['nets']]` to the remaining UIDs and returns the pruned list.

```{r}
exclude_junior_staff <- function(colnet) {
  out = list()
  out[['data']] <-  colnet[['data']] |>
    filter(!position_22 %in% c(
      'PhD Candidate', "Staff", "Postdoctoral Researcher"
    ))
  uids = out[['data']] |> pull(uid) |> unique()

  for (i in 1:length(colnet[['nets']])) {
    net = colnet[['nets']][[i]][uids, uids]
    out[['nets']][[i]] = net
  }
  return(out)
}
```

### `select_junior_staff()`

Subset a colnet object to junior scholars only, retaining only PhD candidates and postdoctoral researchers.

#### Arguments

**`colnet`**

> A colnet list with `data` (demographics tibble with a `position_22` column) and `nets` (list of adjacency matrices).

#### Method

The function filters `colnet[['data']]` to scholars whose `position_22` is `"PhD Candidate"` or `"Postdoctoral Researcher"`. It then subsets each adjacency matrix in `colnet[['nets']]` to the remaining UIDs and returns the pruned list.

```{r}
select_junior_staff <- function(colnet) {
  out = list()
  out[['data']] <-  colnet[['data']] |>
    filter(position_22 %in% c(
      'PhD Candidate', "Postdoctoral Researcher"
    ))
  uids = out[['data']] |> pull(uid) |> unique()

  for (i in 1:length(colnet[['nets']])) {
    net = colnet[['nets']][[i]][uids, uids]
    out[['nets']][[i]] = net
  }
  return(out)
}
```

### `mask_structural_missing()`

Set structurally missing observations to a sentinel value in all network matrices of a colnet object.

#### Arguments

**`colnet`**

> A colnet list as produced by `fcolnet2()`, with `nets` (a nested list of adjacency matrices by type and wave) and `data` (a demographics tibble with `uid` and `uni*` affiliation columns).

**`include_ties`**

> Logical. If `TRUE`, scholars with zero ties are also masked regardless of affiliation; if `FALSE` (the default), only scholars who have both zero ties and a missing affiliation for that wave are masked.

**`value`**

> Numeric. The sentinel value assigned to masked rows and columns. Defaults to `10`.

#### Method

For each combination of network type and wave index, the function identifies scholars who have zero row sums in the network matrix (optionally combined with `include_ties`) and for whom the corresponding wave-year university affiliation (`uni22`, `uni24`, or `uni25`) is `NA`. Rows and columns corresponding to these structurally absent scholars are set to `value` in the matrix, and the modified matrix is stored back in `colnet`.

```{r}
mask_structural_missing <- function(colnet, include_ties = FALSE, value = 10){
  for (type in names(colnet[['nets']])){
    for (i in 1:length(colnet[['nets']][[type]])){
      net <- colnet[['nets']][[type]][[i]]

      no_ties <- rowSums(net) == 0 * include_ties
      abscent <- colnet[['data']] |>
        column_to_rownames('uid') |>
        pull(paste0("uni", c(22, 24, 25)[i])) |>
        is.na()

      exclude <- no_ties & abscent
      net[exclude, ] <- value
      net[ ,exclude] <- value

      colnet[['nets']][[type]][[i]] <- net
    }
  }

  return(colnet)
}
```

### `build_results()`

Orchestrate building a colnet and computing wave summaries and ego statistics for all and senior-only samples.

#### Arguments

**`type`**

> Character string. Authorship type passed to `fcolnet2()` (e.g. `"first"`, `"last"`, `"all"`).

**`mode`**

> Character string. Network mode passed to `summarize_colnet_wave()` and `calculate_ego_results()` (e.g. `"out"`, `"all"`).

#### Method

The function calls `fcolnet2()` with the globally defined `uni_levels`, applies `harmonize_covariates()` and `mask_structural_missing()`. It extracts the first two waves and node data, then defines an internal helper `make_res()` that calls `summarize_colnet_wave()` and `calculate_ego_results()` for a given wave–node data combination. It computes results for both the full sample and the senior-only subsample (via `exclude_junior_staff()`). Returns a named list with elements `all` and `sen`.

```{r}
build_results <- function(type, mode) {
  colnet <- fcolnet2(data,
      university = toupper(uni_levels),
      type = type,
      waves = NULL
    ) |>
    harmonize_covariates(what = 'data') |>
    mask_structural_missing(value = NA)

  wave1 = colnet[['nets']][[1]]
  wave2 = colnet[['nets']][[2]]
  node_data = colnet$data

  make_res <- function(w1, w2, nd, mode) {
    res <- summarize_colnet_wave(
      wave = w1, node_data = nd, wave_next = w2,
      directed = TRUE, mode = mode, i_index_var = "i10_index_21"
    )
    res$ego <- calculate_ego_results(w1, nd, mode = mode)
    res
  }

  colnet2 <- exclude_junior_staff(colnet)
  wave12 = colnet2[['nets']][[1]]
  wave22 = colnet2[['nets']][[2]]
  node_data2 = colnet2$data

  list(
    all = make_res(wave1,  wave2,  node_data,  mode),
    sen = make_res(wave12, wave22, node_data2, mode)
  )
}
```

## Application

```{r}
data <- freadRDS2('data', location = "./data/analysis/")
topics <-freadRDS2('topics', location = './data/analysis/')
```

```{r}
uni_levels = tolower(c("EUR", "RU", "UL", "UU", "UVA", "UVG", "UVT", "VU"))

res_first <- build_results('first', 'out')
res_last  <- build_results('last',  'out')
res_all   <- build_results('all',   'all')

results <- list(
  first_all = res_first$all, first_sen = res_first$sen,
  last_all  = res_last$all,  last_sen  = res_last$sen,
  all_all   = res_all$all,   all_sen   = res_all$sen
)
```

```{r}
# Combine ego results
ego_res <- names(results) |>
  purrr::imap(function(name, i) {
    results[[i]][["ego"]] |>
      rename_with(~ paste0(name, "_", .), c(female, male))
  }) |>
  purrr::reduce(left_join, by = "name")

# Combine network summary results
net_res <- names(results) |>
  purrr::imap(function(name, i) {
    results[[i]][["summary"]] |>
      rename(observed = value) |>
      bind_rows(results[[i]][['morans_i']]) |>
      select(statistic, observed) |>
      rename_with(~ paste0(name, "_", .), c(observed))
  }) |>
  purrr::reduce(left_join, by = "statistic") |>
  filter(!str_detect(statistic, 'uni25'), !str_detect(statistic, 'uni24'))
```

```{r}
ego_res |> writexl::write_xlsx(file.path('results', 'tables', 'ego_stats.xlsx'))

net_res |> writexl::write_xlsx(file.path('results', 'tables', 'net_stats.xlsx'))
```
