---
title: Main Analysis
date: today

format:
  html:
    code-links:
      - href: 0a.custom_functions.html
        icon: file-code
        text: Custom Functions

execute:
  eval: false
---

## Getting Started

```{r}
#| warning: false
#| output: false
#| echo: false
BASEDIR = stringr::str_replace_all(getwd(), '/results', '')
setwd(BASEDIR)
```

```{r}
#| warning: false
#| output: false
# load custom functions
source("src/utils/custom_functions.r")

# clear the global environment and set dependencies
.clear_global_environment()
.load_quarto_dependencies()

BASEDIR = stringr::str_replace_all(getwd(), '/results', '')
setwd(BASEDIR)
```

```{r}
#| warning: false
#| output: false

# load and activate packages
library(tidyverse)
library(RSiena)
library(igraph)
library(dplyr)
library(tibble)
library(glue)
```

## Functions

### `make_adjacency_matrix()`

Build a co-authorship adjacency matrix for a specified set of scholars and time window.

#### Arguments

**`data`**

> Named list with a `works` element (named list of per-scholar works tibbles containing `uid`, `work_id`, `authorships`, `publication_date`, and `publication_year`).

**`uids`**

> Character vector of focal scholar UIDs defining the rows and columns of the output matrix.

**`type`**

> Character string controlling which author position is treated as the focal node. `"first"` and `"last"` use the author at that position; `"all"` creates a full crossing of all co-authors on each work. Defaults to `"first"`.

**`min_year`**

> Integer year or Date giving the lower bound of the publication window (inclusive).

**`max_year`**

> Integer year or Date giving the upper bound of the publication window (inclusive).

**`weighted`**

> Logical. If `TRUE`, matrix values reflect the count of shared publications; if `FALSE`, values are binarized to `1L`. Defaults to `FALSE`.

#### Method

The function binds the works for the requested `uids`, filters to the specified publication window (using `publication_date` for Date inputs and `publication_year` for integers), selects `uid`, `work_id`, and `authorships`, deduplicates by `work_id`, and unnests authorships. Depending on `type`, it either builds ego–alter pairs using the author at the named position or crosses all author pairs per work. Co-author pairs are counted and optionally binarized. The result is pivoted to a wide adjacency matrix, padded to the full `uids` dimension with zeros for absent pairs, and the diagonal is set to `0`.

```{r}
make_adjacency_matrix <- function(
    data, uids, type = 'first', min_year = 2020, max_year = 2026, weighted = FALSE
  ) {

  # make edgelist from works
  edges <- data[['works']][uids] |>
    purrr::keep(~ nrow(.x) > 0) |>
    bind_rows()

  if (inherits(min_year, "Date")) {
    edges <- edges |> filter(publication_date >= min_year, publication_date <= max_year)
  } else {
    edges <- edges |> filter(publication_year >= min_year, publication_year <= max_year)
  }
  edges <- edges |>
    select(uid, work_id, authorships) |>
    rename(e_uid = uid) |>
    distinct(work_id, .keep_all = TRUE) |>
    unnest(authorships)

  if (type != 'all') {
    edges <- edges |>
      group_by(work_id) |>
      mutate(
        e_uid = uid[author_position == type][1]
      ) |> ungroup() |>
      filter(!is.na(e_uid), e_uid != uid) |>
      select(work_id, e_uid, uid)
  } else {
    edges <- edges |>
      group_by(work_id) |>
      reframe(
        tidyr::crossing(
          e_uid = uid,
          uid   = uid
        )
      ) |>
      ungroup() |>
      filter(!is.na(e_uid), e_uid != uid) |>
      select(work_id, e_uid, uid)
  }

  # calculate the number of edges between ego and alters.
  edge_counts <- edges |> count(e_uid, uid, name="w")

  # convert counts to incidences
  if (!weighted) {
    edge_counts <- edge_counts %>% mutate(w = 1L)
  }

  # all nodes appearing as ego or alter
  nodes <- sort(unique(c(edge_counts$e_uid, edge_counts$uid)))

  # complete matrix grid and pivot to wide
  adj <- edge_counts %>%
    rename(from = e_uid, to = uid) %>%
    complete(from = nodes, to = nodes, fill = list(w = 0)) %>%
    pivot_wider(names_from = to, values_from = w) %>%
    arrange(from)

  # convert to matrix and set rownames
  mat <- adj %>% as.data.frame()

  row_ids <- mat$from
  mat$from <- NULL

  mat <- as.matrix(mat)
  rownames(mat) <- row_ids
  diag(mat) <- 0

  # make sure the matrices are consistent across waves
  all_uids <- as.character(uids)

  # indices of existing rows/cols in desired order
  ri <- match(all_uids, rownames(mat))
  ci <- match(all_uids, colnames(mat))

  # start with full NA matrix
  mat_full <- matrix(0L, nrow = length(all_uids), ncol = length(all_uids),
                    dimnames = list(all_uids, all_uids))

  # copy overlap block
  mat_full[!is.na(ri), !is.na(ci)] <- mat[ri[!is.na(ri)], ci[!is.na(ci)], drop = FALSE]
  mat <- mat_full

  mat
}
```

### `fcolnet2()`

Build a multi-wave co-authorship network filtered by university, position, and discipline.

#### Arguments

**`data`**

> Named list with a `demographics` element (containing `uid`, `university_*`, `discipline_*`, and `position_*` columns) and a `works` element used by `make_adjacency_matrix()`.

**`university`**

> Character vector of university abbreviations to include. Defaults to all eight valid Dutch universities.

**`position`**

> Character vector of academic positions to include. Defaults to all five valid positions.

**`discipline`**

> Character vector of disciplines to include. Defaults to `"Sociology"` and `"Political Sciences"`.

**`type`**

> Character string or `NULL`. If `NULL`, adjacency matrices are built for `"first"`, `"last"`, and `"all"` authorship types; otherwise the single specified type is used.

**`waves`**

> List of two-element vectors specifying the start and end of each observation wave. Defaults to three waves roughly corresponding to 2019–2022, 2023–2024, and 2025–2026.

#### Method

The function validates each filter argument against its allowed values using an internal helper. It filters `demographics` to scholars who appear with at least one of the specified universities, disciplines, and positions across the three measurement years. For each wave (and optionally each authorship type), it calls `make_adjacency_matrix()` with the corresponding year bounds. The function returns a list with `data` (the filtered demographics tibble) and `nets` (the nested list of adjacency matrices).

```{r}
fcolnet2 <- function(
    data,
    university = NULL,
    position   = NULL,
    discipline = NULL,
    type = NULL,
    waves = list(c(2019, 2022), c(2023, 2024), c(2025, 2026))
  ){

  valid_disciplines  <- c("Sociology", "Political Sciences")
  valid_universities <- c("UVA","EUR","UL","VU","UVG","UU","RU","UVT")
  valid_positions    <- c("Associate Professor","PhD Candidate","Full Professor",
                          "Researcher or Lecturer","Assistant Professor")
  valid_dates        <- list(c(ymd('20201219'), ymd('20221219')),
                             c(ymd('20221220'), ymd('20240419')),
                             c(ymd('20240420'), ymd('20260201')))

  # helper: set default + validate
  set_and_check <- function(x, valid, name = deparse(substitute(x))) {
    x <- if (is.null(x)) valid else x
    bad <- setdiff(unique(na.omit(x)), valid)
    if (length(bad)) stop("Invalid ", name, ": ", paste(bad, collapse = ", "), call. = FALSE)
    x
  }

  # set and check values
  discipline <- set_and_check(discipline, valid_disciplines,  "discipline")
  university <- set_and_check(university, valid_universities, "university")
  position   <- set_and_check(position,   valid_positions,    "position")
  waves      <- if(is.null(waves)) valid_dates else waves
  min_year   <- min(unlist(waves))
  max_year   <- max(unlist(waves))

  # step 1: make selection of nodes
  authors <- data[["demographics"]] |>
    filter(
      if_any(c(university_22, university_24, university_25),
            ~ .x |>
              toupper() |>
              str_detect(paste(university, collapse = "|"))),
      if_any(c(discipline_22, discipline_24, discipline_25),
            ~ .x %in% discipline),
      if_any(c(position_22, position_24, position_25),
            ~ .x %in% position)
    ) |>
    arrange(uid)
  uids <- authors |> pull(uid) |> sort()

  # step 3: create empty matrixes (wave, i, j)
  nwaves <- length(waves)
  nets <- list()
  if (is.null(type)){
    types = c('first', 'last', 'all')
    for (t in types){
      for (w in 1:length(waves)){
        nets[[t]][[w]] <- make_adjacency_matrix(
          data, uids, type = t, min_year = waves[[w]][1], max_year = waves[[w]][2])
      }
    }
  } else {
    for (w in 1:length(waves)){
      nets[[w]] <- make_adjacency_matrix(
        data, uids, type, min_year = waves[[w]][1], max_year = waves[[w]][2])
    }
  }

  # step 4: fill nets
  output <- list(
    data = authors,
    nets = nets
  )

  return(output)
}
```

### `harmonize_covariates()`

Standardize and rename covariate columns in a colnet data sub-list for use in RSiena models.

#### Arguments

**`data`**

> A colnet list whose `data` (or other named) sub-element contains demographics columns for `gender`, `discipline_*`, and `university_*_first`.

**`what`**

> Character string naming the sub-element of `data` to transform. Defaults to `'data'`.

#### Method

The function adds a `female` integer indicator (1 if `gender == "female"`, 0 otherwise) and a `soc` integer indicator (1 if the first non-missing discipline across the three measurement years is `"Sociology"`). It renames `university_*_first` columns to `uni*p` (e.g. `university_22_first` → `uni22p`) and converts all `uni*p` columns to ordered factors using a predefined set of eight university abbreviations as levels.

```{r}
harmonize_covariates <- function(data, what = 'data'){
  uni_levels = c("EUR", "RU" , "UL", "UU", "UVA", "UVG", "UVT", "VU")

  data[[what]] = data[[what]] |>
    mutate(
      female = as.integer(gender == 'female'),
      soc = pmap_chr(
        list(discipline_22, discipline_24, discipline_25),
        ~ unique(na.omit(c(...)))[1]
      ),
      soc = as.integer(soc == 'Sociology')
    ) |>
    rename_with(
      ~ .x |>
        str_replace('university_', 'uni') |>
        str_replace('_first$', 'p')
    ) |>
    mutate(
      across(
        starts_with('uni') & ends_with('p'),
        ~ factor(tolower(.x),
        levels = tolower(uni_levels),
        ordered = TRUE)
      )
    )

  return(data)
}
```

### `mask_structural_missing()`

Set structurally missing observations to a sentinel value in all network matrices of a colnet object.

#### Arguments

**`colnet`**

> A colnet list as produced by `fcolnet2()`, with `nets` (a nested list of adjacency matrices by type and wave) and `data` (a demographics tibble with `uid` and `uni*` affiliation columns).

**`include_ties`**

> Logical. If `TRUE`, scholars with zero ties are also masked regardless of affiliation; if `FALSE` (the default), only scholars who have both zero ties and a missing affiliation for that wave are masked.

**`value`**

> Numeric. The sentinel value assigned to masked rows and columns. Defaults to `10`.

#### Method

For each combination of network type and wave index, the function identifies scholars who have zero row sums in the network matrix (optionally combined with `include_ties`) and for whom the corresponding wave-year university affiliation (`uni22`, `uni24`, or `uni25`) is `NA`. Rows and columns corresponding to these structurally absent scholars are set to `value` in the matrix, and the modified matrix is stored back in `colnet`.

```{r}
mask_structural_missing <- function(colnet, include_ties = FALSE, value = 10){
  for (type in names(colnet[['nets']])){
    for (i in 1:length(colnet[['nets']][[type]])){
      net <- colnet[['nets']][[type]][[i]]

      no_ties <- rowSums(net) == 0 * include_ties
      abscent <- colnet[['data']] |>
        column_to_rownames('uid') |>
        pull(paste0("uni", c(22, 24, 25)[i])) |>
        is.na()

      exclude <- no_ties & abscent
      net[exclude, ] <- value
      net[ ,exclude] <- value

      colnet[['nets']][[type]][[i]] <- net
    }
  }

  return(colnet)
}
```

### `add_interdisciplinarity()`

Align the subfield interdisciplinarity matrix to the scholars in a colnet object.

#### Arguments

**`colnet`**

> A colnet list containing a `data` element with a `uid` column.

**`topics`**

> Named list with an `intdis` sub-list whose `subfield` element is a matrix with rows indexed by UID and columns indexed by year.

#### Method

The function extracts the sorted UIDs from `colnet$data`, initializes an output matrix with those UIDs as rows and the same column names as `topics[["intdis"]][["subfield"]]`, and fills the rows present in both the output and the source matrix. The result is stored in `colnet[["intdis"]]` and the updated list is returned.

```{r}
add_interdisciplinarity = function(colnet, topics){
  uids = colnet$data |> pull(uid) |> unique() |> sort()
  intdis <- topics[["intdis"]][["subfield"]]

  colnet[["intdis"]] <- {
    out <- matrix(
      NA_real_,
      nrow = length(uids),
      ncol = ncol(intdis),
      dimnames = list(uids, colnames(intdis))
    )

    common_uids <- intersect(uids, rownames(intdis))
    out[common_uids, ] <- intdis[common_uids, , drop = FALSE]
    out
  }

  return(colnet)
}
```

### `add_dissimilarity()`

Align per-year subfield dissimilarity matrices to the scholars in a colnet object.

#### Arguments

**`colnet`**

> A colnet list containing a `data` element with a `uid` column.

**`topics`**

> Named list with a `dissim` sub-list whose `subfield` element is a named list of square UID × UID dissimilarity matrices (one per year).

#### Method

For each year's dissimilarity matrix in `topics[["dissim"]][["subfield"]]`, the function finds the UIDs common to `colnet$data`, the matrix rows, and the matrix columns, initializes a new `n × n` NA matrix (where `n = length(uids)`), and fills the overlapping block. The resulting named list of aligned matrices is stored in `colnet[["dissim"]]` and the updated list is returned.

```{r}
add_dissimilarity <- function(colnet, topics) {
  uids = colnet$data |> pull(uid) |> unique() |> sort()

  colnet[["dissim"]] <- purrr::map(
    topics[["dissim"]][["subfield"]],
    \(mat) {
      common_uids <- Reduce(intersect, list(uids, rownames(mat), colnames(mat)))

      out <- matrix(
        NA_real_,
        nrow = length(uids),
        ncol = length(uids),
        dimnames = list(uids, uids)
      )

      out[common_uids, common_uids] <- mat[common_uids, common_uids, drop = FALSE]
      out
    }
  )

  return(colnet)
}
```

### `add_distances()`

Align per-year institutional travel-time distance matrices to the scholars in a colnet object.

#### Arguments

**`colnet`**

> A colnet list containing a `data` element with a `uid` column.

**`data`**

> Named list containing a `distances` element (a named list of per-year UID × UID travel-time distance matrices, as produced in `09.operationalizations.qmd`).

#### Method

For each year's distance matrix in `data[["distances"]]`, the function finds the UIDs common to `colnet$data`, the matrix rows, and the matrix columns, initializes a new `n × n` NA matrix, and fills the overlapping block. The resulting named list of aligned matrices is stored in `colnet[["distances"]]` and the updated list is returned.

```{r}
add_distances <- function(colnet, data) {
  uids = colnet$data |> pull(uid) |> unique() |> sort()

  colnet[["distances"]] <- purrr::map(
    data[["distances"]],
    \(mat) {
      common_uids <- Reduce(intersect, list(uids, rownames(mat), colnames(mat)))

      out <- matrix(
        NA_real_,
        nrow = length(uids),
        ncol = length(uids),
        dimnames = list(uids, uids)
      )

      out[common_uids, common_uids] <- mat[common_uids, common_uids, drop = FALSE]
      out
    }
  )

  return(colnet)
}
```

### `includeDefaultEffects()`

Add a standard set of structural and covariate RSiena effects to an effects object for both network outcomes.

#### Arguments

**`myeff`**

> An RSiena effects object as returned by `getEffects()`.

#### Method

The function includes `density`, `recip`, and `inPop` structural effects for both the `first` and `last` network outcomes. It then loops over all university covariates in `uni` and adds `egoX` effects for each to both outcomes, using `glue()` to construct the interaction term names. The updated effects object is returned.

```{r}
includeDefaultEffects <- function(myeff){
  myeff <- includeEffects(myeff, density, recip, inPop) #name='first'
  myeff <- includeEffects(myeff, density, recip, inPop, name='last')
  # myeff <- includeEffects(myeff, antiInIso) # does not work in later models

  # # add node characteristics
  # myeff <- includeEffects(myeff, egoX, interaction1 = "soc")
  # myeff <- includeEffects(myeff, egoX, interaction1 = "soc", name='last')
  # # sameX soc collinear with university (uu has no pol-sci)
  for (name in names(uni)){
    myeff <- includeEffects(myeff, egoX, interaction1 = glue("uni${name}"))
    myeff <- includeEffects(myeff, egoX, interaction1 = glue("uni${name}"), name='last')
    # myeff <- includeEffects(myeff, simX, interaction1 = glue("uni${name}"))
  }

  return(myeff)
}
```

### `tidy_rsiena()`

Extract and format RSiena estimation results as a tidy tibble.

#### Arguments

**`ans`**

> An RSiena results object as returned by `siena07()`.

#### Method

The function builds a tibble from the included effects in `ans$effects`, joining in `theta` (estimates), `se` (standard errors), and `tconv` (per-effect convergence t-ratios). It appends a row for the overall maximum convergence ratio (`tconv.max`). For each included effect it computes a z-statistic, two-tailed p-value, significance stars (`+`, `*`, `**`, `***`), and formatted strings for the estimate (with stars) and standard error. The returned tibble contains columns for dependent variable, effect name, short name, interaction terms, estimate, SE, z, p-value, stars, convergence t-ratio, and the two formatted strings.

```{r}
tidy_rsiena <- function(ans) {
  # extract effects
  effects = tibble(
      depvar = ans$effects$name,
      effect = ans$effects$effectName,
      short  = ans$effects$shortName,
      interaction1 = ans$effects$interaction1,
      interaction2 = ans$effects$interaction2,
      include = ans$effects$include,
      estimate = ans$theta,
      se = ans$se,
      conv_t = ans$tconv
    ) |>
    filter(include)

  #extract max conv rate
  tconv.max = tibble(
    depvar = unique(ans$effects$name),
    effect = 'Overall maximum convergence ratio',
    short = 'tconv.max',
    estimate = ans$tconv.max[[1]]
  )

  # combine and format
  effects |>
    bind_rows(tconv.max) |>
    mutate(
      z = estimate / se,
      z = ifelse(short %in% c('rate', 'tconv.max'), NA, z),
      p_value = 2 * pnorm(-abs(z)),
      stars = case_when(
        p_value < 0.001 ~ "***",
        p_value < 0.01  ~ "**",
        p_value < 0.05  ~ "*",
        p_value < 0.1   ~ "+",
        TRUE            ~ ""
      ),
      festimate = sprintf("%.3f%s", estimate, stars),
      fse = sprintf("(%.3f)", se),
      fse = ifelse(short == 'tconv.max', '', fse),
      z = round(z, 3),
      p_value = round(p_value, 4),
      conv_t = round(conv_t, 3)
    )  |>
    select(
      depvar, effect, short, interaction1, interaction2,
      estimate, se, z, p_value, stars, conv_t, festimate, fse
    )
}
```

## Application

```{r}
data <- freadRDS2('data', location = "./data/analysis/")
topics <-freadRDS2('topics', location = './data/analysis/')
```

```{r}
# uni_levels = tolower(c("EUR", "RU" , "UL", "UU", "UVA", "UVG", "UVT", "VU"))
uni_levels = tolower(c(
  "RU"
  # , "UU"
  # , "UVA", "UVG"
))

colnet <- fcolnet2(data,
    discipline = "Sociology",
    university = toupper(uni_levels),
    type = NULL,
    waves = NULL
  ) |>
  harmonize_covariates(what='data') |>
  mask_structural_missing()

wave1 = colnet[['nets']][['first']][[1]]
wave2 = colnet[['nets']][['first']][[2]]
wave3 = colnet[['nets']][['first']][[3]]

wave1l = colnet[['nets']][['last']][[1]]
wave2l = colnet[['nets']][['last']][[2]]
wave3l = colnet[['nets']][['last']][[3]]
```

```{r}
# set dependent variable
net_array <- array(
    data = c(wave1, wave2, wave3),
    dim = c(dim(wave1), 3)
)
first <- sienaDependent(net_array, )

net_array <- array(
    data = c(wave1l, wave2l, wave3l),
    dim = c(dim(wave1l), 3)
)
last <- sienaDependent(net_array, )

# set constant variables as numeric covariances

gender <- coCovar(colnet[['data']]$female)
soc <- coCovar(colnet[['data']]$soc)

# set values for time varying universities
uni = colnet[['data']] |>
  select(uni22p, uni24p, uni25p)

uni <- uni_levels[-1] |>
  set_names() |>
  toupper() |>
  map(function(level) {
    mat <- uni |>
      transmute(
        u22 = as.integer(str_detect(uni22p, level)),
        u24 = as.integer(str_detect(uni24p, level)),
        u25 = as.integer(str_detect(uni25p, level)) # excluded for model specification
      ) |>
      as.matrix()

    varCovar(mat[, c(1, 2)], centered = FALSE)
  })

# functions
funcs = colnet[['data']] |> select(position_22, position_24, position_25)
funcs_levels = c(
  "PhD Candidate", "Researcher or Lecturer",
  "Assistant Professor", "Associate Professor",
  "Full Professor"
)

funcs = funcs_levels[-1] |>
  set_names() |>
  map(function(level){
    mat <- funcs |>
      transmute(
        f22 = as.integer(position_22 == level),
        f24 = as.integer(position_25 == level),
        f25 = as.integer(position_25 == level) # excluded for model specification
      ) |>
      as.matrix()

    varCovar(mat[, c(1, 2)], centered = FALSE)
  })
```

```{r}
# set migrant constant covars
migrant_groups = c('NL', 'M4', 'other')
mig <- migrant_groups[-1] |>
  set_names() |>
  map(function(group){
    val <- as.integer(colnet[['data']]$migrant == group)

    coCovar(val)
  })
```

```{r}
colnet = colnet |>
  add_interdisciplinarity(topics) |>
  add_dissimilarity(topics) |>
  add_distances(data)

intdis = colnet[['intdis']][,1:2] |> varCovar(centered = TRUE)
dissim = colnet[['dissim']][c('2022', '2024')] |> simplify2array() |> varDyadCovar(centered = TRUE)
distan = colnet[['distances']][c('2022', '2024')] |> simplify2array() |> varDyadCovar(centered = TRUE)
```

```{r}
# combine variables into Rsiena data object
mydata <- sienaDataCreate(
  first, last, gender, soc,
  mig$M4, mig$other,
  # uni$ru, uni$ul, uni$uu, uni$uva, uni$uvg, uni$uvt, uni$vu,
  # uni$uu,
  # uni$uva, uni$uvg,
  funcs[[1]], funcs[[2]], funcs[[3]], funcs[[4]],
  intdis, dissim, distan
)

setwd(BASEDIR)
if(!dir.exists("results")) dir.create("results")
setwd(paste0(BASEDIR, '/results'))
print01Report(mydata, modelname = "./init")

# model instability: perhaps throwing cases withh full missingness out
```

```{r}
myAlgorithm <- sienaAlgorithmCreate(
  projname = "init",
  # n3 = 3000 # for publication ready results
)
```

```{r}
myeff <- getEffects(mydata)
myeff <- includeEffects(myeff, density, recip, inPopSqrt, outActSqrt,gwespFF, name='first')
myeff <- includeEffects(myeff, inIso, name='first')
myeff <- includeEffects(myeff, egoX, interaction1="gender", name='first')
myeff <- includeEffects(myeff, altX, interaction1="gender", name='first')
myeff <- includeEffects(myeff, sameX, interaction1="gender", name='first')
myeff <- includeEffects(myeff, egoX, interaction1="funcs[[1]]", name='first')
myeff <- includeEffects(myeff, egoX, interaction1="funcs[[2]]", name='first')
myeff <- includeEffects(myeff, egoX, interaction1="funcs[[3]]", name='first')
myeff <- includeEffects(myeff, egoX, interaction1="funcs[[4]]", name='first')
myeff <- includeEffects(myeff, altX, interaction1="funcs[[1]]", name='first')
myeff <- includeEffects(myeff, altX, interaction1="funcs[[2]]", name='first')
myeff <- includeEffects(myeff, altX, interaction1="funcs[[3]]", name='first')
myeff <- includeEffects(myeff, altX, interaction1="funcs[[4]]", name='first')
myeff <- includeEffects(myeff, X, interaction1="dissim", name='first')
myeff <- includeInteraction(myeff, inPop, altX, interaction1=c("", "gender"))
myeff <- includeInteraction(myeff, altX, altX, interaction1=c("funcs[[4]]", "gender"))

# estimate model
ansm1 <- siena07(
  myAlgorithm,
  data = mydata,
  effects = myeff,
  returnDeps = TRUE
)
ansm1
```

```{r}
myeff <- getEffects(mydata)
myeff <- includeDefaultEffects(myeff)

myAlgorithm <- sienaAlgorithmCreate(
  projname = "init",
  # n3 = 3000 # for publication ready results
)
# estimate model
ansm1 <- siena07(
  myAlgorithm,
  data = mydata,
  effects = myeff,
  returnDeps = TRUE
)

ansm1
save(ansm1, file = './ansm1')
```

```{r}
# myeff <- getEffects(mydata)
# myeff <- includeDefaultEffects(myeff)
myeff <- includeEffects(myeff, egoX, interaction1 = "gender") # altX, sameX
myeff <- includeEffects(myeff, altX, interaction1 = "gender")
myeff <- includeEffects(myeff, sameX, interaction1 = "gender")
myeff <- includeEffects(myeff, egoX, interaction1 = "gender", name='last') # altX, sameX
myeff <- includeEffects(myeff, altX, interaction1 = "gender", name='last')
myeff <- includeEffects(myeff, sameX, interaction1 = "gender", name='last')
# too few minority in sample leave out for now.
# myeff <- includeEffects(myeff, egoX, altX, simX, interaction1 = 'mig$M4') # altX, sameX
# myeff <- includeEffects(myeff, egoX, altX, simX, interaction1 = 'mig$other') # altX, sameX

ansm2 <- siena07(
  myAlgorithm,
  data = mydata,
  effects = myeff,
  returnDeps = TRUE
)

save(ansm2, file = './ansm2')
ansm2
```

```{r}
# myeff <- getEffects(mydata)
# myeff <- includeDefaultEffects(myeff)
myeff <- includeEffects(myeff, egoX, altX, interaction1 = 'funcs[[1]]')
myeff <- includeEffects(myeff, egoX, altX, interaction1 = 'funcs[[2]]')
myeff <- includeEffects(myeff, egoX, altX, interaction1 = 'funcs[[3]]')
myeff <- includeEffects(myeff, egoX, altX, interaction1 = 'funcs[[4]]')
myeff <- includeEffects(myeff, egoX, altX, interaction1 = 'funcs[[1]]', name='last')
myeff <- includeEffects(myeff, egoX, altX, interaction1 = 'funcs[[2]]', name='last')
myeff <- includeEffects(myeff, egoX, altX, interaction1 = 'funcs[[3]]', name='last')
myeff <- includeEffects(myeff, egoX, altX, interaction1 = 'funcs[[4]]', name='last')

ansm3 <- siena07(
  myAlgorithm,
  data = mydata,
  effects = myeff,
  returnDeps = TRUE,
)

save(ansm3, file = './ansm3')

ansm3
```

```{r}
# myeff <- getEffects(mydata)
# myeff <- includeDefaultEffects(myeff)
myeff <- includeEffects(myeff, egoX, altX, interaction1 = 'intdis') # simX does noet work
myeff <- includeEffects(myeff, egoX, altX, interaction1 = 'intdis', name='last') # simX does noet work
myeff <- includeEffects(myeff, X, interaction1 = 'dissim')
myeff <- includeEffects(myeff, X, interaction1 = 'dissim', name='last')
myeff <- includeEffects(myeff, X, interaction1 = "distan")
myeff <- includeEffects(myeff, X, interaction1 = "distan", name='last')

ansm4 <- siena07(
  myAlgorithm,
  data = mydata,
  effects = myeff,
  returnDeps = TRUE,
)

save(ansm4, file = './ansm4')
ansm4
```

```{r}
res = list()
res[['ansm1']] = tidy_rsiena(ansm1)
res[['ansm2']] = tidy_rsiena(ansm2)
res[['ansm3']] = tidy_rsiena(ansm3)
res[['ansm4']] = tidy_rsiena(ansm4)
```

```{r}
res_table <- names(res) |>
  purrr::imap(function(name, i) {
    res[[i]] |>
      mutate(
        effect = str_remove(effect, "first: ") |> str_remove("last: ")
      ) |>
      select(depvar, effect, festimate, fse) |>
      rename_with(~ paste0(str_remove(name, 'ans'), "_", .), c(festimate, fse))
  }) |>
  purrr::reduce(full_join, by = join_by(depvar, effect)) |>
  arrange(depvar)

res_table |> writexl::write_xlsx(file.path(BASEDIR, 'results', 'res_table.xlsx'))
res_table
```

```{r}
setwd(BASEDIR)
```
