---
title: "Interdisciplinarity measures"
date: "Last compiled on `r format(Sys.time(), '%B, %Y')`"
output: 
  html_document:
    css: tweaks.css
    toc:  true
    toc_float: true
    number_sections: false
    code_folding: show
    code_download: yes
eval: false
---




```{r, globalsettings, echo=FALSE, warning=FALSE}

library(knitr)
#library(rgl)
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE, warning = FALSE, message = FALSE,comment = "#>", cache=TRUE, class.source=c("test"), class.output=c("test2"), cache.lazy = FALSE)
options(width = 100)
rgl::setupKnitr()

colorize <- function(x, color) {sprintf("<span style='color: %s;'>%s</span>", color, x) }

```

```{r klippy, echo=FALSE, include=TRUE, eval=TRUE}
klippy::klippy(position = c('top', 'right'))
#klippy::klippy(color = 'darkred')
#klippy::klippy(tooltip_message = 'Click to copy', tooltip_success = 'Done')
```


----

This lab journal demonstrates how we construct our measures of interdisciplinarity. 
  

----

Starting with an empty workspace.

```{r, echo=FALSE}

rm(list = ls())

```



# Custom functions

- `package.check`: Check if packages are installed (and install if not) in R ([source](https://vbaliga.github.io/verify-that-r-packages-are-installed-and-loaded/)).  


```{r, results='hide'}

rm(list=ls())

fpackage.check <- function(packages) {
  lapply(packages, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  })
}

```


---  

# Packages

- `tidyverse`: for tidy data manipulations
- `slider`: for calculating rolling sums (summing publications and IDR scores over multiple years)
- `Matrix`: for creating cosine similarity matrices based on cooccurrence of subfields
- `proxy`: for calculating cosine similarity
- `car`: for calculating Variance Inflation Matrix



```{r, results='hide'}

packages = c("tidyverse", "slider", "Matrix", "proxy", "car")

fpackage.check(packages)

```


--- 

# Input


We load the following datasets:

* [oa_map]("./data/OpenAlex_topic_mapping_table.csv"): mapping table to transform OpenAlex topics into fields and subfields, and vice versa

* [phds_variables]("./data/processed/phds_variables.rda"): processed dataset containing all PhDs with covariates.

* [oa_allpubs_disci]("./data/oa_allpubs_disci.rda"): all publications sourced from OpenAlex with associated subfields and subfield scores, scraping and cleaning of this dataset detailed in [openalex_publications](oa_articles.html)
    

```{r data}


oa_map <- read.csv(file="./data/OpenAlex_topic_mapping_table.csv")

load(file="./data/processed/phds_variables_site.rda")

# renaming so the code works
phds_variables <- phds_variables_site
rm(phds_variables_site)

load(file="./data/openalex_pubs/all_publications/oa_allpubs_disci_site.rda")

# same procedure
oa_allpubs_disci <- oa_allpubs_disci_site
rm(oa_allpubs_disci_site)

```



# Publications

Wide to long


```{r}

oa_allpubs_disci %>%
  arrange(id, year) %>%
  filter(!is.na(id) & !is.na(work_id) & id!="NA" & work_id!="NA") %>%
  mutate(subfieldid_1 = str_extract(subfield1_id, "[:digit:]{4}"),
         subfieldid_2 = str_extract(subfield2_id, "[:digit:]{4}"),
         subfieldid_3 = str_extract(subfield3_id, "[:digit:]{4}"),
         subfieldscore_1 = subfield1_score,
         subfieldscore_2 = subfield2_score,
         subfieldscore_3 = subfield3_score,
         workid = work_id) %>%
  select(id, workid, subfieldid_1, subfieldid_2, subfieldid_3, subfieldscore_1, subfieldscore_2, subfieldscore_3) -> pubs


# selecting only one row per work ID, so we don't have as much data to work with
pubs <- pubs %>% select(workid, starts_with("subfield")) %>% distinct(workid, .keep_all=TRUE)

# shaping the data to long
pubs <- reshape(pubs, direction = "long", varying = c(2:7), sep="_", idvar="workid", timevar="subfield_no")


# aggregating the scores of topics with the same subfield 
pubs %>% 
  mutate(subfieldscore = as.numeric(subfieldscore)) %>%
  filter(!is.na(subfieldscore)) %>%
  arrange(workid, subfield_no) %>%
  group_by(workid, subfieldid) %>% 
  summarize(subfieldscore = sum(subfieldscore)) -> pubs

# sorting the subfield scores from high-to-low and naming the subfields (1-3) accordingly 
pubs %>%
  group_by(workid) %>%
  arrange(desc(subfieldscore)) %>%
  mutate(subfield_n = row_number()) -> pubs

# making our publications dataframe matchable for the subfield data
oa_allpubs_disci %>%
  filter(!is.na(id) & !is.na(work_id) & id!="NA" & work_id!="NA") %>%
  mutate(workid = work_id) %>%
  select(id, workid, year) %>%
  arrange(id, year) -> pubs0

# adding the subfield scores
pubs0 %>%
  left_join(pubs, by="workid") %>%
  arrange(id, workid, subfield_n) %>%
  filter(!is.na(subfield_n)) -> pubs

```




# Summing subfields across publications


Summing scores for individuals in the same year, and then calculating a rolling sum of 5 years (7,9 robustness)


```{r}

# reshaping back to wide, so we have all subfield scores per work in 1 row
pubs_wide <- pubs %>%
  mutate(subfield_col = paste0("subfield_", subfieldid)) %>%
  pivot_wider(
    id_cols = c(id, workid, year),  # Variables to retain as identifiers
    names_from = subfield_col,     # Use the combined subfield information for column names
    values_from = subfieldscore    # Fill columns with subfield scores
  )

# First we aggregate subfield scores within years (e.g. multiple works in the same subfield in a year together)
pubs %>%
  filter(year > 1985) %>%
  mutate(subfieldscore = replace_na(subfieldscore, 0)) %>%
  group_by(id, year, subfieldid) %>%
  summarize(subfieldscore = sum(subfieldscore)) -> pubs_summ

pubs_summ$year <- as.numeric(pubs_summ$year)

# Create a full grid of all subfields a scholar has published in in their career
subfields_per_id <- pubs_summ %>%
  ungroup() %>%
  select(subfieldid, id) %>%
  distinct()

# get all id-year combos in our data
years_per_id <- pubs_summ %>%
  ungroup() %>%
  select(id, year) %>%
  distinct()

# create the full grid of subfields associated with each id
full_grid <- years_per_id %>%
  inner_join(subfields_per_id, by="id")


# Join original data to this full grid
pubs_complete <- full_grid %>%
  left_join(pubs_summ, by = c("id", "year", "subfieldid")) %>%
  mutate(subfieldscore = replace_na(subfieldscore, 0))


# We create a rolling sum for these scores, as we want to make this a bit less dependent on yearly changes
pubs_complete %>%
  group_by(id, subfieldid) %>%
  arrange(id, subfieldid, year) %>%
  mutate(rolling_sum = slider::slide_dbl(subfieldscore, ~ sum(.x),
                                 .before = 4,
                                 .complete = FALSE)) %>%
  mutate(rolling_sum = ifelse(is.na(rolling_sum), subfieldscore, rolling_sum)) %>% ungroup() -> pubs_roll5


pubs_complete %>%
  group_by(id, subfieldid) %>%
  arrange(year) %>%
  mutate(rolling_sum = slider::slide_dbl(subfieldscore, ~ sum(.x),
                                 .before = 6,
                                 .complete = FALSE)) %>%
  mutate(rolling_sum = ifelse(is.na(rolling_sum), subfieldscore, rolling_sum)) %>% ungroup() -> pubs_roll7



pubs_complete %>%
  group_by(id, subfieldid) %>%
  arrange(year) %>%
  mutate(rolling_sum = slider::slide_dbl(subfieldscore, ~ sum(.x),
                                 .before = 8,
                                 .complete = FALSE)) %>%
 mutate(rolling_sum = ifelse(is.na(rolling_sum), subfieldscore, rolling_sum)) %>% ungroup() -> pubs_roll9


# removing rows where the rolling sum is 0
pubs_roll5 <- pubs_roll5 %>% filter(rolling_sum!=0)
pubs_roll7 <- pubs_roll7 %>% filter(rolling_sum!=0)
pubs_roll9 <- pubs_roll9 %>% filter(rolling_sum!=0)


head(pubs_summ %>% arrange(id, year))

```



Calculating a rolling sum of publications (to check correlation with IDR measures)


```{r}

pub_counts <- pubs %>%
  group_by(id, year) %>%
  summarize(pub_count = n_distinct(workid), .groups = "drop") %>%
  arrange(id, year)

pub_counts <- pub_counts %>%
  group_by(id) %>%
  arrange(year) %>%
  mutate(pub_count_rolling5 = slider::slide_dbl(pub_count,
                                               ~ sum(.x),
                                               .before = 4,
                                               .complete = FALSE)) %>%
  ungroup() %>%
  mutate(year = as.numeric(year))


```




# Top 3 subfields


```{r}

oa_map %>%
  distinct(subfield_id, .keep_all = TRUE) %>%
  select(subfield_id, subfield_name) %>% 
  mutate(subfield_id = as.character(subfield_id)) %>%
  right_join(pubs_roll5, by=c("subfield_id"="subfieldid")) -> top3_subfields


top3_subfields %>%
  group_by(id, year) %>%
  arrange(desc(subfieldscore), .by_group=TRUE) %>%
  slice_head(n=3) %>%
  mutate(subfield_no = row_number()) %>%
  ungroup() -> top3_subfields

top3_wide <- top3_subfields %>%
  pivot_wider(
    id_cols = c(id, year),
    names_from = subfield_no,
    values_from = c(subfield_id, subfield_name),
    names_glue = "{.value}{subfield_no}"
  )


```



## RAFOLS: BALANCE

Evenness of spread across different subfields (HHI)

```{r}

pubs_balance <- pubs_roll5 %>%
  group_by(id, year) %>%
  mutate(tot_score = sum(rolling_sum, na.rm=TRUE),
         share = rolling_sum / tot_score,
         share_sq = share^2) %>%
  arrange(id, year) %>%
  summarise(HHI = sum(share_sq, na.rm=TRUE),
            balance = 1-HHI) %>% ungroup()



```




## RAFOLS: DISPARITY


### Co-occurrence matrix of subfields in individuals


```{r}

# creating a binary dataframe where each existing person-subfield combination has a row, and we indicate that this combination is there by creating the `present` variable
pubs_summ_binary <- pubs_roll5 %>%
  mutate(present = 1) %>%
  select(id, subfieldid, present) %>%
  distinct()

# sparse matrix, one row per individual, subfields in columns and a 1 if that combination is there
coocc_mat <- pubs_summ_binary %>%
  pivot_wider(names_from = subfieldid, values_from = present, values_fill=0)

# removing the id variable, so we have only the subfield pairings in the row
mat <- as.matrix(coocc_mat[,-1])

# subfields to rows, researchers to cols
mat_t <- t(mat)

# computing cosine similarity
cosine_sim_mat <- simil(mat_t, method = "cosine")


# converting to matrix format
cosine_sim_mat <- as.matrix(cosine_sim_mat)

# calculating distance (i.e. 1-similarity)
cosine_dist_matrix <- 1 - cosine_sim_mat


# check if the distribution of cosine distances is somewhat normal
diff_values <- cosine_dist_matrix[lower.tri(cosine_dist_matrix)]
summary(diff_values)

# Convert cosine_dist_matrix to a long-format data frame
dist_df <- as.data.frame(as.table(cosine_dist_matrix))
colnames(dist_df) <- c("subfield_i", "subfield_j", "d_ij")


```



### Calculating average disparity between combined fields for individuals


```{r}

# Getting all unique subfield pairs per author-year
pubs_disparity <- pubs_roll5 %>%
  distinct(id, year, subfieldid) %>%
  inner_join(., ., by = c("id", "year")) %>%
  filter(subfieldid.x < subfieldid.y) %>%  # avoid duplicates and self-pairs
  rename(subfield_i = subfieldid.x,
         subfield_j = subfieldid.y)

# add distance between subfield pairs
pubs_disparity <- pubs_disparity %>%
  inner_join(dist_df, by = c("subfield_i", "subfield_j"))

# average distance
pubs_disparity_summary <- pubs_disparity %>%
  group_by(id, year) %>%
  summarise(disparity = mean(d_ij, na.rm = TRUE)) %>%
  ungroup()


```




# RAO-STIRLING


Calculating the measures across the 5-year window data specified in `pubs_roll5`


```{r}

# calculate the proportion of the total score that is taken up by given subfield in a year
subfield_props <- pubs_roll5 %>%
  group_by(id, year, subfieldid) %>%
  summarise(score = sum(rolling_sum, na.rm = TRUE)) %>%
  group_by(id, year) %>%
  mutate(tot_score = sum(score, na.rm = TRUE),
         p_i = score / tot_score) %>%
  ungroup()


# creating a pair dataframe, where each row is one pair of subfields that co-occur for a person-year
subfield_pairs <- subfield_props %>%
  inner_join(subfield_props, by = c("id", "year")) %>%
  rename(subfield_i = subfieldid.x,
         subfield_j = subfieldid.y,
         p_i = p_i.x,
         p_j = p_i.y)


# join distance the subfield pairs
subfield_pairs <- subfield_pairs %>%
  inner_join(dist_df, by = c("subfield_i", "subfield_j"))

# calculate pairwise Rao-Stirling diversity (pi*pj*dij = prop subfield i, prop subfield j, distance between i&j)
subfield_pairs <- subfield_pairs %>%
  mutate(contribution = p_i * p_j * d_ij)

# summing the pairwise measures for the Rao-Stirling Diversity in a given year
rao_stirling <- subfield_pairs %>%
  group_by(id, year) %>%
  summarise(rao_stirling = sum(contribution, na.rm = TRUE)) %>%
  ungroup()


```



# Rao-Stirling 7/9


Using a 7 years and 9 years window of publications as a robustness check.


```{r}

# 7 years

# calculate the proportion of the total score that is taken up by given subfield in a year
subfield_props <- pubs_roll7 %>%
  group_by(id, year, subfieldid) %>%
  summarise(score = sum(rolling_sum, na.rm = TRUE)) %>%
  group_by(id, year) %>%
  mutate(tot_score = sum(score, na.rm = TRUE),
         p_i = score / tot_score) %>%
  ungroup()


# creating a pair dataframe, where each row is one pair of subfields that co-occur for a person-year
subfield_pairs <- subfield_props %>%
  inner_join(subfield_props, by = c("id", "year")) %>%
  rename(subfield_i = subfieldid.x,
         subfield_j = subfieldid.y,
         p_i = p_i.x,
         p_j = p_i.y)


# join distance the subfield pairs
subfield_pairs <- subfield_pairs %>%
  inner_join(dist_df, by = c("subfield_i", "subfield_j"))

# calculate pairwise Rao-Stirling diversity (pi*pj*dij = prop subfield i, prop subfield j, distance between i&j)
subfield_pairs <- subfield_pairs %>%
  mutate(contribution = p_i * p_j * d_ij)

# summing the pairwise measures for the Rao-Stirling Diversity in a given year
rao_stirling7 <- subfield_pairs %>%
  group_by(id, year) %>%
  summarise(rao_stirling7 = sum(contribution, na.rm = TRUE)) %>%
  ungroup()


# 9 years 

# calculate the proportion of the total score that is taken up by given subfield in a year
subfield_props <- pubs_roll9 %>%
  group_by(id, year, subfieldid) %>%
  summarise(score = sum(rolling_sum, na.rm = TRUE)) %>%
  group_by(id, year) %>%
  mutate(tot_score = sum(score, na.rm = TRUE),
         p_i = score / tot_score) %>%
  ungroup()


# creating a pair dataframe, where each row is one pair of subfields that co-occur for a person-year
subfield_pairs <- subfield_props %>%
  inner_join(subfield_props, by = c("id", "year")) %>%
  rename(subfield_i = subfieldid.x,
         subfield_j = subfieldid.y,
         p_i = p_i.x,
         p_j = p_i.y)


# join distance the subfield pairs
subfield_pairs <- subfield_pairs %>%
  inner_join(dist_df, by = c("subfield_i", "subfield_j"))

# calculate pairwise Rao-Stirling diversity (pi*pj*dij = prop subfield i, prop subfield j, distance between i&j)
subfield_pairs <- subfield_pairs %>%
  mutate(contribution = p_i * p_j * d_ij)

# summing the pairwise measures for the Rao-Stirling Diversity in a given year
rao_stirling9 <- subfield_pairs %>%
  group_by(id, year) %>%
  summarise(rao_stirling9 = sum(contribution, na.rm = TRUE)) %>%
  ungroup()


```




# Combining the interdisciplinary measures in 1 dataframe


```{r}

pubs_balance %>%
  left_join(pub_counts, by=c("id", "year")) %>%
  left_join(pubs_disparity_summary, by=c("id", "year")) %>%
  left_join(rao_stirling, by=c("id", "year")) %>%
  left_join(rao_stirling7, by=c("id", "year")) %>%
  left_join(rao_stirling9, by=c("id", "year")) -> interdisciplinarity

```



# Calculating correlation between interdisciplinarity measures


```{r}


cor_matrix <- interdisciplinarity %>%
  select(balance, disparity, rao_stirling, pub_count_rolling5) %>%
  na.omit() %>%
  cor()


print(cor_matrix)

```


Please note that this does not match to the table presented in the manuscript, due to the use of synthetic (example) data in this script. 



# Checking Variance Inflation Factor


Cutoff: VIF <5 = acceptable amount of multicollinearity


```{r}

model <- lm(rao_stirling ~ balance + disparity + pub_count_rolling5, data = interdisciplinarity)

vif(model)


```

All good! 



## Saving

```{r, eval=FALSE}

# selecting only interdisciplinarity measures "to keep"
interdisciplinarity %>%
  select(id, year, balance, disparity, rao_stirling, rao_stirling7, rao_stirling9) -> interdisciplinarity

save(interdisciplinarity, file="./data/processed/interdisciplinarity_site.rda")

```

